---
title: "Annotate with clean NumberSubmitters"
output:
  html_document:
    df_print: paged
    code_folding: hide
---

# Aim

The `NumberSubmitters` column in the ClinVar variant files counts all submissions regardless of origin (germline and somatic). This script replaces it with a germline-only count (`NumberSubmitters_clean`) derived from the ClinVar submission summary, using a regex filter on the `OriginCounts` field.

This corrected count is used downstream in `Data_prep_gene_groups.Rmd` to weight variant representation in the long-form training data.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
theme_set(theme_bw(14))
```

# Load ClinVar submission summary and compute germline counts

```{r, message=FALSE, warning=FALSE}
cols <- c(
  "VariationID", "ClinicalSignificance", "DateLastEvaluated", "Description",
  "SubmittedPhenotypeInfo", "ReportedPhenotypeInfo", "ReviewStatus",
  "CollectionMethod", "OriginCounts", "Submitter", "SCV",
  "SubmittedGeneSymbol", "ExplanationOfInterpretation",
  "SomaticClinicalImpact", "Oncogenicity", "ContributesToAggregateClassification"
)

clinvar_subs <- read_tsv(
  "../../Data/Clinvar_060225/submission_summary_2025-02_noheader.txt",
  col_names = cols
)

# Count germline submissions per variant (germline, de novo, parental, inherited)
germline_counts <- clinvar_subs %>%
  filter(grepl("germline|de novo|paternal|maternal|inherited", OriginCounts)) %>%
  count(VariationID, name = "NumberSubmitters_clean")
```

# Helper functions

```{r}
# Load VariationID lookup for a given pathogenicity class
load_variation_ids <- function(path) {
  read_tsv(path, show_col_types = FALSE) %>%
    mutate(variant = paste(Chromosome, PositionVCF, ReferenceAlleleVCF, AlternateAlleleVCF, sep = "_")) %>%
    dplyr::select(variant, VariationID)
}

# Add VariationID, join germline counts, rename NumberSubmitters columns
add_clean_nsub <- function(df, id_lookup) {
  df %>%
    mutate(
      Chromosome = str_remove(chrom, "^chr"),
      variant    = paste(Chromosome, pos, ref, alt, sep = "_")
    ) %>%
    left_join(id_lookup, by = "variant") %>%
    left_join(germline_counts, by = "VariationID") %>%
    dplyr::select(-Chromosome, -variant, -VariationID) %>%
    dplyr::rename(NumberSubmitters_old = NumberSubmitters)
}
```

# Process pathogenic variants

```{r, message=FALSE, warning=FALSE}
clinvar_path <- read_tsv(
  "../../Data/Variant_lists/Clinvar_DDG2P_missense_path_annot_260526_public.tsv.gz"
)

path_ids <- load_variation_ids(
  "../../Data/Clinvar_060225/Clinvar_build_38_DDG2P_germline_path_missense.tsv.gz"
)

clinvar_path_clean <- add_clean_nsub(clinvar_path, path_ids)
```

# Process benign variants

```{r, message=FALSE, warning=FALSE}
clinvar_benign <- read_tsv(
  "../../Data/Variant_lists/Clinvar_DDG2P_missense_benign_annot_260526_public.tsv.gz"
)

benign_ids <- load_variation_ids(
  "../../Data/Clinvar_060225/Clinvar_build_38_DDG2P_germline_benign_missense.tsv.gz"
)

clinvar_benign_clean <- add_clean_nsub(clinvar_benign, benign_ids)
```

# Process VUS variants

```{r, message=FALSE, warning=FALSE}
clinvar_VUS <- read_tsv(
  "../../Data/Variant_lists/Clinvar_DDG2P_missense_VUS_annot_260526_public.tsv.gz"
)

VUS_ids <- load_variation_ids(
  "../../Data/Clinvar_060225/Clinvar_build_38_DDG2P_germline_VUS_conflicting_missense.tsv.gz"
)

clinvar_VUS_clean <- add_clean_nsub(clinvar_VUS, VUS_ids)
```

# Diagnostic: correlation between old and new NumberSubmitters

Expected: variants with somatic-dominated classifications (BRAF V600E, PIK3CA H1047R) drift downwards; genuine germline-only variants remain unchanged.

```{r}
clinvar_path_clean %>%
  mutate(
    name = paste(gene, codon_change),
    diff = NumberSubmitters_old - NumberSubmitters_clean
  ) %>%
  ggplot(aes(x = NumberSubmitters_old, y = NumberSubmitters_clean)) +
  geom_point(alpha = 0.4) +
  ggrepel::geom_text_repel(
    data = . %>% filter(diff > 10),
    aes(label = name),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    x = "NumberSubmitters (original — germline + somatic)",
    y = "NumberSubmitters (clean — germline only)",
    title = "Pathogenic variants: old vs. clean NumberSubmitters"
  )
```

Note: a small number of variants (e.g. BRAF V600E) retain non-zero germline submission counts after filtering. Inspection of these entries suggests they represent high-level mosaic events or imprecise origin labelling in ClinVar; they cannot be reliably removed without manual curation.

# Export

```{r}
write_tsv(clinvar_path_clean,  "../../Data/Variant_lists/Clinvar_DDG2P_missense_path_annot_260526_public.tsv.gz")
write_tsv(clinvar_benign_clean, "../../Data/Variant_lists/Clinvar_DDG2P_missense_benign_annot_260526_public.tsv.gz")
write_tsv(clinvar_VUS_clean,    "../../Data/Variant_lists/Clinvar_DDG2P_missense_VUS_annot_260526_public.tsv.gz")
```
