---
title: "Annotate variants for somatic-germline project"
output: html_document
date: "2025-02-11"
---

# Aim

**Documentation record** — outputs already exist in `Data/Variant_lists/`. Re-run only on the cluster where large annotation inputs are available.

Annotate five variant lists (ClinVar pathogenic, benign, VUS; 31k de novo mutations; UK Biobank) with somatic evidence:

- COSMIC variant and codon-level counts (genomic + targeted sequencing)
- TwinsUK sperm nanoseq variant and codon-level counts (published)
- Buccal nanoseq variant and codon-level counts (published)
- Cancer hotspot status (hotspots v2, excluding frameshift/splice-dominated residues)

Variant lists enter this script already annotated with codon position, consequence, and dbNSFP computational scores (REVEL, ESM1b, AlphaMissense) from VEP and dbNSFP pipeline (cluster step — not reproduced here).

Published data sources:
- Sperm nanoseq: Rahbari et al., TwinsUK
- Buccal nanoseq: Abascal et al.
- Cancer hotspots v2: Chang et al.

```{r}
library(tidyverse)
```

```{r, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Load variant lists

```{r}
# DATA_SOURCE: pre-annotated variant lists (VEP + dbNSFP pipeline, cluster resource)
# Files: consequence_dbNSFP_annotated/*.tsv.gz
DNMs <- read_tsv("# DATA_SOURCE: DNM_31k_missense_SNV_csq_ENSGID_annotated.tsv.gz") %>%
  mutate(var_ID = paste(chrom, pos, ref, alt, sep = "_")) %>%
  unique()

Clinvar_benign <- read_tsv("# DATA_SOURCE: Clinvar_benign_DDG2P_missense_SNV_csq_ENSGID_annotated.tsv.gz") %>%
  mutate(var_ID = paste(chrom, pos, ref, alt, sep = "_")) %>%
  unique()

Clinvar_path <- read_tsv("# DATA_SOURCE: Clinvar_path_DDG2P_missense_SNV_csq_ENSGID_annotated.tsv.gz") %>%
  mutate(var_ID = paste(chrom, pos, ref, alt, sep = "_")) %>%
  unique()

Clinvar_VUS <- read_tsv("# DATA_SOURCE: Clinvar_VUS_DDG2P_missense_SNV_csq_ENSGID_annotated.tsv.gz") %>%
  mutate(var_ID = paste(chrom, pos, ref, alt, sep = "_")) %>%
  unique()

Biobank <- read_tsv("# DATA_SOURCE: UK_biobank_missense_DDG2P_csq_ENSGID_annotated.tsv.gz") %>%
  mutate(var_ID = paste(chrom, pos, ref, alt, sep = "_")) %>%
  unique()
```

# COSMIC variant counts

```{r}
# DATA_SOURCE: COSMIC variant counts (genomic + targeted), cluster resource
# File: COSMIC_count_genomes_and_targeted.tsv.gz
COSMIC <- read_tsv("# DATA_SOURCE: COSMIC_count_genomes_and_targeted.tsv.gz") %>%
  mutate(chrom = paste0("chr", chrom))

join_and_replace_cosmic <- function(df) {
  df %>%
    left_join(COSMIC, by = c("chrom", "pos", "ref", "alt")) %>%
    mutate(
      n_COSMIC = replace_na(n_COSMIC, 0),
      n_COSMIC_genome = replace_na(n_COSMIC_genome, 0),
      n_COSMIC_targeted = replace_na(n_COSMIC_targeted, 0)
    )
}

DNMs_1         <- join_and_replace_cosmic(DNMs)
Clinvar_benign_1 <- join_and_replace_cosmic(Clinvar_benign)
Clinvar_path_1   <- join_and_replace_cosmic(Clinvar_path)
Clinvar_VUS_1    <- join_and_replace_cosmic(Clinvar_VUS)
Gnomad_biobank_1 <- join_and_replace_cosmic(Biobank)
```

```{r}
# DATA_SOURCE: COSMIC codon-level missense counts, cluster resource
# File: COSMIC_DDG2P_missense_SNV_codon_count.tsv.gz
COSMIC_codon <- read_tsv("# DATA_SOURCE: COSMIC_DDG2P_missense_SNV_codon_count.tsv.gz")

join_and_replace_cosmic_codon <- function(df) {
  df %>%
    left_join(COSMIC_codon, by = c("transcript", "codon")) %>%
    mutate(
      n_COSMIC_genome_codon_count   = replace_na(n_COSMIC_genome_codon_count, 0),
      n_COSMIC_targeted_codon_count = replace_na(n_COSMIC_targeted_codon_count, 0),
      n_COSMIC_codon_count          = replace_na(n_COSMIC_codon_count, 0)
    )
}

DNMs_2           <- join_and_replace_cosmic_codon(DNMs_1)
Clinvar_benign_2 <- join_and_replace_cosmic_codon(Clinvar_benign_1)
Clinvar_path_2   <- join_and_replace_cosmic_codon(Clinvar_path_1)
Clinvar_VUS_2    <- join_and_replace_cosmic_codon(Clinvar_VUS_1)
Gnomad_biobank_2 <- join_and_replace_cosmic_codon(Gnomad_biobank_1)
```

# Sperm nanoseq (TwinsUK — published)

```{r}
# DATA_SOURCE: TwinsUK sperm nanoseq variant-level counts, cluster resource
# File: Sperm_seq_DDG2P_missense_MANE_190126.tsv.gz
# Only TwinsUK columns (n_sperm_twinsUK, n_sperm_exomes_twinsUK) retained
sperm_nanoseq <- read_tsv("# DATA_SOURCE: Sperm_seq_DDG2P_missense_MANE_190126.tsv.gz") %>%
  dplyr::select(chrom, pos, ref, alt, n_sperm_twinsUK, n_sperm_exomes_twinsUK)

join_and_replace_sperm <- function(df) {
  df %>%
    left_join(sperm_nanoseq, by = c("chrom", "pos", "ref", "alt")) %>%
    mutate(
      n_sperm_twinsUK        = replace_na(n_sperm_twinsUK, 0),
      n_sperm_exomes_twinsUK = replace_na(n_sperm_exomes_twinsUK, 0)
    )
}

DNMs_3           <- join_and_replace_sperm(DNMs_2)
Clinvar_benign_3 <- join_and_replace_sperm(Clinvar_benign_2)
Clinvar_path_3   <- join_and_replace_sperm(Clinvar_path_2)
Clinvar_VUS_3    <- join_and_replace_sperm(Clinvar_VUS_2)
Gnomad_biobank_3 <- join_and_replace_sperm(Gnomad_biobank_2)
```

```{r}
# DATA_SOURCE: TwinsUK sperm nanoseq codon-level counts, cluster resource
# File: Sperm_seq_missense_MANE_codon_count_190126.tsv.gz
sperm_nanoseq_codon <- read_tsv("# DATA_SOURCE: Sperm_seq_missense_MANE_codon_count_190126.tsv.gz") %>%
  dplyr::select(-n_sperm_codon_count, n_sperm_exomes_codon_count)

join_and_replace_sperm_codon <- function(df) {
  df %>%
    left_join(sperm_nanoseq_codon, by = c("transcript", "codon")) %>%
    mutate(
      n_sperm_twinsUK_codon_count        = replace_na(n_sperm_twinsUK_codon_count, 0),
      n_sperm_exomes_twinsUK_codon_count = replace_na(n_sperm_exomes_twinsUK_codon_count, 0)
    )
}

DNMs_4           <- join_and_replace_sperm_codon(DNMs_3)
Clinvar_benign_4 <- join_and_replace_sperm_codon(Clinvar_benign_3)
Clinvar_path_4   <- join_and_replace_sperm_codon(Clinvar_path_3)
Clinvar_VUS_4    <- join_and_replace_sperm_codon(Clinvar_VUS_3)
Gnomad_biobank_4 <- join_and_replace_sperm_codon(Gnomad_biobank_3)
```

# Buccal nanoseq (published)

```{r}
# DATA_SOURCE: buccal nanoseq variant-level counts, cluster resource
# File: Buccal_seq_DDG2P_missense_MANE_240426.csv
buccal_nanoseq <- read_csv("# DATA_SOURCE: Buccal_seq_DDG2P_missense_MANE_240426.csv") %>%
  dplyr::select(chrom, pos, ref, alt, n_buccal)

join_and_replace_buccal <- function(df) {
  df %>%
    left_join(buccal_nanoseq, by = c("chrom", "pos", "ref", "alt")) %>%
    mutate(n_buccal = replace_na(n_buccal, 0))
}

DNMs_5           <- join_and_replace_buccal(DNMs_4)
Clinvar_benign_5 <- join_and_replace_buccal(Clinvar_benign_4)
Clinvar_path_5   <- join_and_replace_buccal(Clinvar_path_4)
Clinvar_VUS_5    <- join_and_replace_buccal(Clinvar_VUS_4)
Gnomad_biobank_5 <- join_and_replace_buccal(Gnomad_biobank_4)
```

```{r}
# DATA_SOURCE: buccal nanoseq codon-level counts, cluster resource
# File: Buccal_seq_missense_MANE_codon_count_240426.csv
buccal_nanoseq_codon <- read_csv("# DATA_SOURCE: Buccal_seq_missense_MANE_codon_count_240426.csv")

join_and_replace_buccal_codon <- function(df) {
  df %>%
    left_join(buccal_nanoseq_codon, by = c("transcript", "codon")) %>%
    mutate(n_buccal_codon_count = replace_na(n_buccal_codon_count, 0))
}

DNMs_6           <- join_and_replace_buccal_codon(DNMs_5)
Clinvar_benign_6 <- join_and_replace_buccal_codon(Clinvar_benign_5)
Clinvar_path_6   <- join_and_replace_buccal_codon(Clinvar_path_5)
Clinvar_VUS_6    <- join_and_replace_buccal_codon(Clinvar_VUS_5)
Gnomad_biobank_6 <- join_and_replace_buccal_codon(Gnomad_biobank_5)
```

# Cancer hotspot annotation

Residues where >50% of observed variants are frameshift/splice-dominated are excluded — these are not relevant to missense variant interpretation.

```{r}
hotspot_res <- read_tsv("../../Data/Cancer_data/cancerhotspots/hotspots_v2_residues.txt")

hotspot_res_filt <- hotspot_res %>%
  filter(Type == "single residue") %>%
  separate_rows(Variants, sep = "\\|") %>%
  separate(Variants, into = c("aa", "count"), sep = ":", convert = TRUE) %>%
  mutate(disruptive = aa %in% c("*", "sp")) %>%
  group_by(Gene, Residue, Type, `Q-value`, Samples, `Tumor Type Composition`) %>%
  summarise(
    disruptive_count = sum(count[disruptive], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(disruptive_count / Samples <= 0.5) %>%
  inner_join(hotspot_res, by = c("Gene", "Residue", "Type", "Q-value", "Samples", "Tumor Type Composition"))

hotspot_res_filt_for_join <- hotspot_res_filt %>%
  dplyr::select(Gene, Residue, `Q-value`) %>%
  dplyr::rename(hotcodon_q_value = `Q-value`)

add_hotspot <- function(df) {
  df %>%
    mutate(ref_aa    = str_extract(codon_change, "(?<=\\d)[A-Z](?=>)")) %>%
    mutate(ref_codon = paste0(ref_aa, as.character(codon))) %>%
    left_join(hotspot_res_filt_for_join, by = c("gene" = "Gene", "ref_codon" = "Residue")) %>%
    dplyr::select(-ref_aa, -ref_codon) %>%
    mutate(hotcodon_cancer = ifelse(!is.na(hotcodon_q_value), TRUE, FALSE))
}

DNMs_7           <- add_hotspot(DNMs_6)
Clinvar_path_7   <- add_hotspot(Clinvar_path_6)
Clinvar_benign_7 <- add_hotspot(Clinvar_benign_6)
Clinvar_VUS_7    <- add_hotspot(Clinvar_VUS_6)
Gnomad_biobank_7 <- add_hotspot(Gnomad_biobank_6)
```

# Export

```{r}
write_tsv(DNMs_7,           "../../Data/Variant_lists/DNMs_31k_missense_annot_240426_public.tsv.gz")
write_tsv(Clinvar_benign_7, "../../Data/Variant_lists/Clinvar_DDG2P_missense_benign_annot_240426_public.tsv.gz")
write_tsv(Clinvar_path_7,   "../../Data/Variant_lists/Clinvar_DDG2P_missense_path_annot_240426_public.tsv.gz")
write_tsv(Clinvar_VUS_7,    "../../Data/Variant_lists/Clinvar_DDG2P_missense_VUS_annot_240426_public.tsv.gz")
write_tsv(Gnomad_biobank_7, "../../Data/Variant_lists/Biobank_missense_variants_annot_240426_public.tsv.gz")
```
