---
title: "Data preparation — gene group ML model input"
output: html_document
date: "2025-02-20"
---

# Aims

Prepare data for input into supervised ML models and single-feature AUC generation, stratified by gene group.

Steps:

1. Import ClinVar pathogenic, DNM, and Biobank variant lists; remove known pathogenic variants from Biobank
2. Standardise column names (rename `_score` suffix, rename TwinsUK sperm columns to generic `n_sperm`)
3. Combine pathogenic + benign into unique-form dataframes; annotate with gene group flags (Set 1 definitions)
4. Export pre-split whole datasets (unique and long form) for single-feature AUC and downstream analysis
5. Select ML feature columns; slice by gene group; split 70/30 train/test
6. Expand train/test to long_uncontaminated form (train/test split on unique data, then expand)
7. Export all train/test splits for ML model building

Run script 4 (`Annating_with_clean_NumberSubmissions.Rmd`) before this script — it adds `NumberSubmitters_clean` (germline-only submission count) used here for long-form expansion.

```{r, message=FALSE}
library(tidyverse)
library(caret)
library(glue)
library(rlang)

dir.create("../../Data/ML_model_input/Gene_group_models", recursive = TRUE, showWarnings = FALSE)
dir.create("../../Data/ML_model_input/Gene_group_hold_out_models", recursive = TRUE, showWarnings = FALSE)
```

# Import variant lists

```{r}
Clinvar_path <- read_tsv("../../Data/Variant_lists/Clinvar_DDG2P_missense_path_annot_260526_public.tsv.gz")
DNMs         <- read_tsv("../../Data/Variant_lists/DNMs_31k_missense_annot_260526_public.tsv.gz")
Biobank      <- read_tsv("../../Data/Variant_lists/Biobank_missense_variants_annot_260526_public.tsv.gz")
```

# Remove pathogenic variants from Biobank

```{r}
Biobank <- Biobank %>%
  filter(!var_ID %in% c(Clinvar_path$var_ID, DNMs$var_ID))
```

# Standardise columns

Select features of interest, convert "." to NA in dbNSFP columns, remove `_score` suffix, rename TwinsUK sperm columns to generic names.

```{r}
desired_CEPs <- c(
  "SIFT_score", "Polyphen2_HVAR_score", "MutationAssessor_score",
  "MetaSVM_score", "M-CAP_score", "ESM1b_score",
  "PrimateAI_score", "VARITY_R_score", "CADD_raw",
  "AlphaMissense_score", "REVEL_score",
  "popEVE", "ESM1v"
)

CEPs_to_rename <- c(
  "SIFT_score", "Polyphen2_HVAR_score", "MutationAssessor_score",
  "MetaSVM_score", "M-CAP_score", "ESM1b_score",
  "PrimateAI_score", "VARITY_R_score",
  "AlphaMissense_score", "REVEL_score"
)

standardise_variant_df <- function(df, id_col = NULL) {
  shared_cols <- c(
    "gene", "var_ID", "chrom", "pos", "ref", "alt",
    all_of(desired_CEPs),
    "codon", "codon_change",
    "n_COSMIC", "n_COSMIC_genome", "n_COSMIC_targeted",
    "n_COSMIC_codon_count", "n_COSMIC_genome_codon_count", "n_COSMIC_targeted_codon_count",
    "hotcodon_cancer",
    "n_sperm_twinsUK", "n_sperm_exomes_twinsUK",
    "n_sperm_twinsUK_codon_count", "n_sperm_exomes_twinsUK_codon_count",
    "n_buccal", "n_buccal_codon_count"
  )
  if (!is.null(id_col)) shared_cols <- c(shared_cols, id_col)

  df %>%
    dplyr::select(any_of(shared_cols)) %>%
    mutate(across(where(is.character) & any_of(desired_CEPs), ~ na_if(., "."))) %>%
    rename_with(~ str_remove(., "_score$"), any_of(CEPs_to_rename)) %>%
    dplyr::rename(
      n_sperm             = n_sperm_twinsUK,
      n_sperm_exomes      = n_sperm_exomes_twinsUK,
      n_sperm_codon_count = n_sperm_twinsUK_codon_count,
      n_sperm_exomes_codon_count = n_sperm_exomes_twinsUK_codon_count
    )
}

Clinvar_path <- standardise_variant_df(Clinvar_path, id_col = "NumberSubmitters_clean")
DNMs         <- standardise_variant_df(DNMs,         id_col = "n_DNM")
Biobank      <- standardise_variant_df(Biobank)
```

# Build unique-form path vs benign dataframes

```{r}
clinvar_mutation_data_unique <- bind_rows(
  Clinvar_path %>% mutate(Pathogenicity_category = "Pathogenic"),
  Biobank      %>% mutate(Pathogenicity_category = "Benign")
)

DNM_mutation_data_unique <- bind_rows(
  DNMs    %>% mutate(Pathogenicity_category = "Pathogenic"),
  Biobank %>% mutate(Pathogenicity_category = "Benign")
)
```

# Annotate with gene group flags (Set 1)

```{r}
altfunc          <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_AltFunc.csv")
altfunc_stringent <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_AltFunc_stringent.csv")
lof              <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_LOF.csv")
other            <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_other.csv")
RASopathies      <- read_csv("../../Data/Gene_definitions/RASopathy_genes_VCEP_030425.csv") %>% pull(gene)
altfunc_minus_RAS <- setdiff(altfunc$gene_symbol, RASopathies)

annotate_gene_groups <- function(df) {
  df %>%
    mutate(
      altfunc_DD_cancer_gene            = gene %in% altfunc$gene_symbol,
      altfunc_stringent_DD_cancer_gene  = gene %in% altfunc_stringent$gene_symbol,
      lof_DD_cancer_gene                = gene %in% lof$gene_symbol,
      other_DD_cancer_gene              = gene %in% other$gene_symbol,
      RASopathy_gene                    = gene %in% RASopathies,
      altfunc_minus_RASopathy_gene      = gene %in% altfunc_minus_RAS
    )
}

clinvar_mutation_data_unique_set1 <- annotate_gene_groups(clinvar_mutation_data_unique)
DNM_mutation_data_unique_set1     <- annotate_gene_groups(DNM_mutation_data_unique)
```

# Export whole datasets (pre-split)

Used by `Gene_group_single_feature_AUC_AUPRC_F1_MCC.Rmd` and downstream analysis scripts. Long form expands each pathogenic row by `NumberSubmitters_clean` (germline-only count from script 4) or `n_DNM`.

```{r}
write_csv(
  clinvar_mutation_data_unique_set1 %>% dplyr::select(-codon, -codon_change),
  "../../Data/ML_model_input/Gene_group_models/Clinvar_vs_Biobank_unique_set1_240426.csv.gz"
)

clinvar_mutation_data_long_set1 <- clinvar_mutation_data_unique_set1 %>%
  mutate(NumberSubmitters_clean = ifelse(is.na(NumberSubmitters_clean), 1, NumberSubmitters_clean)) %>%
  uncount(NumberSubmitters_clean, .remove = FALSE)

write_csv(
  clinvar_mutation_data_long_set1 %>% dplyr::select(-codon, -codon_change),
  "../../Data/ML_model_input/Gene_group_models/Clinvar_vs_Biobank_long_set1_240426.csv.gz"
)

write_csv(
  DNM_mutation_data_unique_set1 %>% dplyr::select(-codon, -codon_change),
  "../../Data/ML_model_input/Gene_group_models/DNM_vs_Biobank_unique_set1_240426.csv.gz"
)

DNM_mutation_data_long_set1 <- DNM_mutation_data_unique_set1 %>%
  mutate(n_DNM = ifelse(is.na(n_DNM), 1, n_DNM)) %>%
  uncount(n_DNM, .remove = FALSE)

write_csv(
  DNM_mutation_data_long_set1 %>% dplyr::select(-codon, -codon_change),
  "../../Data/ML_model_input/Gene_group_models/DNM_vs_Biobank_long_set1_240426.csv.gz"
)
```

# ML feature selection

```{r}
cols <- c(
  "var_ID", "gene", "NumberSubmitters_clean", "n_DNM",
  "altfunc_DD_cancer_gene", "altfunc_stringent_DD_cancer_gene",
  "altfunc_minus_RASopathy_gene", "lof_DD_cancer_gene",
  "other_DD_cancer_gene", "RASopathy_gene",
  "Pathogenicity_category",
  "n_COSMIC_codon_count", "n_sperm_codon_count", "n_buccal_codon_count",
  "REVEL", "ESM1b", "AlphaMissense"
)

nm  <- c("clinvar_mutation_data_unique_set1", "DNM_mutation_data_unique_set1")
lst <- mget(nm) |> lapply(\(df) dplyr::select(df, dplyr::any_of(cols)))
list2env(setNames(lst, nm), environment())
```

# Slice into gene groups

```{r}
groups <- c(
  altfunc_DD_cancer_gene            = "altfunc",
  altfunc_stringent_DD_cancer_gene  = "altfunc_stringent",
  altfunc_minus_RASopathy_gene      = "altfunc_minus_RAS",
  lof_DD_cancer_gene                = "lof",
  other_DD_cancer_gene              = "other",
  RASopathy_gene                    = "RASopathy"
)

drop_cols     <- names(groups)
all_group_label <- "all"

stub <- function(x) sub(
  "^(.+?)_mutation_data_unique_set(\\d+)(?:_filt2)?$",
  "\\1_unique_set\\2", x
)

grid <- tidyr::expand_grid(
  nm  = nm,
  col = c(names(groups), all_group_label)
)

out <- pmap(grid, function(nm, col) {
  df         <- get(nm)
  group_name <- if (col == all_group_label) "all" else groups[[col]]
  name       <- paste0(stub(nm), "_", group_name)

  sliced <- if (col == all_group_label) {
    df %>% dplyr::select(-all_of(drop_cols))
  } else {
    df %>% filter(.data[[col]] == TRUE) %>% dplyr::select(-all_of(drop_cols))
  }
  setNames(list(sliced), name)
}) |> flatten()

list2env(out, .GlobalEnv)
```

# Train / test split (70/30)

```{r}
set.seed(40)

split_train_test <- function(df_list) {
  result <- list()
  for (df_name in names(df_list)) {
    df <- df_list[[df_name]]
    idx <- createDataPartition(y = df$Pathogenicity_category, p = 0.7, list = FALSE)
    result[[paste0(df_name, "_train")]] <- df[idx, ]
    result[[paste0(df_name, "_test")]]  <- df[-idx, ]
  }
  result
}

train_test_splits <- split_train_test(out)
list2env(train_test_splits, envir = .GlobalEnv)
```

# Create long_uncontaminated data

Train/test split on unique data, then expand to long form to avoid train/test contamination while maximising power.

```{r}
expand_to_long_uncontaminated <- function(data_name) {
  gene_groups <- c("altfunc", "lof", "other", "RASopathy", "altfunc_minus_RAS", "all")
  sets   <- c("set1")
  result <- list()

  for (group in gene_groups) {
    for (set in sets) {
      train_df <- get(paste0(data_name, "_unique_", set, "_", group, "_train"))
      test_df  <- get(paste0(data_name, "_unique_", set, "_", group, "_test"))

      if (data_name == "clinvar") {
        train_df <- train_df %>% mutate(NumberSubmitters_clean = ifelse(is.na(NumberSubmitters_clean), 1, NumberSubmitters_clean))
        test_df  <- test_df  %>% mutate(NumberSubmitters_clean = ifelse(is.na(NumberSubmitters_clean), 1, NumberSubmitters_clean))
        train_long <- train_df %>% uncount(NumberSubmitters_clean, .remove = FALSE)
        test_long  <- test_df  %>% uncount(NumberSubmitters_clean, .remove = FALSE)
      } else {
        train_df <- train_df %>% mutate(n_DNM = ifelse(is.na(n_DNM), 1, n_DNM))
        test_df  <- test_df  %>% mutate(n_DNM = ifelse(is.na(n_DNM), 1, n_DNM))
        train_long <- train_df %>% uncount(n_DNM, .remove = FALSE)
        test_long  <- test_df  %>% uncount(n_DNM, .remove = FALSE)
      }

      result[[paste0(data_name, "_long_uncontaminated_", set, "_", group, "_train")]] <- train_long
      result[[paste0(data_name, "_long_uncontaminated_", set, "_", group, "_test")]]  <- test_long
    }
  }
  result
}

long_uncontaminated_dfs <- c(
  expand_to_long_uncontaminated("clinvar"),
  expand_to_long_uncontaminated("DNM")
)

list2env(long_uncontaminated_dfs, envir = .GlobalEnv)
```

# Export ML model input files

```{r}
data_types  <- c("clinvar_unique", "DNM_unique", "clinvar_long_uncontaminated", "DNM_long_uncontaminated")
sets        <- c("set1")
gene_groups <- c("altfunc", "lof", "other", "RASopathy", "altfunc_minus_RAS", "all")

df_list <- character(0)
for (data in data_types) {
  for (gene in gene_groups) {
    for (set in sets) {
      df_list <- c(df_list,
                   paste0(data, "_", set, "_", gene, "_train"),
                   paste0(data, "_", set, "_", gene, "_test"))
    }
  }
}

output_folder <- "../../Data/ML_model_input/Gene_group_models/"

for (df_name in df_list) {
  write_csv(get(df_name), file = file.path(output_folder, paste0(df_name, "_240426.csv.gz")))
}
```
