---
title: "Define somatic selection gene groups"
output:
  html_document:
    df_print: paged
    code_folding: hide
date: "2026-02-24"
---

# Aim

**Documentation record** — outputs already exist in `Data/Gene_definitions/`. Re-run only if gene list sources change.

Define the union of somatic selection genes (sperm and buccal nanoseq) and COSMIC cancer genes, then create phenotype-defined subsets:

-   `Somatic_selection_cancer_genes.csv` — all genes under somatic selection or in COSMIC CGC
-   `Macrocephaly_monoallelic_DD_somatic_genes.csv` — macrocephaly/megalencephaly panel genes that are also monoallelic DDG2P and somatic selection genes

The analysis section of `Pathway_GO_analysis.Rmd` (Script 10) reads these outputs.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
```

# Define somatic selection genes and cancer genes

```{r, message=FALSE, warning=FALSE}
sperm_selection_genes <- c(
  "PTPN11", "MIB1", "RIT1", "FGFR3", "EP300", "FGFR2", "KDM5B", "NF1", "SMAD6",
  "CUL3", "RASA2", "PRRC2A", "PTEN", "ROBO1", "DDX3X", "CSNK2B", "KRAS", "PPM1D",
  "ARID1A", "BRAF", "HRAS", "KMT2E", "SCAF4", "BMPR2", "TCF12", "CCAR2", "DHX9", "NSD1",
  "LZTR1", "ARHGAP35", "CBL", "SSX1", "RBM12", "SMAD4", "FAM222B",
  "KDM5C", "KMT2D", "AR", "CTNNB1", "RAF1"
)

buccal_selection_genes <- c(
  "NOTCH1", "FAT1", "TP53", "NOTCH2", "ATM", "MGA", "KMT2C", "ASXL1", "CHEK2",
  "TET2", "PPM1D", "BCORL1", "ARID1A", "SETD2", "ZFP36L2", "ARID2", "EP300", "NOTCH3",
  "KDM5C", "RAC1", "ARID1B", "RBM10", "KDM6A", "STAG2", "EPHA2", "CIC", "SF3B1",
  "ZFP36L1", "EGFR", "DNMT3A", "PIK3R1", "TP63", "ARID5B", "AJUBA", "FGFR3", "ZNF750",
  "FUBP1", "CUL3", "KLF5", "PAX9", "FOXP1", "ETV6", "RHOA", "RARG", "SPOP", "SMARCB1",
  "CCND1", "HLA-B", "EIF1AX", "PIK3CA"
)

somatic_selection_genes <- union(sperm_selection_genes, buccal_selection_genes)

cancer_gene_census <- read_tsv("../../Data/Gene_definitions/Cosmic_CancerGeneCensus_v98_GRCh37.tsv.gz",
                               show_col_types = FALSE)
cancer_genes <- cancer_gene_census %>% pull(GENE_SYMBOL) %>% unique()

somatic_selection_and_cancer_genes <- union(somatic_selection_genes, cancer_genes)

write_csv(
  data.frame(gene = somatic_selection_and_cancer_genes),
  "../../Data/Gene_definitions/Somatic_selection_cancer_genes.csv"
)
```

# Phenotype-defined gene lists

## Macrocephaly / megalencephaly

DDG2P monoallelic genes with macrocephaly HPO term intersected with PanelApp Australia macrocephaly/megalencephaly panel, then further subset to somatic selection genes.

```{r, message=FALSE, warning=FALSE}
DDG2P <- read_csv("../../Data/DDG2P/G2P_DD_2025-04-16.csv") %>% clean_names()

DDG2P_monoallelic <- DDG2P %>%
  filter(allelic_requirement %in% c("monoallelic_autosomal", "monoallelic_X_heterozygous"))

macrocephaly_HPO <- "HP:0000256"

DDG2P_macrocephaly_mono <- DDG2P_monoallelic %>%
  filter(str_detect(phenotypes, macrocephaly_HPO)) %>%
  pull(gene_symbol) %>%
  unique()
```

```{r, warning=FALSE, message=FALSE}
# DATA_SOURCE: PanelApp Australia — Macrocephaly/Megalencephaly panel (green genes)
# Download from https://panelapp.agha.umccr.org/ and save to Data/DDG2P/ or similar
# File used: PanelApp_Aus_Macrocephaly_Megalencephaly_green_genes_120326.tsv
panelapp_macrocephaly <- read_tsv("../../Data/DDG2P/PanelApp_Aus_Macrocephaly_Megalencephaly_green_genes_120326.tsv") %>%
  clean_names() %>%
  filter(gene_symbol %in% DDG2P_monoallelic$gene_symbol) %>%
  pull(gene_symbol) %>%
  unique()

panelapp_macrocephaly_somatic <- intersect(panelapp_macrocephaly, somatic_selection_and_cancer_genes)

write_csv(
  data.frame(gene = panelapp_macrocephaly_somatic),
  "../../Data/Gene_definitions/Macrocephaly_monoallelic_DD_somatic_genes.csv"
)
```

Gene counts:

```{r}
cat("DDG2P macrocephaly HPO genes:", length(DDG2P_macrocephaly_mono), "\n")
cat("PanelApp macrocephaly + DDG2P monoallelic:", length(panelapp_macrocephaly), "\n")
cat("PanelApp macrocephaly + somatic selection:", length(panelapp_macrocephaly_somatic), "\n")
```
