---
title: "Define cancer DD gene mechanism sets"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: true
---

# Aim

**Documentation record** — outputs already exist in `Data/Gene_definitions/`. Re-run only if DDG2P or COSMIC CGC inputs change.

Define DD genes with monoallelic inheritance that overlap with somatic selection genes or known cancer genes (COSMIC CGC), then assign each to one of three mechanism groups:

-   **AltFunc**: AltFunc DD genes that are COSMIC oncogenes with a missense mutation pattern
-   **LOF**: LOF_AD DD genes that are COSMIC TSGs
-   **Other**: remaining monoallelic DDG2P genes that overlap with somatic selection or cancer genes
-   **RASopathy**: ClinGen VCEP RASopathy gene panel (defined separately)

Gene set definitions use Set 1 criteria (COSMIC oncogenes with missense filter). See `Archived_scripts/` for Set 2 (COSMIC oncogenes, no missense filter) and Set 3 (oncoKB) exploratory definitions.

```{r}
library(tidyverse)
library(janitor)
library(DT)
```

# Define somatic selection genes

Source: Inigo and Rahelah's Twins UK papers (sperm and buccal nanoseq).

```{r}
sperm_selection_genes <- c(
  "PTPN11", "MIB1", "RIT1", "FGFR3", "EP300", "FGFR2", "KDM5B", "NF1", "SMAD6",
  "CUL3", "RASA2", "PRRC2A", "PTEN", "ROBO1", "DDX3X", "CSNK2B", "KRAS", "PPM1D",
  "ARID1A", "BRAF", "HRAS", "KMT2E", "SCAF4", "BMPR2", "TCF12", "CCAR2", "DHX9", "NSD1",
  "LZTR1", "ARHGAP35", "CBL", "SSX1", "RBM12", "SMAD4", "FAM222B",
  "KDM5C", "KMT2D", "AR", "CTNNB1", "RAF1"
)

buccal_selection_genes <- c(
  "NOTCH1", "FAT1", "TP53", "NOTCH2", "ATM", "MGA", "KMT2C", "ASXL1", "CHEK2",
  "TET2", "PPM1D", "BCORL1", "ARID1A", "SETD2", "ZFP36L2", "ARID2", "EP300", "NOTCH3",
  "KDM5C", "RAC1", "ARID1B", "RBM10", "KDM6A", "STAG2", "EPHA2", "CIC", "SF3B1",
  "ZFP36L1", "EGFR", "DNMT3A", "PIK3R1", "TP63", "ARID5B", "AJUBA", "FGFR3", "ZNF750",
  "FUBP1", "CUL3", "KLF5", "PAX9", "FOXP1", "ETV6", "RHOA", "RARG", "SPOP", "SMARCB1",
  "CCND1", "HLA-B", "EIF1AX", "PIK3CA"
)

somatic_selection_genes <- union(sperm_selection_genes, buccal_selection_genes)
```

# Define cancer gene groups from COSMIC

```{r}
cancer_gene_census <- read_tsv("../../Data/Gene_definitions/Cosmic_CancerGeneCensus_v98_GRCh37.tsv.gz",
                               show_col_types = FALSE)
cancer_genes <- cancer_gene_census %>% pull(GENE_SYMBOL) %>% unique()
somatic_selection_and_cancer_genes <- union(somatic_selection_genes, cancer_genes)

cancer_gene_census_annotated <- cancer_gene_census %>%
  mutate(onco_mis = case_when(
    # Missense filter is important for oncogenes: many are pathogenic via fusion or amplification
    # rather than missense, which would not overlap with germline AF variants
    grepl("oncogene", ROLE_IN_CANCER) & grepl("Mis", MUTATION_TYPES) ~ "Oncogene_missense",
    TRUE ~ "Cancer_gene_other"
  )) %>%
  mutate(TSG = case_when(
    grepl("TSG", ROLE_IN_CANCER) ~ "TSG",
    TRUE ~ "Cancer_gene_other"
  ))

oncogene_missense <- cancer_gene_census_annotated %>%
  filter(onco_mis == "Oncogene_missense") %>%
  pull(GENE_SYMBOL) %>% unique()

TSG <- cancer_gene_census_annotated %>%
  filter(TSG == "TSG") %>%
  pull(GENE_SYMBOL) %>% unique()
```

# DDG2P 2025 — Set 1 gene group definitions

```{r}
DDG2P_2025 <- read_csv("../../Data/DDG2P/G2P_DD_2025-04-16.csv") %>% clean_names()

DDG2P_2025_dominant <- DDG2P_2025 %>%
  filter(allelic_requirement %in% c("monoallelic_X_heterozygous", "monoallelic_autosomal")) %>%
  pull(gene_symbol) %>% unique()
```

Define DD dominant genes that are also under somatic selection and export:

```{r}
DD_2025_dominant_somatic <- data.frame(gene_symbol = DDG2P_2025_dominant) %>%
  filter(gene_symbol %in% somatic_selection_and_cancer_genes)

write_csv(DD_2025_dominant_somatic,
          "../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes.csv")
```

```{r}
DDG2P_2025_mechanisms <- read_csv("../../Data/DDG2P/DDG2P_2025_annotated_with_gene_groups_040725.csv")

AltFunc_genes_2025 <- DDG2P_2025_mechanisms %>%
  filter(AltFunc == 1) %>%
  pull(gene_symbol) %>% unique()

AltFunc_stringent_genes_2025 <- DDG2P_2025_mechanisms %>%
  filter(AltFunc_stringent == 1) %>%
  pull(gene_symbol) %>% unique()

LOF_AD_genes_2025 <- DDG2P_2025_mechanisms %>%
  filter(LOF_AD == 1) %>%
  pull(gene_symbol) %>% unique()

# Set 1: AltFunc DD genes × COSMIC oncogenes with missense pattern
AltFunc_onco_genes_2025 <- intersect(AltFunc_genes_2025, oncogene_missense)
AltFunc_stringent_onco_gene_2025 <- intersect(AltFunc_stringent_genes_2025, oncogene_missense)

# Set 1: LOF_AD DD genes × COSMIC TSGs
LOF_AD_TSGs_2025 <- intersect(LOF_AD_genes_2025, TSG)

# Set 1: remaining dominant DDG2P genes that overlap with somatic selection / cancer genes
DD_cancer_other_2025 <- intersect(DDG2P_2025_dominant, somatic_selection_and_cancer_genes) %>%
  setdiff(LOF_AD_TSGs_2025) %>%
  setdiff(AltFunc_onco_genes_2025)
```

## Gene group sizes

```{r}
data.frame(
  Group = c("AltFunc (liberal)", "AltFunc (stringent)", "LOF", "Other", "Dominant somatic"),
  n = c(
    length(AltFunc_onco_genes_2025),
    length(AltFunc_stringent_onco_gene_2025),
    length(LOF_AD_TSGs_2025),
    length(DD_cancer_other_2025),
    nrow(DD_2025_dominant_somatic)
  )
) %>% datatable()
```

Note: AltFunc and LOF groups are not mutually exclusive. Overlap:

```{r}
intersect(AltFunc_onco_genes_2025, LOF_AD_TSGs_2025)
```

## Export Set 1

```{r}
write_csv(data.frame(gene_symbol = AltFunc_onco_genes_2025),
          "../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_AltFunc.csv")

write_csv(data.frame(gene_symbol = AltFunc_stringent_onco_gene_2025),
          "../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_AltFunc_stringent.csv")

write_csv(data.frame(gene_symbol = LOF_AD_TSGs_2025),
          "../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_LOF.csv")

write_csv(data.frame(gene_symbol = DD_cancer_other_2025),
          "../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_other.csv")
```

# Define RASopathy genes

Source: ClinGen RASopathy VCEP gene list (<https://clinicalgenome.org/affiliation/50021/>). NF1 and LZTR1 excluded — LOF mechanism, phenotypically distinct, and dominate ClinVar counts for the group.

```{r}
RASopathy_genes_VCEP <- c(
  "BRAF", "HRAS", "KRAS", "MAP2K1", "MAP2K2", "NRAS",
  "PTPN11", "RAF1", "RIT1", "SHOC2", "SOS1", "SOS2"
)

write_csv(data.frame(gene = RASopathy_genes_VCEP),
          "../../Data/Gene_definitions/RASopathy_genes_VCEP_030425.csv")
```
