---
title: "Per-gene single-feature AUC, AUPRC, F1, MCC (hold-out gene models)"
output: html_document
date: "2026-04-30"
---

# Aim

For each individual gene (across all gene groups) and dataset, calculate classification metrics
for individual features (somatic evidence scores and computational effect predictors):

- AUC + 95% CI
- AUPRC
- Sensitivity, specificity, precision, recall, accuracy, balanced accuracy, F1, MCC at MCC-optimal threshold

Only genes with ≥ 10 benign and ≥ 10 pathogenic variants are included. For each gene, features
with > 20% missingness in pathogenic variants, or fewer than 10 non-NA variants in either class,
are excluded.

Output: `Data/AUC_tables/Gene_group_hold_out_models/Single_feature_AUCs_individual_genes_<date>.csv`

```{r setup, message=FALSE}
library(tidyverse)
library(pROC)
library(PRROC)
library(parallel)

date    <- "240426"
n_cores <- max(1L, detectCores() - 1L)
message("Using ", n_cores, " cores")

dir.create("../../../Data/AUC_tables/Gene_group_hold_out_models",
           recursive = TRUE, showWarnings = FALSE)

buccal_panel_genes <- read_csv(
  "../../../Data/Buccal_data/Lawson_supplementary_table1_buccal_nanoseq_panel.csv",
  show_col_types = FALSE) %>% pull(gene) %>% unique()
buccal_predictors <- c("n_buccal", "n_buccal_codon_count")
```

# Import data

```{r load-data, message=FALSE}
input_folder <- "../../../Data/ML_model_input/Gene_group_models/"

clinvar_mutation_data_unique <- read_csv(
  file.path(input_folder, paste0("Clinvar_vs_Biobank_unique_set1_", date, ".csv.gz")),
  show_col_types = FALSE)
clinvar_mutation_data_long   <- read_csv(
  file.path(input_folder, paste0("Clinvar_vs_Biobank_long_set1_",   date, ".csv.gz")),
  show_col_types = FALSE)
DNM_mutation_data_unique     <- read_csv(
  file.path(input_folder, paste0("DNM_vs_Biobank_unique_set1_",     date, ".csv.gz")),
  show_col_types = FALSE)
DNM_mutation_data_long       <- read_csv(
  file.path(input_folder, paste0("DNM_vs_Biobank_long_set1_",       date, ".csv.gz")),
  show_col_types = FALSE)

data_sets <- c("clinvar_mutation_data_unique", "clinvar_mutation_data_long",
               "DNM_mutation_data_unique",     "DNM_mutation_data_long")
```

# Metrics functions

Copied verbatim from `Gene_group_single_feature_AUC_AUPRC_F1_MCC.Rmd` so per-gene metrics are
computed identically to the gene-group version.

```{r metrics-functions}
inverted_predictors <- c("ESM1b", "ESM1v", "popEVE", "SIFT")

midpoint_thresholds <- function(x, max_candidates = 500L) {
  vals <- sort(unique(x[!is.na(x)]))
  if (length(vals) <= 1L) return(vals)
  mids <- (vals[-1L] + vals[-length(vals)]) / 2
  candidates <- c(vals[1L] - 1e-12, mids, vals[length(vals)] + 1e-12)
  if (length(candidates) > max_candidates)
    candidates <- unname(quantile(candidates, probs = seq(0, 1, length.out = max_candidates)))
  candidates
}

safe_div <- function(num, den) ifelse(den > 0, num / den, NA_real_)

add_metrics_single_feature <- function(dat, predictor_col_name,
                                       inverted_list = inverted_predictors) {
  dat <- dat %>%
    mutate(Pathogenicity_category = case_when(
      Pathogenicity_category == "Pathogenic" ~ 1,
      Pathogenicity_category == "Benign"     ~ 0,
      TRUE                                   ~ NA_real_
    ))

  dat2 <- dat %>%
    filter(!is.na(Pathogenicity_category) & !is.na(.data[[predictor_col_name]]))

  na_row <- list(
    Predictor = predictor_col_name,
    AUC_full = NA_real_, AUC_full_lower = NA_real_, AUC_full_upper = NA_real_,
    AUPRC = NA_real_, Threshold = NA_real_,
    Sensitivity = NA_real_, Specificity = NA_real_,
    Precision = NA_real_, Recall = NA_real_, Accuracy = NA_real_,
    BalancedAccuracy = NA_real_, F1 = NA_real_, MCC = NA_real_,
    Observed_direction = NA_character_,
    Expected_direction = if (predictor_col_name %in% inverted_list)
      "lower=pathogenic (higher=benign)" else "higher=pathogenic",
    Consistent_with_expected = NA
  )

  if (nrow(dat2) == 0 || length(unique(dat2$Pathogenicity_category)) < 2) return(na_row)

  score <- as.numeric(dat2[[predictor_col_name]])
  y     <- as.integer(dat2$Pathogenicity_category)

  r <- pROC::roc(response = y, predictor = score, levels = c(0, 1),
                 direction = "auto", quiet = TRUE, na.rm = TRUE)

  auc_val <- round(as.numeric(pROC::auc(r)), 4)
  auc_ci  <- pROC::ci.auc(r)
  dir_sym <- r$direction
  obs_dir <- if (identical(dir_sym, "<")) "higher=pathogenic (controls < cases)"
             else                         "higher=benign (controls > cases)"
  exp_dir <- if (predictor_col_name %in% inverted_list)
               "lower=pathogenic (higher=benign)" else "higher=pathogenic"
  consistent <- (grepl("benign", obs_dir)     && grepl("benign", exp_dir)) ||
                (grepl("pathogenic", obs_dir) && grepl("pathogenic", exp_dir))

  inverted  <- predictor_col_name %in% inverted_list
  pr_score  <- if (inverted) -score else score
  pos_s     <- pr_score[y == 1]
  neg_s     <- pr_score[y == 0]
  auprc_val <- if (length(pos_s) > 0 && length(neg_s) > 0)
    round(as.numeric(PRROC::pr.curve(pos_s, neg_s, curve = FALSE)$auc.integral), 4)
  else NA_real_

  candidates <- midpoint_thresholds(score)
  best_mcc <- -Inf; best_thr <- NA_real_; best_sens <- NA_real_; best_spec <- NA_real_

  for (thr_cand in candidates) {
    pred_pos <- if (!inverted) as.integer(score >= thr_cand) else as.integer(score <= thr_cand)
    TP <- as.numeric(sum(pred_pos == 1 & y == 1, na.rm = TRUE))
    TN <- as.numeric(sum(pred_pos == 0 & y == 0, na.rm = TRUE))
    FP <- as.numeric(sum(pred_pos == 1 & y == 0, na.rm = TRUE))
    FN <- as.numeric(sum(pred_pos == 0 & y == 1, na.rm = TRUE))
    s  <- safe_div(TP, TP + FN); p <- safe_div(TN, TN + FP)
    dn <- sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))
    m  <- if (!is.na(dn) && dn > 0) ((TP * TN) - (FP * FN)) / dn else NA_real_
    if (!is.na(m) &&
        (m > best_mcc ||
         (isTRUE(all.equal(m, best_mcc)) && !is.na(s) && (is.na(best_sens) || s > best_sens)) ||
         (isTRUE(all.equal(m, best_mcc)) && isTRUE(all.equal(s, best_sens)) &&
          !is.na(p) && (is.na(best_spec) || p > best_spec)))) {
      best_mcc <- m; best_thr <- thr_cand; best_sens <- s; best_spec <- p
    }
  }

  pred_pos <- if (!is.na(best_thr)) {
    if (!inverted) as.integer(score >= best_thr) else as.integer(score <= best_thr)
  } else rep(NA_integer_, length(score))

  TP <- as.numeric(sum(pred_pos == 1 & y == 1, na.rm = TRUE))
  TN <- as.numeric(sum(pred_pos == 0 & y == 0, na.rm = TRUE))
  FP <- as.numeric(sum(pred_pos == 1 & y == 0, na.rm = TRUE))
  FN <- as.numeric(sum(pred_pos == 0 & y == 1, na.rm = TRUE))
  N  <- TP + TN + FP + FN

  prec <- safe_div(TP, TP + FP); rec <- safe_div(TP, TP + FN)
  acc  <- if (N > 0) (TP + TN) / N else NA_real_
  bal  <- mean(c(best_sens, best_spec), na.rm = TRUE)
  f1   <- if (!is.na(prec) && !is.na(rec) && (prec + rec) > 0) 2 * prec * rec / (prec + rec) else NA_real_
  dn2  <- sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))
  mcc  <- if (!is.na(dn2) && dn2 > 0) ((TP * TN) - (FP * FN)) / dn2 else NA_real_

  list(
    Predictor = predictor_col_name,
    AUC_full = auc_val, AUC_full_lower = round(auc_ci[1], 4), AUC_full_upper = round(auc_ci[3], 4),
    AUPRC = auprc_val,
    Threshold        = if (!is.na(best_thr))  round(best_thr, 6) else NA_real_,
    Sensitivity      = if (!is.na(best_sens)) round(best_sens, 4) else NA_real_,
    Specificity      = if (!is.na(best_spec)) round(best_spec, 4) else NA_real_,
    Precision        = if (!is.na(prec))      round(prec, 4) else NA_real_,
    Recall           = if (!is.na(rec))       round(rec, 4) else NA_real_,
    Accuracy         = if (!is.na(acc))       round(acc, 4) else NA_real_,
    BalancedAccuracy = if (!is.na(bal))       round(bal, 4) else NA_real_,
    F1               = if (!is.na(f1))        round(f1, 4) else NA_real_,
    MCC              = if (!is.na(mcc))       round(mcc, 4) else NA_real_,
    Observed_direction       = obs_dir,
    Expected_direction       = exp_dir,
    Consistent_with_expected = consistent
  )
}

make_metrics_df <- function(my_data, cols, inverted_list = inverted_predictors) {
  bind_rows(lapply(cols, \(col)
    as_tibble(add_metrics_single_feature(my_data, col, inverted_list))
  ))
}
```

# Feature list

```{r feature-list}
AUC_cols_all <- c(
  "n_COSMIC", "n_COSMIC_genome", "n_COSMIC_targeted",
  "n_COSMIC_codon_count", "n_COSMIC_genome_codon_count", "n_COSMIC_targeted_codon_count",
  "hotcodon_cancer",
  "n_sperm", "n_sperm_exomes", "n_sperm_codon_count", "n_sperm_exomes_codon_count",
  "n_buccal", "n_buccal_codon_count",
  "AlphaMissense", "REVEL", "ESM1b",
  "SIFT", "Polyphen2_HVAR", "MutationAssessor", "MetaSVM", "M-CAP",
  "PrimateAI", "VARITY_R", "CADD_raw", "popEVE", "ESM1v"
)
```

# Compute metrics per gene

```{r compute-metrics, message=FALSE}
# Build (dataset_object_name, dataset_label, gene) combo table
combos <- do.call(rbind, lapply(data_sets, function(df_name) {
  df       <- get(df_name)
  nm_parts <- str_split(df_name, "_")[[1]]
  dataset_label <- paste0(nm_parts[1], "_", nm_parts[4])  # e.g. "clinvar_unique"

  genes <- df %>%
    group_by(gene) %>%
    count(Pathogenicity_category) %>%
    pivot_wider(values_from = n, names_from = Pathogenicity_category) %>%
    filter(coalesce(Benign, 0L) > 9L, coalesce(Pathogenic, 0L) > 9L) %>%
    pull(gene)

  if (length(genes) == 0L) return(NULL)
  data.frame(df_name = df_name, dataset = dataset_label, gene = genes,
             stringsAsFactors = FALSE)
}))

message(nrow(combos), " (dataset, gene) combinations to process")

compute_gene_metrics <- function(idx) {
  row     <- combos[idx, ]
  df      <- get(row$df_name, envir = .GlobalEnv)
  gene_df <- filter(df, gene == row$gene)

  pat_df <- filter(gene_df, Pathogenicity_category == "Pathogenic")
  if (nrow(pat_df) == 0L) return(NULL)

  # Stage 1: exclude features with >20% missing in pathogenic variants
  missing_pct <- pat_df %>%
    summarise(across(all_of(AUC_cols_all), ~ mean(is.na(.)) * 100)) %>%
    as.list() %>% unlist()
  cols_low_missing <- names(missing_pct)[!is.na(missing_pct) & missing_pct < 20]

  # Stage 2: require >=10 non-NA in both classes
  cols_enough_data <- gene_df %>%
    select(Pathogenicity_category, any_of(cols_low_missing)) %>%
    pivot_longer(-Pathogenicity_category, names_to = "predictor",
                 values_to = "score", values_drop_na = TRUE) %>%
    count(predictor, Pathogenicity_category, name = "n") %>%
    pivot_wider(names_from = Pathogenicity_category, values_from = n, values_fill = 0L) %>%
    filter(coalesce(Benign, 0L) >= 10L, coalesce(Pathogenic, 0L) >= 10L) %>%
    pull(predictor)

  if (length(cols_enough_data) == 0L) return(NULL)

  make_metrics_df(gene_df, cols_enough_data) %>%
    mutate(dataset = row$dataset, gene = row$gene)
}

results_list <- mclapply(seq_len(nrow(combos)), compute_gene_metrics, mc.cores = n_cores)
all_results  <- bind_rows(Filter(Negate(is.null), results_list))

# Mask all metrics to NA for buccal predictors where gene is not on the buccal panel
numeric_metric_cols <- c("AUC_full", "AUC_full_lower", "AUC_full_upper", "AUPRC",
                         "Threshold", "Sensitivity", "Specificity", "Precision",
                         "Recall", "Accuracy", "BalancedAccuracy", "F1", "MCC")
all_results <- all_results %>%
  mutate(.buccal_mask = Predictor %in% buccal_predictors & !gene %in% buccal_panel_genes) %>%
  mutate(across(all_of(numeric_metric_cols),  ~ if_else(.buccal_mask, NA_real_,      .))) %>%
  mutate(across(c(Observed_direction, Expected_direction),
                                               ~ if_else(.buccal_mask, NA_character_, .))) %>%
  mutate(Consistent_with_expected = if_else(.buccal_mask, NA, Consistent_with_expected)) %>%
  select(-.buccal_mask)

message(nrow(all_results), " rows computed across ",
        n_distinct(all_results$gene), " genes and ",
        n_distinct(all_results$dataset), " datasets")

all_results
```

# Export

```{r export}
dir.create("../../../Data/Tables", recursive = TRUE, showWarnings = FALSE)

write_csv(all_results,
  paste0("../../../Data/AUC_tables/Gene_group_hold_out_models/Single_feature_AUCs_individual_genes_",
         date, ".csv"))
write_csv(all_results,
  "../../../Data/Tables/Gene_group_hold_out_single_feature_metrics.csv")
```
