---
title: "DDX3X benchmarking: public data"
output:
  html_document:
    df_print: paged
    code_folding: hide
  pdf_document: default
date: "2026-02-20"
---

# Aim

Assess the predictive power of somatic data to distinguish benign from pathogenic DDX3X
germline variants using published data only.

Data sources:
- DNMs: trio WES/WGS (31k cohort)
- ClinVar: (likely) pathogenic germline variants
- Biobank: variants seen in UK Biobank, excluding DNMs and ClinVar (likely) pathogenic

All lists are subset to missense variants in the MANE select transcript, annotated
with somatic features.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(pROC)
library(janitor)

dir.create("../../Data/MAVE_benchmarking", recursive = TRUE, showWarnings = FALSE)

read_gene <- function(path, gene_name) {
  read_tsv(path, show_col_types = FALSE) %>%
    filter(gene == gene_name) %>%
    dplyr::rename(
      n_sperm             = n_sperm_twinsUK,
      n_sperm_codon_count = n_sperm_twinsUK_codon_count
    )
}

DDX3X_Biobank      <- read_gene("../../Data/Variant_lists/Biobank_missense_variants_annot_260526_public.tsv.gz",  "DDX3X")
DDX3X_clinvar_path <- read_gene("../../Data/Variant_lists/Clinvar_DDG2P_missense_path_annot_260526_public.tsv.gz","DDX3X")
DDX3X_DNMs         <- read_gene("../../Data/Variant_lists/DNMs_31k_missense_annot_260526_public.tsv.gz",          "DDX3X")

DDX3X_clinvar_phenotype_terms <- read_tsv(
  "../../Data/Clinvar_060225/Clinvar_build_38_DDG2P_germline_path_missense.tsv.gz",
  show_col_types = FALSE
) %>% filter(GeneSymbol == "DDX3X")

```

```{r, message=FALSE, warning=FALSE}
desired_CEPs  <- c("PrimateAI_score", "VARITY_R_score", "CADD_raw",
                   "AlphaMissense_score", "REVEL_score")
new_CEP_names <- c("PrimateAI", "VARITY_R", "CADD_raw", "AlphaMissense", "REVEL")

variant_cols <- c("gene", "codon_change",
                  "REVEL_score", "AlphaMissense_score", "CADD_raw",
                  "VARITY_R_score", "PrimateAI_score",
                  "n_COSMIC", "n_COSMIC_codon_count",
                  "n_sperm", "n_sperm_codon_count",
                  "n_buccal", "n_buccal_codon_count",
                  "hotcodon_cancer", "group")

clean_CEPs <- function(df) {
  df %>%
    mutate(across(where(is.character) & all_of(desired_CEPs), ~ na_if(., "."))) %>%
    mutate(across(all_of(desired_CEPs), as.numeric)) %>%
    rename_with(~ str_remove(., "_score$"), all_of(desired_CEPs))
}

pre_cols <- c(setdiff(variant_cols, "group"), "NumberSubmitters_clean", "n_DNM")

clinvar_biobank_df <- bind_rows(
  DDX3X_clinvar_path %>% dplyr::select(any_of(pre_cols)) %>% mutate(group = "Clinvar_pathogenic"),
  DDX3X_Biobank      %>% dplyr::select(any_of(pre_cols)) %>% mutate(group = "Biobank")
) %>% dplyr::select(all_of(variant_cols)) %>% clean_CEPs()

clinvar_long_biobank_df <- bind_rows(
  DDX3X_clinvar_path %>%
    dplyr::select(any_of(pre_cols)) %>%
    mutate(NumberSubmitters_clean = coalesce(NumberSubmitters_clean, 1L)) %>%
    uncount(NumberSubmitters_clean) %>%
    mutate(group = "Clinvar_pathogenic"),
  DDX3X_Biobank %>% dplyr::select(any_of(pre_cols)) %>% mutate(group = "Biobank")
) %>% dplyr::select(all_of(variant_cols)) %>% clean_CEPs()

DNM_biobank_df <- bind_rows(
  DDX3X_DNMs %>%
    dplyr::select(any_of(pre_cols)) %>%
    mutate(n_DNM = coalesce(n_DNM, 1L)) %>%
    uncount(n_DNM) %>%
    mutate(group = "DNM_pathogenic"),
  DDX3X_Biobank %>% dplyr::select(any_of(pre_cols)) %>% mutate(group = "Biobank")
) %>% dplyr::select(all_of(variant_cols)) %>% clean_CEPs()
```

```{r, message=FALSE, warning=FALSE}
MAVE_files <- list.files("../../Data/MAVE/DDX3X_MAVE/score/", pattern = "*.tsv",
                          full.names = TRUE)
MAVE_data  <- lapply(MAVE_files, read_tsv, show_col_types = FALSE) %>% bind_rows()

MAVE_data_select <- MAVE_data %>%
  dplyr::select(chrom, Protein_position, ref_aa, alt_aa,
                NDD_clinical_classification_prob_var_nonfunctional) %>%
  filter(!is.na(ref_aa)) %>%
  unique() %>%
  group_by(chrom, Protein_position, ref_aa, alt_aa) %>%
  # Maximum probability-of-non-function across replicates, discarding NAs.
  # If every replicate is NA the variant has no usable score (kept as NA rather
  # than the -Inf that max(na.rm = TRUE) would otherwise return).
  summarise(
    MAVE_score = if (all(is.na(NDD_clinical_classification_prob_var_nonfunctional))) {
      NA_real_
    } else {
      max(NDD_clinical_classification_prob_var_nonfunctional, na.rm = TRUE)
    },
    .groups = "drop"
  ) %>%
  mutate(codon_change = paste0(Protein_position, ref_aa, ">", Protein_position, alt_aa)) %>%
  dplyr::select(codon_change, MAVE_score)
```

# Number of DDX3X variants

```{r}
cat("n (likely) pathogenic ClinVar variants:", nrow(DDX3X_clinvar_path), "\n")
cat("n ClinVar long form variants:", nrow(clinvar_long_biobank_df %>% filter(group == "Clinvar_pathogenic")), "\n")
cat("n Biobank variants:", nrow(DDX3X_Biobank), "\n")
cat("n unique DNMs (31k):", nrow(DDX3X_DNMs), "\n")
cat("n total DNMs (31k):", sum(DDX3X_DNMs$n_DNM, na.rm = TRUE), "\n")
```

# MAVE data

```{r}
clinvar_biobank_df_MAVE <- clinvar_biobank_df      %>% left_join(MAVE_data_select, by = "codon_change")
clinvar_long_df_MAVE    <- clinvar_long_biobank_df %>% left_join(MAVE_data_select, by = "codon_change")
DNM_biobank_df_MAVE     <- DNM_biobank_df          %>% left_join(MAVE_data_select, by = "codon_change")
```

## Percent missingness

```{r}
clinvar_biobank_df_MAVE %>% summarise(percent_missing = mean(is.na(MAVE_score)) * 100)
DNM_biobank_df_MAVE     %>% filter(group == "DNM_pathogenic") %>%
  summarise(percent_missing = mean(is.na(MAVE_score)) * 100)
```

# AUC of somatic and MAVE data {.tabset}

```{r auc-helper}
features_to_test <- c(new_CEP_names,
                      "n_COSMIC", "n_COSMIC_codon_count",
                      "n_sperm", "n_sperm_codon_count",
                      "n_buccal", "n_buccal_codon_count",
                      "MAVE_score")

calc_aucs <- function(df, path_level) {
  results <- data.frame(feature = character(), AUC = numeric(),
                        CI_lower = numeric(), CI_upper = numeric(),
                        stringsAsFactors = FALSE)
  for (feat in features_to_test) {
    roc_obj   <- roc(df$group, df[[feat]], levels = c("Biobank", path_level),
                     quiet = TRUE)
    ci_val    <- ci.auc(roc_obj)
    results   <- rbind(results, data.frame(feature = feat,
                                           AUC      = as.numeric(auc(roc_obj)),
                                           CI_lower = ci_val[1],
                                           CI_upper = ci_val[3]))
  }
  results
}

colour_features <- function(df) {
  df %>%
    mutate(colour = case_when(
      feature %in% c("n_COSMIC", "n_COSMIC_codon_count") ~ "Cancer",
      feature %in% c("n_sperm", "n_sperm_codon_count")   ~ "Sperm Nanoseq",
      feature %in% c("n_buccal", "n_buccal_codon_count") ~ "Buccal Nanoseq",
      feature == "MAVE_score"                            ~ "MAVE",
      TRUE ~ "Computational effect predictor"
    )) %>%
    mutate(colour = factor(colour, levels = c(
      "Computational effect predictor",
      "Cancer", "Sperm Nanoseq", "Buccal Nanoseq", "MAVE"
    )))
}

feature_order_MAVE <- c(new_CEP_names,
                        "n_COSMIC", "n_COSMIC_codon_count",
                        "n_sperm", "n_sperm_codon_count",
                        "n_buccal", "n_buccal_codon_count",
                        "MAVE_score")
```

```{r compute-aucs}
results      <- calc_aucs(clinvar_biobank_df_MAVE, "Clinvar_pathogenic") %>% colour_features()
results_long <- calc_aucs(clinvar_long_df_MAVE,   "Clinvar_pathogenic") %>% colour_features()
results_DNM  <- calc_aucs(DNM_biobank_df_MAVE,    "DNM_pathogenic")     %>% colour_features()
```

```{r auc-plot-fn}
auc_plot <- function(df, x_label) {
  ggplot(df, aes(y = factor(feature, levels = rev(feature_order_MAVE)),
                 x = AUC, colour = colour)) +
    geom_point(size = 2) +
    geom_errorbar(aes(xmin = CI_lower, xmax = CI_upper), width = 0.5) +
    labs(title = "Value of somatic data in DDX3X germline variant impact prediction",
         x = x_label, y = "", colour = "") +
    theme_minimal()
}
```

## ClinVar unique vs Biobank

```{r}
auc_plot(results, "AUC — ClinVar (likely) pathogenic vs Biobank")
```

## ClinVar long vs Biobank

```{r}
auc_plot(results_long, "AUC — ClinVar all vs Biobank")
```

## DNM long vs Biobank

```{r}
auc_plot(results_DNM, "AUC — DNM vs Biobank")
```

# Export plot data

```{r}
write_csv(results,
          "../../Data/MAVE_benchmarking/DDX3X_MAVE_plot_data_clinvar_unique_published_data.csv")
write_csv(results_long,
          "../../Data/MAVE_benchmarking/DDX3X_MAVE_plot_data_clinvar_long_published_data.csv")
write_csv(results_DNM,
          "../../Data/MAVE_benchmarking/DDX3X_MAVE_plot_data_DNM_long_published_data.csv")
```
