---
title: "FGFR benchmarking: public data"
output:
  html_document:
    df_print: paged
    code_folding: hide
  pdf_document: default
date: "2026-02-20"
---

# Aim

Determine the relative utility of MAVE data, somatic data and CEPs for clinical variant
interpretation in FGFR proteins using TwinsUK sperm sequencing and 31k DNM cohort data.

Subset to FGFR1-3 (FGFR4 is not a known DD gene and has no benchmarking data).

FGFR1-4 MAVE: https://doi.org/10.1038/s41588-025-02431-8

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(DT)
library(janitor)
library(patchwork)
library(pROC)

dir.create("../../Data/MAVE_benchmarking", recursive = TRUE, showWarnings = FALSE)

palette1 <- c("#FF8080", "#804000", "#6699CC")

fgfr_genes <- c("FGFR1", "FGFR2", "FGFR3")

read_fgfr <- function(path) {
  read_tsv(path, show_col_types = FALSE) %>%
    filter(gene %in% fgfr_genes) %>%
    dplyr::rename(
      n_sperm             = n_sperm_twinsUK,
      n_sperm_codon_count = n_sperm_twinsUK_codon_count
    )
}
```

```{r, message=FALSE, warning=FALSE}
FGFR_MAVE <- read_csv(
  "../../Data/MAVE/FGFR_MAVE_Diederichs/Diederichs_FGFR_supplement_2025.csv",
  show_col_types = FALSE
) %>%
  clean_names() %>%
  dplyr::select(fgfr, aa_mut, median_activation, category_activation, category_lof,
                p_value_activation, q_value_activation, discovery_activation) %>%
  dplyr::rename(gene = fgfr)

DNMs_31k         <- read_fgfr("../../Data/Variant_lists/DNMs_31k_missense_annot_240426_public.tsv.gz")
Clinvar_filt_path   <- read_fgfr("../../Data/Variant_lists/Clinvar_DDG2P_missense_path_annot_240426_public.tsv.gz")
Clinvar_filt_vus    <- read_fgfr("../../Data/Variant_lists/Clinvar_DDG2P_missense_VUS_annot_240426_public.tsv.gz")
Clinvar_filt_benign <- read_fgfr("../../Data/Variant_lists/Clinvar_DDG2P_missense_benign_annot_240426_public.tsv.gz")
Biobank_benign      <- read_fgfr("../../Data/Variant_lists/Biobank_missense_variants_annot_240426_public.tsv.gz")

# Raw ClinVar data to retrieve phenotype labels
clinvar <- read_tsv(
  "../../Data/Clinvar_060225/Clinvar_build_38_DDG2P_germline.tsv.gz",
  show_col_types = FALSE
)

clinvar_FGFR <- clinvar %>% filter(GeneSymbol %in% fgfr_genes)

clinvar_FGFR_path <- clinvar_FGFR %>%
  filter(ClinicalSignificance %in%
           c("Pathogenic", "Likely pathogenic", "Pathogenic/Likely pathogenic"))
```

```{r, message=FALSE, warning=FALSE}
FGFR_MAVE_for_join <- FGFR_MAVE %>%
  mutate(codon_change = str_replace(aa_mut, "([A-Z])([0-9]+)([A-Z])", "\\2\\1>\\2\\3")) %>%
  dplyr::select(gene, codon_change, median_activation, p_value_activation, q_value_activation) %>%
  group_by(gene, codon_change) %>%
  summarise(
    MAVE_activation    = max(median_activation),
    p_value_activation = min(p_value_activation),
    q_value_activation = min(q_value_activation),
    .groups = "drop"
  )

benchmarking_dat <- bind_rows(
  DNMs_31k            %>% mutate(source = "DNMs",              MutPred2_score = as.numeric(MutPred2_score)),
  Clinvar_filt_path   %>% mutate(source = "Clinvar_pathogenic", MutPred2_score = as.numeric(MutPred2_score)),
  Clinvar_filt_vus    %>% mutate(source = "Clinvar_VUS",        MutPred2_score = as.numeric(MutPred2_score)),
  Clinvar_filt_benign %>% mutate(source = "Clinvar_benign",     MutPred2_score = as.numeric(MutPred2_score)),
  Biobank_benign      %>% mutate(source = "Biobank_benign",     MutPred2_score = as.numeric(MutPred2_score))
)

benchmarking_dat_annot <- benchmarking_dat %>%
  left_join(FGFR_MAVE_for_join, by = c("gene", "codon_change"))
```

Remove hypogonadism/anosmia entries (LOF phenotypes not applicable to GOF analysis):

```{r}
LOF_clinvar_IDs_to_remove <- clinvar_FGFR_path %>%
  filter(str_detect(PhenotypeList, "hypog|Hypog|anosmia")) %>%
  mutate(variant_ID = paste0("chr", Chromosome, "_", PositionVCF, "_",
                              ReferenceAlleleVCF, "_", AlternateAlleleVCF)) %>%
  pull(variant_ID)

benchmarking_dat_annot_filt <- benchmarking_dat_annot %>%
  mutate(variant_ID = paste(chrom, pos, ref, alt, sep = "_")) %>%
  filter(!variant_ID %in% LOF_clinvar_IDs_to_remove) %>%
  dplyr::select(-variant_ID)
```

# MAVE data summary categories

```{r}
FGFR_MAVE %>% count(category_activation, category_lof) %>% arrange(category_activation == "Neutral")
```

# MAVE coverage

```{r}
benchmarking_dat_annot %>%
  count(source, MAVE_covered = !is.na(MAVE_activation)) %>%
  pivot_wider(names_from = MAVE_covered, values_from = n,
              names_prefix = "MAVE_") %>%
  dplyr::rename(not_covered = MAVE_FALSE, covered = MAVE_TRUE) %>%
  mutate(percent_covered = covered / (covered + not_covered) * 100)
```

## Coverage by gene

```{r}
benchmarking_dat_annot %>%
  filter(source %in% c("Clinvar_pathogenic", "Biobank_benign")) %>%
  group_by(gene, source) %>%
  count(MAVE_covered = !is.na(MAVE_activation)) %>%
  pivot_wider(names_from = MAVE_covered, values_from = n,
              names_prefix = "MAVE_") %>%
  dplyr::rename(not_covered = MAVE_FALSE, covered = MAVE_TRUE) %>%
  mutate(percent_covered = covered / (covered + not_covered) * 100)
```

# MAVE vs ClinVar pathogenicity

## Before LOF entries removed

```{r}
benchmarking_dat_annot %>%
  filter(source %in% c("Clinvar_pathogenic", "Biobank_benign"),
         !is.na(MAVE_activation)) %>%
  ggplot(aes(x = source, y = MAVE_activation)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  scale_y_log10() +
  theme_minimal() +
  labs(title = "FGFR MAVE activation scores by ClinVar pathogenicity",
       x = "", y = "MAVE activation")
```

## After LOF entries removed

```{r}
benchmarking_dat_annot_filt %>%
  filter(source %in% c("Clinvar_pathogenic", "Biobank_benign"),
         !is.na(MAVE_activation)) %>%
  ggplot(aes(x = source, y = MAVE_activation)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  scale_y_log10() +
  theme_minimal() +
  labs(title = "FGFR MAVE activation scores (hypog/anosmia entries removed)",
       x = "", y = "MAVE activation")
```

## DNMs vs Biobank

```{r}
benchmarking_dat_annot %>%
  filter(source %in% c("DNMs", "Biobank_benign"), !is.na(MAVE_activation)) %>%
  ggplot(aes(x = source, y = MAVE_activation)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  scale_y_log10() +
  theme_minimal() +
  labs(title = "FGFR MAVE activation scores: DNMs vs Biobank",
       x = "", y = "MAVE activation")
```

# AUC: ClinVar pathogenic FGFR variants vs Biobank {.tabset}

Three comparisons:
1. ClinVar unfiltered (includes hypog/anosmia)
2. ClinVar filtered to GOF variants
3. ClinVar filtered to GOF, MAVE-covered variants only (like-for-like)

```{r auc-helper}
predictors <- c("AlphaMissense_score", "REVEL_score", "popEVE",
                "CADD_raw", "PrimateAI_score",
                "n_COSMIC", "n_COSMIC_codon_count",
                "n_sperm", "n_sperm_codon_count",
                "MAVE_activation")

calc_roc_results <- function(df) {
  results <- data.frame()
  for (predictor in predictors) {
    df_noNA <- df %>% filter(!is.na(.data[[predictor]]))
    if (nrow(df_noNA) < 2 || n_distinct(df_noNA$label) < 2) next
    roc_obj <- pROC::roc(response = df_noNA$label, predictor = df_noNA[[predictor]],
                          levels = c(0, 1), direction = "<", quiet = TRUE)
    ci_val <- ci.auc(roc_obj)
    results <- rbind(results, data.frame(
      Predictor = predictor,
      AUC      = as.numeric(auc(roc_obj)),
      CI_lower = round(ci_val[1], 4),
      CI_upper = round(ci_val[3], 4)
    ))
  }
  results
}

prep_roc_df <- function(df) {
  df %>%
    mutate(label = case_when(
      source == "Clinvar_pathogenic" ~ 1,
      source == "Biobank_benign"     ~ 0,
      TRUE ~ NA_real_
    )) %>%
    mutate(label = factor(label, levels = c(0, 1))) %>%
    mutate(across(all_of(predictors), as.numeric)) %>%
    mutate(popEVE = -popEVE)
}
```

```{r compute-aucs, warning=FALSE}
roc_clinvar_1 <- benchmarking_dat_annot %>%
  filter(source %in% c("Clinvar_pathogenic", "Biobank_benign")) %>%
  prep_roc_df()

roc_clinvar_2 <- benchmarking_dat_annot_filt %>%
  filter(source %in% c("Clinvar_pathogenic", "Biobank_benign")) %>%
  prep_roc_df()

roc_clinvar_3 <- benchmarking_dat_annot_filt %>%
  filter(!is.na(MAVE_activation),
         source %in% c("Clinvar_pathogenic", "Biobank_benign")) %>%
  prep_roc_df()

roc_results_1 <- calc_roc_results(roc_clinvar_1)
roc_results_2 <- calc_roc_results(roc_clinvar_2)
roc_results_3 <- calc_roc_results(roc_clinvar_3)
```

## Split by dataset {.tabset}

### Unique form data

```{r, fig.height=4, fig.width=10}
plot_dat <- bind_rows(
  roc_results_1 %>% mutate(Dataset = "ClinVar unfiltered"),
  roc_results_2 %>% mutate(Dataset = "ClinVar filtered to GOF"),
  roc_results_3 %>% mutate(Dataset = "ClinVar filtered to GOF, MAVE-covered only")
)

plot_dat %>%
  mutate(fill_factor = case_when(
    Predictor %in% c("n_COSMIC", "n_COSMIC_codon_count",
                     "n_sperm", "n_sperm_codon_count") ~ "Somatic data",
    Predictor == "MAVE_activation"                      ~ "Functional data (MAVE)",
    TRUE ~ "Computational predictors"
  )) %>%
  mutate(Predictor = factor(Predictor, levels = predictors)) %>%
  ggplot(aes(x = Predictor, y = AUC, color = fill_factor, shape = Dataset)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2,
                position = position_dodge(width = 0.5)) +
  coord_flip() +
  theme_minimal() +
  labs(title = "AUC: ClinVar pathogenic FGFR variants vs Biobank",
       x = "", y = "AUC") +
  ylim(0, 1) +
  theme(legend.title = element_blank()) +
  scale_color_manual(values = palette1)
```

### Long form data

Not done

## Split by gene {.tabset}

### Unique form data

```{r, warning=FALSE}
roc_results_by_gene <- lapply(setNames(fgfr_genes, fgfr_genes), function(g) {
  roc_gene <- roc_clinvar_2 %>% filter(gene == g)
  calc_roc_results(roc_gene) %>% mutate(Gene = g)
})

roc_results_by_gene_table <- bind_rows(roc_results_by_gene) %>%
  mutate(Predictor = factor(Predictor, levels = predictors))
```

```{r, fig.height=4, fig.width=15}
roc_results_by_gene_table %>%
  mutate(fill_factor = case_when(
    Predictor %in% c("n_COSMIC", "n_COSMIC_codon_count",
                     "n_sperm", "n_sperm_codon_count") ~ "Somatic data",
    Predictor == "MAVE_activation"                      ~ "Functional data (MAVE)",
    TRUE ~ "Computational predictors"
  )) %>%
  ggplot(aes(x = Predictor, y = AUC, color = fill_factor)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  labs(title = "AUC: ClinVar pathogenic FGFR (GOF) vs Biobank, by gene",
       x = "", y = "AUC") +
  ylim(0, 1) +
  theme(legend.title = element_blank()) +
  facet_wrap(~ Gene) +
  scale_color_manual(values = palette1)
```

### Long form data

```{r, warning=FALSE}
roc_clinvar_long <- bind_rows(
  benchmarking_dat_annot_filt %>% filter(source == "Biobank_benign"),
  benchmarking_dat_annot_filt %>%
    filter(source == "Clinvar_pathogenic") %>%
    mutate(NumberSubmitters_clean = coalesce(NumberSubmitters_clean, 1L)) %>%
    uncount(NumberSubmitters_clean, .remove = FALSE)
) %>%
  prep_roc_df()

roc_results_by_gene_long <- lapply(setNames(fgfr_genes, fgfr_genes), function(g) {
  roc_gene <- roc_clinvar_long %>% filter(gene == g)
  calc_roc_results(roc_gene) %>% mutate(Gene = g)
})

roc_results_by_gene_table_long <- bind_rows(roc_results_by_gene_long) %>%
  mutate(Predictor = factor(Predictor, levels = predictors))
```

```{r, fig.height=4, fig.width=15}
roc_results_by_gene_table_long %>%
  mutate(fill_factor = case_when(
    Predictor %in% c("n_COSMIC", "n_COSMIC_codon_count",
                     "n_sperm", "n_sperm_codon_count") ~ "Somatic data",
    Predictor == "MAVE_activation"                      ~ "Functional data (MAVE)",
    TRUE ~ "Computational predictors"
  )) %>%
  ggplot(aes(x = Predictor, y = AUC, color = fill_factor)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  labs(title = "AUC: ClinVar pathogenic FGFR (GOF) vs Biobank, long form, by gene",
       x = "", y = "AUC") +
  ylim(0, 1) +
  theme(legend.title = element_blank()) +
  facet_wrap(~ Gene) +
  scale_color_manual(values = palette1)
```

# Write out data

```{r}
write_csv(plot_dat,
          "../../Data/MAVE_benchmarking/FGFR_MAVE_plot_dat_public_data.csv")
write_csv(roc_results_by_gene_table,
          "../../Data/MAVE_benchmarking/FGFR_MAVE_plot_split_gene_dat_public_data.csv")
write_csv(roc_results_by_gene_table_long,
          "../../Data/MAVE_benchmarking/FGFR_MAVE_plot_split_gene_LONG_public_data.csv")
```
