---
title: "HRAS benchmarking: public data"
output:
  html_document:
    df_print: paged
    code_folding: hide
  pdf_document: default
date: "2026-02-20"
---

# Aim

Assess the predictive power of somatic data to distinguish benign from pathogenic HRAS
germline variants using published data only.

Data sources:
- ClinVar: (likely) pathogenic germline variants
- Biobank: variants seen in UK Biobank, excluding ClinVar (likely) pathogenic variants

All lists are subset to missense variants in the MANE select transcript, annotated
with somatic features.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(pROC)
library(janitor)
library(stringr)

dir.create("../../Data/MAVE_benchmarking", recursive = TRUE, showWarnings = FALSE)

read_gene <- function(path, gene_name) {
  read_tsv(path, show_col_types = FALSE) %>%
    filter(gene == gene_name) %>%
    dplyr::rename(
      n_sperm             = n_sperm_twinsUK,
      n_sperm_codon_count = n_sperm_twinsUK_codon_count
    )
}

HRAS_Biobank      <- read_gene("../../Data/Variant_lists/Biobank_missense_variants_annot_240426_public.tsv.gz",  "HRAS")
HRAS_clinvar_path <- read_gene("../../Data/Variant_lists/Clinvar_DDG2P_missense_path_annot_240426_public.tsv.gz","HRAS")
HRAS_DNMs         <- read_gene("../../Data/Variant_lists/DNMs_31k_missense_annot_240426_public.tsv.gz",          "HRAS")

HRAS_clinvar_phenotype_terms <- read_tsv(
  "../../Data/Clinvar_060225/Clinvar_build_38_DDG2P_germline_path_missense.tsv.gz",
  show_col_types = FALSE
) %>% filter(GeneSymbol == "HRAS")

```

```{r, message=FALSE, warning=FALSE}
desired_CEPs  <- c("PrimateAI_score", "VARITY_R_score", "CADD_raw",
                   "AlphaMissense_score", "REVEL_score")
new_CEP_names <- c("PrimateAI", "VARITY_R", "CADD_raw", "AlphaMissense", "REVEL")

variant_cols <- c("gene", "codon_change",
                  "REVEL_score", "AlphaMissense_score", "CADD_raw",
                  "VARITY_R_score", "PrimateAI_score",
                  "n_COSMIC", "n_COSMIC_codon_count",
                  "n_sperm", "n_sperm_codon_count",
                  "n_buccal", "n_buccal_codon_count",
                  "hotcodon_cancer", "group")

clean_CEPs <- function(df) {
  df %>%
    mutate(across(where(is.character) & all_of(desired_CEPs), ~ na_if(., "."))) %>%
    mutate(across(all_of(desired_CEPs), as.numeric)) %>%
    rename_with(~ str_remove(., "_score$"), all_of(desired_CEPs))
}

pre_cols <- c(setdiff(variant_cols, "group"), "NumberSubmitters_clean", "n_DNM")

clinvar_biobank_df <- bind_rows(
  HRAS_clinvar_path %>% dplyr::select(any_of(pre_cols)) %>% mutate(group = "Clinvar_pathogenic"),
  HRAS_Biobank      %>% dplyr::select(any_of(pre_cols)) %>% mutate(group = "Biobank")
) %>% dplyr::select(all_of(variant_cols)) %>% clean_CEPs()

clinvar_long_biobank_df <- bind_rows(
  HRAS_clinvar_path %>%
    dplyr::select(any_of(pre_cols)) %>%
    mutate(NumberSubmitters_clean = coalesce(NumberSubmitters_clean, 1L)) %>%
    uncount(NumberSubmitters_clean) %>%
    mutate(group = "Clinvar_pathogenic"),
  HRAS_Biobank %>% dplyr::select(any_of(pre_cols)) %>% mutate(group = "Biobank")
) %>% dplyr::select(all_of(variant_cols)) %>% clean_CEPs()
```

```{r, message=FALSE, warning=FALSE}
raw <- read_csv(
  "../../Data/MAVE/HRAS_MAVE_Bandaru_2017/elife-27810-supp1_Regulated_RAS.csv",
  col_names = FALSE, show_col_types = FALSE
)

codon_numbers <- as.character(unlist(raw[1, ], use.names = FALSE))
wt_by_column  <- as.character(unlist(raw[2, ], use.names = FALSE))

MAVE_data <- raw[-c(1, 2), ] %>%
  dplyr::rename(mut_aa = X1) %>%
  pivot_longer(cols = -mut_aa, names_to = "col_id", values_to = "score") %>%
  mutate(
    col_index = as.integer(str_remove(col_id, "X")),
    codon     = codon_numbers[col_index],
    wt_aa     = wt_by_column[col_index],
    change    = paste0(codon, wt_aa, ">", codon, mut_aa)
  ) %>%
  filter(!is.na(score), !is.na(mut_aa), !is.na(codon), !is.na(wt_aa)) %>%
  dplyr::select(change, score) %>%
  dplyr::rename(MAVE_score = score, codon_change = change) %>%
  mutate(MAVE_score = as.numeric(MAVE_score))
```

# Number of HRAS variants

```{r}
cat("n (likely) pathogenic ClinVar variants:", nrow(HRAS_clinvar_path), "\n")
cat("n ClinVar long form variants:", nrow(clinvar_long_biobank_df %>% filter(group == "Clinvar_pathogenic")), "\n")
cat("n Biobank variants:", nrow(HRAS_Biobank), "\n")
cat("n unique DNMs (31k):", nrow(HRAS_DNMs), "\n")
cat("n total DNMs (31k):", sum(HRAS_DNMs$n_DNM, na.rm = TRUE), "\n")
```

# MAVE data

```{r}
clinvar_biobank_df_MAVE  <- clinvar_biobank_df      %>% left_join(MAVE_data, by = "codon_change")
clinvar_long_df_MAVE     <- clinvar_long_biobank_df %>% left_join(MAVE_data, by = "codon_change")
```

## Percent missingness

```{r}
clinvar_biobank_df_MAVE %>% summarise(percent_missing = mean(is.na(MAVE_score)) * 100)
clinvar_biobank_df_MAVE %>% filter(group == "Clinvar_pathogenic") %>%
  summarise(n_with_MAVE = sum(!is.na(MAVE_score)))
```

# AUC of somatic and MAVE data {.tabset}

```{r auc-helper}
features_to_test <- c(new_CEP_names,
                      "n_COSMIC", "n_COSMIC_codon_count",
                      "n_sperm", "n_sperm_codon_count",
                      "n_buccal", "n_buccal_codon_count",
                      "MAVE_score")

calc_aucs <- function(df, path_level) {
  results <- data.frame(feature = character(), AUC = numeric(),
                        CI_lower = numeric(), CI_upper = numeric(),
                        stringsAsFactors = FALSE)
  for (feat in features_to_test) {
    roc_obj <- roc(df$group, df[[feat]], levels = c("Biobank", path_level), quiet = TRUE)
    ci_val  <- ci.auc(roc_obj)
    results <- rbind(results, data.frame(feature  = feat,
                                         AUC      = as.numeric(auc(roc_obj)),
                                         CI_lower = ci_val[1],
                                         CI_upper = ci_val[3]))
  }
  results
}

colour_features <- function(df) {
  df %>%
    mutate(colour = case_when(
      feature %in% c("n_COSMIC", "n_COSMIC_codon_count") ~ "Cancer",
      feature %in% c("n_sperm", "n_sperm_codon_count")   ~ "Sperm Nanoseq",
      feature %in% c("n_buccal", "n_buccal_codon_count") ~ "Buccal Nanoseq",
      feature == "MAVE_score"                            ~ "MAVE",
      TRUE ~ "Computational effect predictor"
    )) %>%
    mutate(colour = factor(colour, levels = c(
      "Computational effect predictor",
      "Cancer", "Sperm Nanoseq", "Buccal Nanoseq", "MAVE"
    )))
}

feature_order_MAVE <- c(new_CEP_names,
                        "n_COSMIC", "n_COSMIC_codon_count",
                        "n_sperm", "n_sperm_codon_count",
                        "n_buccal", "n_buccal_codon_count",
                        "MAVE_score")
```

```{r compute-aucs}
results3       <- calc_aucs(clinvar_biobank_df_MAVE, "Clinvar_pathogenic") %>% colour_features()
results_long_3 <- calc_aucs(clinvar_long_df_MAVE,   "Clinvar_pathogenic") %>% colour_features()
```

## ClinVar unique vs Biobank

```{r}
ggplot(results3, aes(y = factor(feature, levels = rev(feature_order_MAVE)),
                     x = AUC, colour = colour)) +
  geom_point(size = 2) +
  geom_errorbar(aes(xmin = CI_lower, xmax = CI_upper), width = 0.5) +
  labs(title = "Value of somatic data in HRAS germline variant impact prediction",
       x = "AUC — ClinVar (likely) pathogenic vs Biobank", y = "", colour = "") +
  theme_minimal()
```

## ClinVar long vs Biobank

```{r}
ggplot(results_long_3, aes(y = factor(feature, levels = rev(feature_order_MAVE)),
                            x = AUC, colour = colour)) +
  geom_point(size = 2) +
  geom_errorbar(aes(xmin = CI_lower, xmax = CI_upper), width = 0.5) +
  labs(title = "Value of somatic data in HRAS germline variant impact prediction",
       x = "AUC — ClinVar all vs Biobank", y = "", colour = "") +
  theme_minimal()
```

# Export plot data

```{r}
write_csv(results3,
          "../../Data/MAVE_benchmarking/HRAS_MAVE_plot_data_clinvar_unique_published_data.csv")
write_csv(results_long_3,
          "../../Data/MAVE_benchmarking/HRAS_MAVE_plot_data_clinvar_long_published_data.csv")
```
