---
title: "PTEN benchmarking: public data"
output:
  html_document:
    df_print: paged
    code_folding: hide
  pdf_document: default
date: "2026-02-20"
---

# Aim

Assess the predictive power of somatic data to distinguish benign from pathogenic PTEN
germline variants using published data only.

Data sources:
- DNMs: trio WES/WGS (31k cohort)
- ClinVar: (likely) pathogenic germline variants
- Biobank: variants seen in UK Biobank, excluding DNMs and ClinVar (likely) pathogenic

All lists are subset to missense variants in the MANE select transcript, annotated
with somatic features.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(pROC)
library(janitor)

dir.create("../../Data/MAVE_benchmarking", recursive = TRUE, showWarnings = FALSE)

read_gene <- function(path, gene_name) {
  read_tsv(path, show_col_types = FALSE) %>%
    filter(gene == gene_name) %>%
    dplyr::rename(
      n_sperm             = n_sperm_twinsUK,
      n_sperm_codon_count = n_sperm_twinsUK_codon_count
    )
}

PTEN_Biobank      <- read_gene("../../Data/Variant_lists/Biobank_missense_variants_annot_240426_public.tsv.gz",  "PTEN")
PTEN_clinvar_path <- read_gene("../../Data/Variant_lists/Clinvar_DDG2P_missense_path_annot_240426_public.tsv.gz","PTEN")
PTEN_DNMs         <- read_gene("../../Data/Variant_lists/DNMs_31k_missense_annot_240426_public.tsv.gz",          "PTEN")

PTEN_clinvar_phenotype_terms <- read_tsv(
  "../../Data/Clinvar_060225/Clinvar_build_38_DDG2P_germline_path_missense.tsv.gz",
  show_col_types = FALSE
) %>% filter(GeneSymbol == "PTEN")

```

```{r, message=FALSE, warning=FALSE}
desired_CEPs  <- c("PrimateAI_score", "VARITY_R_score", "CADD_raw",
                   "AlphaMissense_score", "REVEL_score")
new_CEP_names <- c("PrimateAI", "VARITY_R", "CADD_raw", "AlphaMissense", "REVEL")

variant_cols <- c("gene", "codon_change",
                  "REVEL_score", "AlphaMissense_score", "CADD_raw",
                  "VARITY_R_score", "PrimateAI_score",
                  "n_COSMIC", "n_COSMIC_codon_count",
                  "n_sperm", "n_sperm_codon_count",
                  "n_buccal", "n_buccal_codon_count",
                  "hotcodon_cancer", "group")

clean_CEPs <- function(df) {
  df %>%
    mutate(across(where(is.character) & all_of(desired_CEPs), ~ na_if(., "."))) %>%
    mutate(across(all_of(desired_CEPs), as.numeric)) %>%
    rename_with(~ str_remove(., "_score$"), all_of(desired_CEPs))
}

pre_cols <- c(setdiff(variant_cols, "group"), "NumberSubmitters_clean", "n_DNM")

clinvar_biobank_df <- bind_rows(
  PTEN_clinvar_path %>% dplyr::select(any_of(pre_cols)) %>% mutate(group = "Clinvar_pathogenic"),
  PTEN_Biobank      %>% dplyr::select(any_of(pre_cols)) %>% mutate(group = "Biobank")
) %>% dplyr::select(all_of(variant_cols)) %>% clean_CEPs()

clinvar_long_biobank_df <- bind_rows(
  PTEN_clinvar_path %>%
    dplyr::select(any_of(pre_cols)) %>%
    mutate(NumberSubmitters_clean = coalesce(NumberSubmitters_clean, 1L)) %>%
    uncount(NumberSubmitters_clean) %>%
    mutate(group = "Clinvar_pathogenic"),
  PTEN_Biobank %>% dplyr::select(any_of(pre_cols)) %>% mutate(group = "Biobank")
) %>% dplyr::select(all_of(variant_cols)) %>% clean_CEPs()

DNM_biobank_df <- bind_rows(
  PTEN_DNMs %>%
    dplyr::select(any_of(pre_cols)) %>%
    mutate(n_DNM = coalesce(n_DNM, 1L)) %>%
    uncount(n_DNM) %>%
    mutate(group = "DNM_pathogenic"),
  PTEN_Biobank %>% dplyr::select(any_of(pre_cols)) %>% mutate(group = "Biobank")
) %>% dplyr::select(all_of(variant_cols)) %>% clean_CEPs()
```

```{r, message=FALSE, warning=FALSE}
activity_MAVE  <- read_csv("../../Data/MAVE/Fowler_PTEN_analysis/Mighell_PTEN_phosphatase_data.csv",
                            show_col_types = FALSE) %>% clean_names()
abundance_MAVE <- read_tsv("../../Data/MAVE/Fowler_PTEN_analysis/13073_2021_984_MOESM2_ESM.tsv",
                            show_col_types = FALSE)
```

# Number of PTEN variants

```{r}
cat("n (likely) pathogenic ClinVar variants:", nrow(PTEN_clinvar_path), "\n")
cat("n Biobank variants:", nrow(PTEN_Biobank), "\n")
cat("n unique DNMs (31k):", nrow(PTEN_DNMs), "\n")
cat("n total DNMs (31k):", sum(PTEN_DNMs$n_DNM, na.rm = TRUE), "\n")
```

# Phenotypes in ClinVar

```{r}
PTEN_clinvar_phenotype_terms %>% count(PhenotypeList) %>%
  arrange(desc(n)) %>% dplyr::select(n, PhenotypeList)
```

# MAVE data

```{r}
abundance_MAVE_2 <- abundance_MAVE %>%
  mutate(codon_change = paste0(position, start, ">", position, end)) %>%
  dplyr::select(codon_change, score_orig) %>%
  dplyr::rename(abundance_MAVE = score_orig) %>%
  filter(!is.na(abundance_MAVE), !is.na(codon_change))

activity_MAVE_2 <- activity_MAVE %>%
  extract(variant_one_letter,
          into = c("ref", "pos", "alt"),
          regex = "([A-Z])(\\d+)([A-Z])",
          remove = FALSE) %>%
  mutate(codon_change = str_c(pos, ref, ">", pos, alt)) %>%
  dplyr::select(codon_change, cum_score) %>%
  dplyr::rename(activity_MAVE = cum_score) %>%
  filter(!is.na(codon_change), !is.na(activity_MAVE))

clinvar_biobank_df_MAVE <- clinvar_biobank_df %>%
  left_join(activity_MAVE_2, by = "codon_change") %>%
  left_join(abundance_MAVE_2, by = "codon_change")

clinvar_long_df_MAVE <- clinvar_long_biobank_df %>%
  left_join(activity_MAVE_2, by = "codon_change") %>%
  left_join(abundance_MAVE_2, by = "codon_change")

DNM_biobank_df_MAVE <- DNM_biobank_df %>%
  left_join(activity_MAVE_2, by = "codon_change") %>%
  left_join(abundance_MAVE_2, by = "codon_change")
```

## Percent missingness

```{r}
bind_rows(
  clinvar_biobank_df_MAVE %>%
    pivot_longer(c(activity_MAVE, abundance_MAVE), names_to = "MAVE_type", values_to = "val") %>%
    group_by(group, MAVE_type) %>%
    summarise(percent_missing = mean(is.na(val)) * 100, .groups = "drop"),
  DNM_biobank_df_MAVE %>%
    filter(group == "DNM_pathogenic") %>%
    pivot_longer(c(activity_MAVE, abundance_MAVE), names_to = "MAVE_type", values_to = "val") %>%
    group_by(group, MAVE_type) %>%
    summarise(percent_missing = mean(is.na(val)) * 100, .groups = "drop")
) %>% pivot_wider(names_from = MAVE_type, values_from = percent_missing)
```

## MAVE score distributions {.tabset}

### ClinVar unique vs Biobank {.tabset}

#### Abundance

```{r, fig.width=3, fig.height=2}
clinvar_biobank_df_MAVE %>% filter(!is.na(abundance_MAVE)) %>%
  ggplot(aes(x = group, y = abundance_MAVE, fill = group)) +
  geom_boxplot() + theme_minimal() +
  labs(title = "PTEN MAVE abundance", x = "", y = "MAVE abundance") +
  theme(legend.position = "none")
```

#### Activity

```{r, fig.width=3, fig.height=2}
clinvar_biobank_df_MAVE %>% filter(!is.na(activity_MAVE)) %>%
  ggplot(aes(x = group, y = activity_MAVE, fill = group)) +
  geom_boxplot() + theme_minimal() +
  labs(title = "PTEN MAVE activity", x = "", y = "MAVE activity") +
  theme(legend.position = "none")
```

### DNM long vs Biobank {.tabset}

#### Abundance

```{r, fig.width=3, fig.height=2}
DNM_biobank_df_MAVE %>% filter(!is.na(abundance_MAVE)) %>%
  ggplot(aes(x = group, y = abundance_MAVE, fill = group)) +
  geom_boxplot() + theme_minimal() +
  labs(title = "PTEN MAVE abundance", x = "", y = "MAVE abundance") +
  theme(legend.position = "none")
```

#### Activity

```{r, fig.width=3, fig.height=2}
DNM_biobank_df_MAVE %>% filter(!is.na(activity_MAVE)) %>%
  ggplot(aes(x = group, y = activity_MAVE, fill = group)) +
  geom_boxplot() + theme_minimal() +
  labs(title = "PTEN MAVE activity", x = "", y = "MAVE activity") +
  theme(legend.position = "none")
```

# AUC of somatic and MAVE data {.tabset}

```{r auc-helper}
features_to_test <- c(new_CEP_names,
                      "n_COSMIC", "n_COSMIC_codon_count",
                      "n_sperm", "n_sperm_codon_count",
                      "n_buccal", "n_buccal_codon_count",
                      "activity_MAVE", "abundance_MAVE")

calc_aucs <- function(df, path_level) {
  results <- data.frame(feature = character(), AUC = numeric(),
                        CI_lower = numeric(), CI_upper = numeric(),
                        stringsAsFactors = FALSE)
  for (feat in features_to_test) {
    roc_obj <- roc(df$group, df[[feat]], levels = c("Biobank", path_level), quiet = TRUE)
    ci_val  <- ci.auc(roc_obj)
    results <- rbind(results, data.frame(feature  = feat,
                                         AUC      = as.numeric(auc(roc_obj)),
                                         CI_lower = ci_val[1],
                                         CI_upper = ci_val[3]))
  }
  results
}

colour_features <- function(df) {
  df %>%
    mutate(colour = case_when(
      feature %in% c("n_COSMIC", "n_COSMIC_codon_count") ~ "Cancer",
      feature %in% c("n_sperm", "n_sperm_codon_count")   ~ "Sperm Nanoseq",
      feature %in% c("n_buccal", "n_buccal_codon_count") ~ "Buccal Nanoseq",
      feature %in% c("activity_MAVE", "abundance_MAVE")  ~ "MAVE",
      TRUE ~ "Computational effect predictor"
    )) %>%
    mutate(colour = factor(colour, levels = c(
      "Computational effect predictor",
      "Cancer", "Sperm Nanoseq", "Buccal Nanoseq", "MAVE"
    )))
}

feature_order_MAVE <- c(new_CEP_names,
                        "n_COSMIC", "n_COSMIC_codon_count",
                        "n_sperm", "n_sperm_codon_count",
                        "n_buccal", "n_buccal_codon_count",
                        "activity_MAVE", "abundance_MAVE")
```

```{r compute-aucs}
results3       <- calc_aucs(clinvar_biobank_df_MAVE, "Clinvar_pathogenic") %>% colour_features()
results_long_3 <- calc_aucs(clinvar_long_df_MAVE,   "Clinvar_pathogenic") %>% colour_features()
results_DNM3   <- calc_aucs(DNM_biobank_df_MAVE,    "DNM_pathogenic")     %>% colour_features()
```

```{r auc-plot-fn}
auc_plot <- function(df, x_label) {
  ggplot(df, aes(y = factor(feature, levels = rev(feature_order_MAVE)),
                 x = AUC, colour = colour)) +
    geom_point(size = 2) +
    geom_errorbar(aes(xmin = CI_lower, xmax = CI_upper), width = 0.5) +
    labs(title = "Value of somatic data in PTEN germline variant impact prediction",
         x = x_label, y = "", colour = "") +
    theme_minimal()
}
```

## ClinVar unique vs Biobank

```{r}
auc_plot(results3, "AUC — ClinVar (likely) pathogenic vs Biobank")
```

## ClinVar long vs Biobank

```{r}
auc_plot(results_long_3, "AUC — ClinVar all vs Biobank")
```

## DNM long vs Biobank

```{r}
auc_plot(results_DNM3, "AUC — DNM vs Biobank")
```

# Export plot data

```{r}
write_csv(results3,
          "../../Data/MAVE_benchmarking/PTEN_MAVE_plot_data_clinvar_unique_published_data.csv")
write_csv(results_long_3,
          "../../Data/MAVE_benchmarking/PTEN_MAVE_plot_data_clinvar_long_published_data.csv")
write_csv(results_DNM3,
          "../../Data/MAVE_benchmarking/PTEN_MAVE_plot_data_DNM_long_published_data.csv")
```
