---
title: "PTPN11 variant benchmarking: public data"
output:
  html_document:
    df_print: paged
    code_folding: hide
  pdf_document: default
date: "2026-01-15"
---

# Aim

Determine the relative utility of MAVE data, somatic data and CEPs for clinical variant
interpretation in PTPN11 using publicly available data (TwinsUK sperm sequencing and 31k
DNM cohort).

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(DT)
library(janitor)
library(patchwork)
library(pROC)
library(cowplot)

dir.create("../../Data/MAVE_benchmarking", recursive = TRUE, showWarnings = FALSE)
dir.create("../../Data/Figures/Final",     recursive = TRUE, showWarnings = FALSE)

palette1 <- c("#FF8080", "#804000", "#6699CC")

read_ptpn11 <- function(path) {
  read_tsv(path, show_col_types = FALSE) %>%
    filter(gene == "PTPN11") %>%
    dplyr::rename(
      n_sperm             = n_sperm_twinsUK,
      n_sperm_codon_count = n_sperm_twinsUK_codon_count
    )
}
```

```{r, warning=FALSE, message=FALSE}
# Three-letter to one-letter amino acid map
aa3to1 <- c(
  Ala = "A", Arg = "R", Asn = "N", Asp = "D", Cys = "C",
  Glu = "E", Gln = "Q", Gly = "G", His = "H", Ile = "I",
  Leu = "L", Lys = "K", Met = "M", Phe = "F", Pro = "P",
  Ser = "S", Thr = "T", Trp = "W", Tyr = "Y", Val = "V"
)

vcep <- read_csv(
  "../../Data/RASopathy_VCEP/RASopathy_expert_var_curation.csv",
  show_col_types = FALSE
) %>%
  mutate(
    gene = str_extract(`Preferred Variant Title`, "(?<=\\()[A-Z0-9]+(?=\\))"),
    mutation = `Preferred Variant Title` %>%
      str_extract("\\(p\\.[^)]+\\)") %>%
      str_remove_all("\\(|\\)") %>%
      str_remove("^p\\.") %>%
      str_replace("([A-Za-z]{3})(\\d+)([A-Za-z]{3})", "\\2\\1>\\2\\3") %>%
      str_replace_all(aa3to1)
  )

vcep_path   <- vcep %>% filter(Classification %in% c("Pathogenic", "Likely pathogenic"))
vcep_benign <- vcep %>% filter(Classification %in% c("Benign", "Likely Benign"))
vcep_vus    <- vcep %>% filter(Classification == "Uncertain Significance")

PTPN11_MAVE_FL <- read_csv(
  "../../Data/MAVE/PTPN11_MAVE_Jiang_2025/PTPN11_MAVE_SHP2_FL_41467_2025_60641_MOESM8_ESM.csv",
  show_col_types = FALSE
) %>%
  clean_names() %>%
  dplyr::select(wt, pos, mut, enrichment_ave, cancer_high_freq) %>%
  mutate(codon_change = paste0(pos, wt, ">", pos, mut))

PTPN11_MAVE_PTP <- read_csv(
  "../../Data/MAVE/PTPN11_MAVE_Jiang_2025/PTPN11_MAVE_SHP2_PTP_41467_2025_60641_MOESM8_ESM.csv",
  show_col_types = FALSE
) %>%
  clean_names() %>%
  dplyr::select(wt, pos, mut, enrichment_ave) %>%
  mutate(codon_change = paste0(pos, wt, ">", pos, mut))

DNMs_31k <- read_tsv(
  "../../Data/Variant_lists/DNMs_31k_missense_annot_240426_public.tsv.gz",
  show_col_types = FALSE
)

clinvar_raw <- read_tsv(
  "../../Data/Clinvar_060225/Clinvar_build_38_DDG2P_germline.tsv.gz",
  show_col_types = FALSE
)
```

```{r, warning=FALSE, message=FALSE}
Clinvar_filt_path   <- read_ptpn11("../../Data/Variant_lists/Clinvar_DDG2P_missense_path_annot_240426_public.tsv.gz")
Clinvar_filt_vus    <- read_ptpn11("../../Data/Variant_lists/Clinvar_DDG2P_missense_VUS_annot_240426_public.tsv.gz")
Clinvar_filt_benign <- read_ptpn11("../../Data/Variant_lists/Clinvar_DDG2P_missense_benign_annot_240426_public.tsv.gz")
Biobank_benign      <- read_ptpn11("../../Data/Variant_lists/Biobank_missense_variants_annot_240426_public.tsv.gz")
```

# Defining NSML (LOF) vs Noonan (GOF) variants

NSML/LEOPARD mutations from https://doi.org/10.1159/000342251:

```{r}
NSML_mutations <- c(
  "279Y>279C", "279Y>279S", "461A>461T", "464G>464A",
  "468T>468M", "468T>468P", "498R>498W", "498R>498L",
  "506Q>506P", "510Q>510E", "510Q>510P"
)

Noonan_mutations <- vcep %>%
  filter(gene == "PTPN11",
         Classification %in% c("Pathogenic", "Likely Pathogenic"),
         Condition %in% c("Noonan syndrome", "RASopathy"),
         !is.na(mutation),
         !mutation %in% c("510Q>510P", "279Y>279C")) %>%
  pull(mutation)
```

## MAVE scores by pathogenicity category

```{r}
plot_clinvar <- bind_rows(
  Clinvar_filt_path   %>% dplyr::select(chrom, pos, ref, alt, NumberSubmitters_clean, codon_change) %>% mutate(category = "(likely)Pathogenic"),
  Clinvar_filt_vus    %>% dplyr::select(chrom, pos, ref, alt, NumberSubmitters_clean, codon_change) %>% mutate(category = "VUS/conflicting"),
  Clinvar_filt_benign %>% dplyr::select(chrom, pos, ref, alt, NumberSubmitters_clean, codon_change) %>% mutate(category = "(likely)Benign"),
  Biobank_benign      %>% dplyr::select(chrom, pos, ref, alt, codon_change)                  %>% mutate(category = "UK_Biobank")
) %>%
  left_join(PTPN11_MAVE_FL %>% dplyr::select(codon_change, enrichment_ave), by = "codon_change") %>%
  mutate(NSML_mutation = if_else(codon_change %in% NSML_mutations, "Yes", "No"),
         category = factor(category, levels = c("(likely)Pathogenic", "VUS/conflicting",
                                                 "(likely)Benign", "UK_Biobank")))

plot_vcep <- bind_rows(
  data.frame(codon_change = vcep_path  %>% filter(gene == "PTPN11") %>% pull(mutation), category = "(likely)Pathogenic"),
  data.frame(codon_change = vcep_vus   %>% filter(gene == "PTPN11") %>% pull(mutation), category = "VUS"),
  data.frame(codon_change = vcep_benign %>% filter(gene == "PTPN11") %>% pull(mutation), category = "(likely)Benign"),
  Biobank_benign %>% dplyr::select(chrom, pos, ref, alt, codon_change) %>% mutate(category = "UK_Biobank")
) %>%
  left_join(PTPN11_MAVE_FL %>% dplyr::select(codon_change, enrichment_ave), by = "codon_change") %>%
  mutate(NSML_mutation = if_else(codon_change %in% NSML_mutations, "Yes", "No"),
         category = factor(category, levels = c("(likely)Pathogenic", "VUS", "(likely)Benign", "UK_Biobank")))

plot_DNMs <- bind_rows(
  DNMs_31k %>%
    filter(grepl("missense", consequence), gene == "PTPN11") %>%
    dplyr::select(chrom, pos, ref, alt, n_DNM, codon_change) %>%
    mutate(category = "DNMs"),
  Biobank_benign %>%
    dplyr::select(chrom, pos, ref, alt, codon_change) %>%
    mutate(category = "UK_Biobank")
) %>%
  left_join(PTPN11_MAVE_FL %>% dplyr::select(codon_change, enrichment_ave), by = "codon_change") %>%
  mutate(NSML_mutation = if_else(codon_change %in% NSML_mutations, "Yes", "No"),
         category = factor(category, levels = c("DNMs", "UK_Biobank")))
```

### MAVE coverage

#### ClinVar and Biobank variants

```{r}
plot_clinvar %>%
  group_by(category) %>%
  summarise(n = n(), n_with_MAVE = sum(!is.na(enrichment_ave)),
            percent_with_MAVE = n_with_MAVE / n * 100)
```

#### VCEP variants

```{r}
plot_vcep %>%
  group_by(category) %>%
  summarise(n = n(), n_with_MAVE = sum(!is.na(enrichment_ave)),
            percent_with_MAVE = n_with_MAVE / n * 100)
```

#### DNM variants

```{r}
plot_DNMs %>%
  group_by(category) %>%
  summarise(n = n(), n_with_MAVE = sum(!is.na(enrichment_ave)),
            percent_with_MAVE = n_with_MAVE / n * 100)
```

### ClinVar vs Biobank

```{r}
plot_clinvar %>% filter(!is.na(enrichment_ave)) %>%
  ggplot(aes(x = category, y = enrichment_ave)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.7, aes(color = NSML_mutation)) +
  theme_minimal() +
  labs(title = "PTPN11 MAVE enrichment scores by ClinVar pathogenicity",
       x = "", y = "MAVE enrichment score (full length)")
```

### VCEP vs Biobank

```{r}
plot_vcep %>% filter(!is.na(enrichment_ave)) %>%
  ggplot(aes(x = category, y = enrichment_ave)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.7, aes(color = NSML_mutation)) +
  theme_minimal() +
  labs(title = "PTPN11 MAVE enrichment scores by VCEP pathogenicity",
       x = "", y = "MAVE enrichment score (full length)")
```

### DNMs vs Biobank

```{r}
plot_DNMs %>% filter(!is.na(enrichment_ave)) %>%
  ggplot(aes(x = category, y = enrichment_ave)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.7, aes(color = NSML_mutation)) +
  theme_minimal() +
  labs(title = "PTPN11 MAVE enrichment scores: DNMs vs Biobank",
       x = "", y = "MAVE enrichment score (full length)")
```

## AUC to distinguish ClinVar pathogenic PTPN11 variants from Biobank variants

Three variant set definitions:
1. All ClinVar (likely) pathogenic vs Biobank
2. Stringent: ClinVar Pathogenic / VCEP pathogenic vs (likely) benign + common Biobank (AF > 5e-6)
3. Stringent subset minus NSML mutations

```{r auc-helper}
predictors <- c("AlphaMissense_score", "REVEL_score", "popEVE",
                "CADD_raw", "PrimateAI_score",
                "n_COSMIC", "n_COSMIC_codon_count",
                "n_sperm", "n_sperm_codon_count",
                "enrichment_ave")

calc_roc_df <- function(df) {
  results <- data.frame()
  for (predictor in predictors) {
    df_noNA <- df %>% filter(!is.na(.data[[predictor]]))
    if (nrow(df_noNA) < 2 || n_distinct(df_noNA$label) < 2) next
    roc_obj <- roc(df_noNA$label, df_noNA[[predictor]], quiet = TRUE)
    ci_val <- ci.auc(roc_obj)
    results <- rbind(results, data.frame(
      Predictor = predictor,
      AUC      = as.numeric(auc(roc_obj)),
      CI_upper = round(ci_val[3], 4),
      CI_lower = round(ci_val[1], 4)
    ))
  }
  results
}

prep_roc <- function(df, path_level, benign_levels) {
  df %>%
    mutate(label = case_when(
      category == path_level                ~ 1,
      category %in% benign_levels           ~ 0,
      TRUE                                  ~ NA_real_
    )) %>%
    filter(!is.na(label)) %>%
    left_join(PTPN11_MAVE_FL %>% dplyr::select(codon_change, enrichment_ave),
              by = "codon_change") %>%
    mutate(NSML_mutation = if_else(codon_change %in% NSML_mutations, "Yes", "No"),
           label = factor(label, levels = c(0, 1))) %>%
    filter(!is.na(enrichment_ave)) %>%
    mutate(across(all_of(predictors), as.numeric))
}
```

```{r compute-aucs, warning=FALSE}
roc_df_base <- bind_rows(
  Clinvar_filt_path %>%
    dplyr::select(chrom, pos, ref, alt, codon_change, all_of(setdiff(predictors, "enrichment_ave"))) %>%
    mutate(category = "(likely)Pathogenic"),
  Biobank_benign %>%
    dplyr::select(chrom, pos, ref, alt, codon_change, all_of(setdiff(predictors, "enrichment_ave"))) %>%
    mutate(category = "UK_Biobank")
)

roc_clinvar       <- prep_roc(roc_df_base, "(likely)Pathogenic", "UK_Biobank")
roc_clinvar_no_NSML <- roc_clinvar %>% filter(NSML_mutation == "No")

Clinvar_path_stringent_ids <- clinvar_raw %>%
  filter(GeneSymbol == "PTPN11", ClinicalSignificance == "Pathogenic") %>%
  mutate(variant = paste0("chr", Chromosome, "_", PositionVCF, "_",
                          ReferenceAlleleVCF, "_", AlternateAlleleVCF)) %>%
  pull(variant)

roc_df_stringent <- bind_rows(
  Clinvar_filt_path %>%
    mutate(variant = paste(chrom, pos, ref, alt, sep = "_")) %>%
    filter(variant %in% Clinvar_path_stringent_ids) %>%
    dplyr::select(chrom, pos, ref, alt, codon_change,
                  all_of(setdiff(predictors, "enrichment_ave"))) %>%
    mutate(category = "Pathogenic"),
  Clinvar_filt_benign %>%
    dplyr::select(chrom, pos, ref, alt, codon_change,
                  all_of(setdiff(predictors, "enrichment_ave"))) %>%
    mutate(category = "(likely)Benign"),
  Biobank_benign %>%
    filter(AF > 0.000005) %>%
    dplyr::select(chrom, pos, ref, alt, codon_change,
                  all_of(setdiff(predictors, "enrichment_ave"))) %>%
    mutate(category = "UK_Biobank")
)

roc_clinvar_stringent        <- prep_roc(roc_df_stringent, "Pathogenic", c("(likely)Benign", "UK_Biobank"))
roc_clinvar_stringent_no_NSML <- roc_clinvar_stringent %>% filter(NSML_mutation == "No")

roc_results            <- calc_roc_df(roc_clinvar)
roc_results_no_NSML    <- calc_roc_df(roc_clinvar_no_NSML)
roc_results_stringent  <- calc_roc_df(roc_clinvar_stringent)
roc_results_stringent_no_NSML <- calc_roc_df(roc_clinvar_stringent_no_NSML)
```

```{r, fig.width=10, fig.height=4}
plot_dat <- bind_rows(
  roc_results_stringent_no_NSML %>% mutate(Dataset = "Stringent subset minus NSML"),
  roc_results_stringent          %>% mutate(Dataset = "Stringent subset"),
  roc_results                    %>% mutate(Dataset = "All variants")
)

plot_dat %>%
  mutate(fill_factor = case_when(
    Predictor %in% c("n_COSMIC", "n_COSMIC_codon_count",
                     "n_sperm", "n_sperm_codon_count") ~ "Somatic data",
    Predictor == "enrichment_ave"                       ~ "Functional data (MAVE)",
    TRUE ~ "Computational predictors"
  )) %>%
  mutate(Predictor = factor(Predictor, levels = predictors)) %>%
  ggplot(aes(x = Predictor, y = AUC, color = fill_factor, shape = Dataset)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2,
                position = position_dodge(width = 0.5)) +
  coord_flip() +
  theme_minimal() +
  labs(title = "AUC: ClinVar pathogenic PTPN11 variants vs Biobank",
       x = "", y = "AUC") +
  ylim(0, 1) +
  theme(legend.title = element_blank()) +
  scale_color_manual(values = c("Somatic data" = "#6699CC",
                                "Functional data (MAVE)" = "#804000",
                                "Computational predictors" = "#FF8080"))
```

### Unique vs long form

```{r, warning=FALSE}
roc_clinvar_long <- bind_rows(
  Clinvar_filt_path %>%
    dplyr::select(chrom, pos, ref, alt, codon_change,
                  all_of(setdiff(predictors, "enrichment_ave")),
                  NumberSubmitters_clean) %>%
    mutate(NumberSubmitters_clean = coalesce(NumberSubmitters_clean, 1L)) %>%
    uncount(NumberSubmitters_clean) %>%
    mutate(category = "(likely)Pathogenic"),
  Biobank_benign %>%
    dplyr::select(chrom, pos, ref, alt, codon_change,
                  all_of(setdiff(predictors, "enrichment_ave"))) %>%
    mutate(category = "UK_Biobank")
) %>%
  mutate(label = case_when(
    category == "(likely)Pathogenic" ~ 1,
    category == "UK_Biobank"         ~ 0,
    TRUE ~ NA_real_
  )) %>%
  filter(!is.na(label)) %>%
  left_join(PTPN11_MAVE_FL %>% dplyr::select(codon_change, enrichment_ave), by = "codon_change") %>%
  mutate(NSML_mutation = if_else(codon_change %in% NSML_mutations, "Yes", "No"),
         label = factor(label, levels = c(0, 1))) %>%
  filter(!is.na(enrichment_ave)) %>%
  mutate(across(all_of(predictors), as.numeric))

roc_results_long <- calc_roc_df(roc_clinvar_long)
```

```{r, fig.width=10, fig.height=5}
unique_vs_long <- bind_rows(
  roc_results      %>% mutate(Form = "Unique"),
  roc_results_long %>% mutate(Form = "Long")
)

unique_vs_long %>%
  mutate(fill_factor = case_when(
    Predictor %in% c("n_COSMIC", "n_COSMIC_codon_count",
                     "n_sperm", "n_sperm_codon_count") ~ "Somatic data",
    Predictor == "enrichment_ave"                       ~ "Functional data (MAVE)",
    TRUE ~ "Computational predictors"
  )) %>%
  mutate(Predictor = factor(Predictor, levels = predictors)) %>%
  ggplot(aes(x = Predictor, y = AUC, color = fill_factor)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2,
                position = position_dodge(width = 0.5)) +
  coord_flip() + facet_wrap(~ Form) +
  theme_minimal() +
  labs(title = "AUC: ClinVar pathogenic PTPN11 variants vs Biobank — unique vs long form",
       x = "", y = "AUC") +
  ylim(0, 1) +
  theme(legend.title = element_blank()) +
  scale_color_manual(values = c("Somatic data" = "#6699CC",
                                "Functional data (MAVE)" = "#804000",
                                "Computational predictors" = "#FF8080"))
```

# Write out data

```{r}
write_csv(plot_dat,
          "../../Data/MAVE_benchmarking/PTPN11_MAVE_plot_dat_public_data.csv")
write_csv(unique_vs_long,
          "../../Data/MAVE_benchmarking/PTPN11_MAVE_plot_dat_unique_long_public_data.csv")
```

# Appendix: Codon-level mutation distribution and MAVE scores

```{r appendix-data, message=FALSE, warning=FALSE}
# DATA_SOURCE: pre-processed per-codon mutation count files (from annotation pipeline)
# Expected paths when available:
#   Data/Clinvar_060225/Clinvar_path_DDG2P_codingcsq_for_gene_vis.tsv.gz
#   Data/Cancer_data/COSMIC_DDG2P_for_gene_vis.tsv.gz  (cluster resource)
#   Data/Variant_lists/DNM_list_for_gene_vis.tsv.gz
#   Data/Variant_lists/Sperm_seq_for_gene_vis.tsv.gz
#   Data/MAVE/PTPN11_MAVE_Jiang_2025/PTPN11_MAVE_FL_for_gene_vis.tsv.gz

appendix_ok <- tryCatch({
  clinvar_vis <<- readr::read_tsv(
    "../../Data/Clinvar_060225/Clinvar_path_DDG2P_codingcsq_for_gene_vis.tsv.gz",
    show_col_types = FALSE
  )
  COSMIC_vis <<- readr::read_tsv(
    "../../Data/Cancer_data/COSMIC_DDG2P_for_gene_vis.tsv.gz",
    show_col_types = FALSE
  )
  DNMs_vis <<- readr::read_tsv(
    "../../Data/Variant_lists/DNM_list_for_gene_vis.tsv.gz",
    show_col_types = FALSE
  )
  sperm_vis <<- readr::read_tsv(
    "../../Data/Variant_lists/Sperm_seq_for_gene_vis.tsv.gz",
    show_col_types = FALSE
  )
  MAVE_vis <<- readr::read_tsv(
    "../../Data/MAVE/PTPN11_MAVE_Jiang_2025/PTPN11_MAVE_FL_for_gene_vis.tsv.gz",
    show_col_types = FALSE
  )
  TRUE
}, error = function(e) {
  message("Appendix data not available locally — skipping appendix plots: ", conditionMessage(e))
  FALSE
})

missense_colors <- c("missense" = "#882255", "non_missense" = "#6699CC")
aa_levels <- c("A","R","N","D","C","Q","E","G","H","I",
               "L","K","M","F","P","S","T","W","Y","V","stop")
```

```{r appendix-helpers}
xaxis_codons <- function(max_codon) {
  if (is.na(max_codon) || !is.finite(max_codon) || max_codon <= 0) max_codon <- 1
  if (max_codon <= 600) {
    scale_x_continuous(limits = c(0, max_codon + 20),  breaks = seq(0, max_codon + 20,  by = 100))
  } else if (max_codon <= 1200) {
    scale_x_continuous(limits = c(0, max_codon + 50),  breaks = seq(0, max_codon + 50,  by = 250))
  } else {
    scale_x_continuous(limits = c(0, max_codon + 100), breaks = pretty(c(0, max_codon)))
  }
}

left_labeled_panel <- function(p, label, strip_width = 0.26, label_size = 0.9) {
  strip <- wrap_elements(
    full = grid::textGrob(label, rot = 0, x = unit(0.02, "npc"),
                          just = c("left", "centre"),
                          gp = grid::gpar(fontface = "bold", cex = label_size))
  )
  (strip | p) + plot_layout(widths = c(strip_width, 1))
}

mave_heatmap_plot <- function(MAVE_gene, max_codon) {
  mave_df <- MAVE_gene %>%
    transmute(codon = pos,
              mut   = factor(mut, levels = rev(aa_levels)),
              enrichment_ave) %>%
    tidyr::complete(codon = seq_len(max_codon),
                    mut   = factor(rev(aa_levels), levels = rev(aa_levels)))
  lim <- max(abs(mave_df$enrichment_ave), na.rm = TRUE)
  if (!is.finite(lim) || lim == 0) lim <- 1

  ggplot(mave_df, aes(x = codon, y = mut, fill = enrichment_ave)) +
    geom_tile(height = 0.95, width = 0.95) +
    xlab("Codon") + ylab(NULL) +
    xaxis_codons(max_codon) +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                         midpoint = 0, limits = c(-lim, lim), na.value = "grey80",
                         name = "MAVE") +
    theme_bw() +
    theme(panel.grid = element_blank(), axis.text.y = element_blank(),
          axis.ticks.y = element_blank(), axis.title.y = element_blank(),
          legend.position = "right")
}

generate_gene_panel_plots <- function(gene_symbol, clinvar, COSMIC, DNMs, sperm, MAVE,
                                      panel_labels = c("left", "none")) {
  panel_labels <- match.arg(panel_labels)
  clinvar_g <- clinvar %>% filter(gene == gene_symbol)
  COSMIC_g  <- COSMIC  %>% filter(gene == gene_symbol)
  DNMs_g    <- DNMs    %>% filter(gene == gene_symbol)
  sperm_g   <- sperm   %>% filter(gene == gene_symbol)

  max_codon <- suppressWarnings(max(c(
    if (nrow(clinvar_g) > 0) max(clinvar_g$codon, na.rm = TRUE) else -Inf,
    if (nrow(COSMIC_g)  > 0) max(COSMIC_g$codon,  na.rm = TRUE) else -Inf,
    if (nrow(DNMs_g)    > 0) max(DNMs_g$codon,    na.rm = TRUE) else -Inf,
    if (nrow(sperm_g)   > 0) max(sperm_g$codon,   na.rm = TRUE) else -Inf,
    if (nrow(MAVE)      > 0) max(MAVE$pos,         na.rm = TRUE) else -Inf
  ), na.rm = TRUE))
  if (!is.finite(max_codon) || max_codon <= 0) max_codon <- 1

  base_theme <- theme(plot.title = element_blank(), legend.position = "none",
                      axis.title.y = element_blank())

  make_bar_panel <- function(dat, y_col) {
    dat %>%
      mutate(missense = factor(missense, levels = c("missense", "non_missense"))) %>%
      ggplot(aes(x = codon, y = .data[[y_col]], fill = missense, colour = missense)) +
      geom_col(width = 1, linewidth = 0.25, na.rm = TRUE) + xlab("") +
      xaxis_codons(max_codon) +
      scale_fill_manual(values = missense_colors, breaks = c("missense", "non_missense"), drop = FALSE) +
      scale_colour_manual(values = missense_colors, breaks = c("missense", "non_missense"),
                          drop = FALSE, guide = "none") +
      base_theme
  }

  clinvar_plot <- make_bar_panel(clinvar_g, "NumberSubmitters_clean")
  dnm_plot     <- make_bar_panel(DNMs_g,   "n_DNM")
  cosmic_plot  <- make_bar_panel(COSMIC_g, "n_COSMIC")
  sperm_plot   <- make_bar_panel(sperm_g,  "n_sperm")
  MAVE_plot    <- mave_heatmap_plot(MAVE, max_codon) +
    theme(plot.margin = margin(5.5, 5.5, 5.5, 18))

  labels <- c("n ClinVar P/LP", "n DNM", "n COSMIC", "n Sperm nanoseq", "MAVE")

  if (panel_labels == "left") {
    combined <-
      left_labeled_panel(clinvar_plot, labels[1]) /
      left_labeled_panel(dnm_plot,     labels[2]) /
      left_labeled_panel(cosmic_plot,  labels[3]) /
      left_labeled_panel(sperm_plot,   labels[4]) /
      left_labeled_panel(MAVE_plot,    labels[5]) +
      plot_layout(heights = c(1, 1, 1, 1, 1.3))
  } else {
    combined <- clinvar_plot / dnm_plot / cosmic_plot / sperm_plot / MAVE_plot +
      plot_layout(heights = c(1, 1, 1, 1, 1.3))
  }

  gene_title <- wrap_elements(
    full = grid::textGrob(gene_symbol, x = unit(0, "npc"), just = c("left", "top"),
                          gp = grid::gpar(fontface = "bold", cex = 1.25))
  )
  gene_title / combined + plot_layout(heights = c(0.08, 1))
}
```

```{r ptpn11-codon-plot, fig.width=8, fig.height=8, warning=FALSE, eval=appendix_ok}
PTPN11_panel <- generate_gene_panel_plots(
  gene_symbol  = "PTPN11",
  clinvar      = clinvar_vis,
  COSMIC       = COSMIC_vis,
  DNMs         = DNMs_vis,
  sperm        = sperm_vis,
  MAVE         = PTPN11_MAVE_FL,
  panel_labels = "left"
)

PTPN11_panel

ggsave("../../Data/Figures/Final/PTPN11_MAVE_codon_plot.png",
       PTPN11_panel, width = 8, height = 8, dpi = 300)
```
