---
title: "Somatic germline paper inline numbers"
output:
  html_document:
    df_print: paged
    code_folding: hide
date: "2026-02-05"
---

# Aim
Compute all inline numerical claims made in the manuscript text. Sections follow the order of appearance in the manuscript.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(janitor)

date <- "240426"

recode_data_name <- function(x) {
  dplyr::case_match(x,
    "clinvar_long_uncontaminated" ~ "clinvar_count_weighted",
    "DNM_long_uncontaminated"     ~ "DNM_count_weighted",
    "clinvar_long"                ~ "clinvar_count_weighted",
    "DNM_long"                    ~ "DNM_count_weighted",
    .default = x
  )
}
```

```{r, warning=FALSE, include=FALSE}
Clinvar_vs_Biobank <- read_csv("../../Data/ML_model_input/Gene_group_models/Clinvar_vs_Biobank_unique_set1_240426.csv.gz")
DNM_vs_Biobank     <- read_csv("../../Data/ML_model_input/Gene_group_models/DNM_vs_Biobank_unique_set1_240426.csv.gz")
tmp_clinvar        <- read_tsv("../../Data/Variant_lists/Clinvar_DDG2P_missense_path_annot_240426_public.tsv.gz")
```

```{r, warning=FALSE, message=FALSE}
altfunc           <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_AltFunc.csv") %>% pull(gene_symbol)
altfunc_stringent <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_AltFunc_stringent.csv") %>% pull(gene_symbol)
lof               <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_LOF.csv") %>% pull(gene_symbol)
other             <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_other.csv") %>% pull(gene_symbol)
RASopathies       <- read_csv("../../Data/Gene_definitions/RASopathy_genes_VCEP_030425.csv", show_col_types = FALSE) %>% pull(gene)
altfunc_minus_RAS <- setdiff(altfunc, RASopathies)
macrocephaly      <- read_csv("../../Data/Gene_definitions/Macrocephaly_monoallelic_DD_somatic_genes.csv") %>% pull(gene)
```

```{r, warning=FALSE, include=FALSE}
AUCs_individual_features <- read_csv(paste0("../../Data/AUC_tables/Gene_group_models/AUC_table_gene_groups_single_features_", date, ".csv")) %>%
  mutate(data_name = recode_data_name(data_name))

AUCs_LR_model <- read_csv(paste0("../../Data/AUC_tables/Gene_group_models/AUC_table_gene_groups_LR_models_", date, ".csv")) %>%
  mutate(Predictor = paste0(cep, "_LR_model_", test_or_train, "_data"),
         data_name = recode_data_name(data_name))

AUCs <- bind_rows(AUCs_LR_model, AUCs_individual_features) %>%
  dplyr::select(Predictor, data_name, gene_group,
                AUC_full, AUC_full_lower, AUC_full_upper, everything()) %>%
  dplyr::select(-any_of("gene_set"))

n_genes <- read_csv(paste0("../../Data/AUC_tables/Gene_group_models/Gene_counts_per_group_", date, ".csv")) %>%
  mutate(data_name = recode_data_name(data_name))
n_vars  <- read_csv(paste0("../../Data/AUC_tables/Gene_group_models/Variant_counts_per_group_", date, ".csv")) %>%
  mutate(data_name = recode_data_name(data_name))
```

```{r, warning=FALSE, message=FALSE}
gene_AUCs_individual_features <- read_csv("../../Data/AUC_tables/Gene_group_hold_out_models/Single_feature_AUCs_individual_genes_240426.csv") %>%
  mutate(dataset = recode_data_name(dataset))
gene_AUCs_LR_model            <- read_csv("../../Data/AUC_tables/Gene_group_hold_out_models/AUC_table_gene_holdout_LR_models_240426.csv") %>%
  mutate(dataset = recode_data_name(dataset))
n_vars_per_gene               <- read_csv("../../Data/AUC_tables/Gene_group_hold_out_models/Variant_counts_per_gene_240426.csv")

n_vars_per_gene_clinvar <- n_vars_per_gene %>%
  filter(data_set %in% c("clinvar_mutation_data_unique", "clinvar_mutation_data_long")) %>%
  pivot_wider(names_from = data_set, values_from = c(n_path_variants, n_benign_variants)) %>%
  dplyr::select(-n_benign_variants_clinvar_mutation_data_long) %>%
  dplyr::rename(
    n_benign            = n_benign_variants_clinvar_mutation_data_unique,
    n_unique_pathogenic = n_path_variants_clinvar_mutation_data_unique,
    n_pathogenic        = n_path_variants_clinvar_mutation_data_long
  )
```

```{r, warning=FALSE, message=FALSE}
LR <- read_csv("../../Data/LR_calculations/LR_table.csv")
```

```{r, warning=FALSE, message=FALSE}
PTPN11_dat <- read_csv("../../Data/MAVE_benchmarking/PTPN11_MAVE_plot_dat_unique_long_public_data.csv")
PTEN_MAVE  <- read_csv("../../Data/MAVE_benchmarking/PTEN_MAVE_plot_data_clinvar_long_published_data.csv")
DDX3X_MAVE <- read_csv("../../Data/MAVE_benchmarking/DDX3X_MAVE_plot_data_clinvar_long_published_data.csv")
HRAS_MAVE  <- read_csv("../../Data/MAVE_benchmarking/HRAS_MAVE_plot_data_clinvar_long_published_data.csv")
```

```{r, warning=FALSE, message=FALSE}
PTPN11_MAVE_vis <- read_csv("../../Data/MAVE/PTPN11_MAVE_Jiang_2025/PTPN11_MAVE_SHP2_FL_41467_2025_60641_MOESM8_ESM.csv") %>%
  clean_names() %>%
  dplyr::select(wt, pos, mut, enrichment_ave)
```

```{r, warning=FALSE, message=FALSE}
VUS_acmg        <- read_tsv("../../Data/auto_ACMG_output/Clinvar_VUS_ACMG_240426.tsv")
VUS_acmg_somatic <- read_tsv("../../Data/auto_ACMG_output/Clinvar_VUS_ACMG_somatic_240426.tsv")
path_acmg        <- read_tsv("../../Data/auto_ACMG_output/Clinvar_path_ACMG_240426.tsv")
path_acmg_somatic <- read_tsv("../../Data/auto_ACMG_output/Clinvar_path_ACMG_somatic_240426.tsv")
benign_acmg      <- read_tsv("../../Data/auto_ACMG_output/Clinvar_benign_ACMG_240426.tsv")
benign_acmg_somatic <- read_tsv("../../Data/auto_ACMG_output/Clinvar_benign_ACMG_somatic_240426.tsv")
```

```{r}
# Per-gene somatic AUC summary (clinvar_count_weighted dataset)
genes_clinvar_long_auc <- gene_AUCs_individual_features %>%
  filter(dataset == "clinvar_count_weighted") %>%
  filter(Predictor %in% c("n_COSMIC_codon_count", "n_sperm_codon_count", "n_buccal_codon_count")) %>%
  group_by(gene) %>%
  summarise(max_AUC_lower = max(AUC_full_lower, na.rm = TRUE), .groups = "drop") %>%
  filter(is.finite(max_AUC_lower)) %>%
  mutate(
    auc_lower_06 = max_AUC_lower > 0.6,
    auc_lower_08 = max_AUC_lower > 0.8
  )

genes_clinvar_long <- genes_clinvar_long_auc$gene
```

---

# 1. Variant counts

```{r}
# Pathogenic variants in clinvar_count_weighted (including duplicate entries per gene count)
n_path_clinvar_cw <- n_vars %>%
  filter(data_name == "clinvar_count_weighted", gene_group == "all") %>%
  pull(n_path_variants)

# Pathogenic variants clinvar unique
n_path_clinvar_unique <- n_vars %>%
  filter(data_name == "clinvar_unique", gene_group == "all") %>%
  pull(n_path_variants)

# Benign (UK Biobank) variants
n_benign <- n_vars %>%
  filter(data_name == "clinvar_count_weighted", gene_group == "all") %>%
  pull(n_benign_variants)

# DNM variants (count weighted)
n_dnm_cw <- n_vars %>%
  filter(data_name == "DNM_count_weighted", gene_group == "all") %>%
  pull(n_path_variants)

# DNM variants (unique)
n_dnm_unique <- n_vars %>%
  filter(data_name == "DNM_unique", gene_group == "all") %>%
  pull(n_path_variants)

# Gene counts
n_genes_total <- n_genes %>%
  filter(data_name == "clinvar_count_weighted", gene_group == "all") %>%
  pull(gene_count)

cat(
  "Pathogenic variants (clinvar count weighted):", n_path_clinvar_cw, "\n",
  "Pathogenic variants (clinvar unique):", n_path_clinvar_unique, "\n",
  "Benign variants (UK Biobank):", n_benign, "\n",
  "DNM variants (count weighted):", n_dnm_cw, "\n",
  "DNM variants (unique):", n_dnm_unique, "\n",
  "Genes with clinvar data:", n_genes_total, "\n"
)
```

```{r}
# Number of genes in each gene group
cat(
  "RASopathy genes:", length(RASopathies), "\n",
  "AltFunc genes:", length(altfunc), "\n",
  "Non-RASopathy AltFunc:", length(altfunc_minus_RAS), "\n",
  "LOF genes:", length(lof), "\n",
  "Other genes:", length(other), "\n"
)
```

---

# 2. Gene group AUCs

```{r}
# RASopathy AUC for key predictors in clinvar_count_weighted
ras_aucs <- AUCs %>%
  filter(gene_group == "RASopathy", data_name == "clinvar_count_weighted") %>%
  filter(Predictor %in% c("n_COSMIC_codon_count", "n_sperm_codon_count",
                          "AlphaMissense", "REVEL",
                          "REVEL_LR_model_test_data")) %>%
  dplyr::select(Predictor, AUC_full, AUC_full_lower, AUC_full_upper) %>%
  arrange(desc(AUC_full))

cat("RASopathy AUCs (clinvar_count_weighted):\n")
print(ras_aucs)
```

```{r}
# AltFunc AUC for key predictors
altfunc_aucs <- AUCs %>%
  filter(gene_group == "altfunc", data_name == "clinvar_count_weighted") %>%
  filter(Predictor %in% c("n_COSMIC_codon_count", "n_sperm_codon_count",
                          "AlphaMissense", "REVEL",
                          "REVEL_LR_model_test_data")) %>%
  dplyr::select(Predictor, AUC_full, AUC_full_lower, AUC_full_upper) %>%
  arrange(desc(AUC_full))

cat("AltFunc AUCs (clinvar_count_weighted):\n")
print(altfunc_aucs)
```

```{r}
# AUC uplift from adding somatic data to REVEL (test data, RASopathy)
ras_uplift <- AUCs_LR_model %>%
  filter(gene_group == "RASopathy", test_or_train == "test",
         data_name == "clinvar_count_weighted", cep == "REVEL") %>%
  dplyr::select(AUC_full, AUC_full_lower, AUC_full_upper,
                AUC_cep_only, AUC_diff, delong_p_value)

cat("RASopathy REVEL+somatic uplift (test):\n")
print(ras_uplift)
```

---

# 3. Likelihood ratios

```{r}
# COSMIC codon count LRs at ACMG-relevant thresholds (RASopathy genes)
cosmic_lr_ras <- LR %>%
  filter(feature == "n_COSMIC_codon_count",
         gene_group == "RASopathies",
         data_set == "clinvar_unique_public") %>%
  dplyr::select(rule, threshold, sensitivity, specificity,
                LR_pos, LR_pos_low95, LR_pos_high95) %>%
  arrange(as.numeric(gsub("[^0-9.]", "", threshold)))

cat("COSMIC codon count LRs for RASopathy genes:\n")
print(cosmic_lr_ras)
```

```{r}
# Hotcodon LRs (COSMIC hotspot)
hotcodon_lr <- LR %>%
  filter(feature %in% c("hotcodon_cancer", "n_COSMIC"),
         gene_group %in% c("RASopathies", "AllGenes"),
         data_set == "clinvar_unique_public") %>%
  dplyr::select(gene_group, feature, rule, threshold, sensitivity, specificity,
                LR_pos, LR_pos_low95, LR_pos_high95)

cat("Hotcodon/COSMIC count LRs:\n")
print(hotcodon_lr)
```

---

# 4. Number of genes with somatic predictive power {.tabset}

## AUC lower CI > 0.6

```{r}
n_genes_06 <- sum(genes_clinvar_long_auc$auc_lower_06)
n_genes_total_analyzed <- nrow(genes_clinvar_long_auc)

cat(
  "Genes analyzed (clinvar_count_weighted):", n_genes_total_analyzed, "\n",
  "Genes with AUC lower CI > 0.6:", n_genes_06, "\n",
  "Percentage:", round(100 * n_genes_06 / n_genes_total_analyzed, 1), "%\n"
)
```

```{r}
# By gene group
gene_group_map <- c(
  setNames(rep("RASopathy", length(RASopathies)), RASopathies),
  setNames(rep("AltFunc (non-RAS)", length(altfunc_minus_RAS)), altfunc_minus_RAS),
  setNames(rep("LOF", length(lof)), lof),
  setNames(rep("Other", length(other)), other)
)

genes_clinvar_long_auc_grouped <- genes_clinvar_long_auc %>%
  mutate(gene_group = gene_group_map[gene]) %>%
  filter(!is.na(gene_group))

genes_clinvar_long_auc_grouped %>%
  group_by(gene_group) %>%
  summarise(
    n_genes      = n(),
    n_auc_06     = sum(auc_lower_06),
    pct_auc_06   = round(100 * sum(auc_lower_06) / n(), 1),
    n_auc_08     = sum(auc_lower_08),
    pct_auc_08   = round(100 * sum(auc_lower_08) / n(), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_auc_06))
```

## AUC lower CI > 0.8

```{r}
n_genes_08 <- sum(genes_clinvar_long_auc$auc_lower_08)

cat(
  "Genes with AUC lower CI > 0.8:", n_genes_08, "\n",
  "Percentage:", round(100 * n_genes_08 / n_genes_total_analyzed, 1), "%\n"
)
```

---

# 4b. Individual gene LR models: somatic vs REVEL comparison

```{r}
# Single-feature: genes where any somatic codon count predictor beats REVEL
# (somatic AUC lower CI > REVEL AUC upper CI, non-overlapping CIs)
# Restricted to the 195 dominant DD somatic selection genes

dominant_somatic_genes <- union(altfunc, union(lof, other))

revel_sf <- gene_AUCs_individual_features %>%
  filter(dataset == "clinvar_count_weighted", Predictor == "REVEL",
         gene %in% dominant_somatic_genes) %>%
  select(gene, AUC_revel_upper = AUC_full_upper)

somatic_sf <- gene_AUCs_individual_features %>%
  filter(dataset == "clinvar_count_weighted",
         Predictor %in% c("n_COSMIC_codon_count",
                          "n_sperm_codon_count",
                          "n_buccal_codon_count"),
         gene %in% dominant_somatic_genes) %>%
  select(gene, Predictor, somatic_lower = AUC_full_lower)

somatic_vs_revel <- somatic_sf %>%
  left_join(revel_sf, by = "gene") %>%
  mutate(beats_revel = !is.na(somatic_lower) & !is.na(AUC_revel_upper) &
           somatic_lower > AUC_revel_upper)

n_genes_with_revel_sf <- n_distinct(revel_sf$gene)

n_somatic_beats_revel <- somatic_vs_revel %>%
  group_by(gene) %>%
  summarise(any_beats = any(beats_revel), .groups = "drop") %>%
  filter(any_beats) %>%
  nrow()

cat(
  "Dominant DD somatic selection genes (altfunc + lof + other):", length(dominant_somatic_genes), "\n",
  "With REVEL AUC (clinvar count-weighted, ≥10 unique P/B variants):", n_genes_with_revel_sf, "\n",
  "Where any somatic codon count predictor beats REVEL",
  "(lower CI > upper CI of REVEL):", n_somatic_beats_revel, "\n"
)
```

```{r}
# Hold-out LR models (REVEL + somatic vs REVEL alone)
# DeLong p < 0.05 = significant difference; direction determined by AUC_full vs AUC_cep_only

lr_revel_test <- gene_AUCs_LR_model %>%
  filter(dataset == "clinvar_count_weighted",
         CEP == "REVEL",
         test_or_train == "test",
         gene %in% dominant_somatic_genes) %>%
  mutate(auc_diff = AUC_full - AUC_cep_only,
         sig      = delong_p_value < 0.05)

n_lr_genes_tested <- nrow(lr_revel_test)

n_sig_delong_uplift  <- sum(lr_revel_test$sig & lr_revel_test$auc_diff > 0)
n_sig_delong_hurt    <- sum(lr_revel_test$sig & lr_revel_test$auc_diff < 0)
n_nonsig             <- sum(!lr_revel_test$sig)

# Gene group membership of genes where somatic hurts
gene_group_map_lr <- lr_revel_test %>%
  mutate(gene_group = case_when(
    gene %in% RASopathies ~ "RASopathy",
    gene %in% altfunc     ~ "AltFunc",
    gene %in% lof         ~ "LOF",
    gene %in% other       ~ "Other",
    TRUE                  ~ "Unknown"
  ))

hurt_by_group <- gene_group_map_lr %>%
  filter(sig, auc_diff < 0) %>%
  count(gene_group, name = "n_hurt")

cat(
  "Dominant DD somatic selection genes with hold-out REVEL LR model test results",
  "(clinvar count-weighted):", n_lr_genes_tested, "\n\n",
  "Significant uplift   (DeLong p<0.05, combined AUC > REVEL):", n_sig_delong_uplift, "\n",
  "Significant worsened (DeLong p<0.05, combined AUC < REVEL):", n_sig_delong_hurt,   "\n",
  "Non-significant      (DeLong p>=0.05):                     ", n_nonsig,            "\n\n",
  "Gene group breakdown of genes where somatic significantly hurts:\n"
)
print(hurt_by_group)
```

---

# 5. VUS/conflicting variant resolution

```{r}
# How many VUS in RASopathy genes have any somatic data?
vus_somatic_data <- VUS_acmg_somatic %>%
  filter(gene %in% RASopathies) %>%
  mutate(
    has_cosmic  = n_COSMIC_codon_count > 0,
    has_sperm   = n_sperm_exomes_codon_count > 0 | n_sperm_twinsUK_codon_count > 0,
    has_any_somatic = has_cosmic | has_sperm
  )

cat(
  "VUS in RASopathy genes:", nrow(vus_somatic_data), "\n",
  "With COSMIC codon count data:", sum(vus_somatic_data$has_cosmic, na.rm = TRUE), "\n",
  "With sperm nanoseq data:", sum(vus_somatic_data$has_sperm, na.rm = TRUE), "\n",
  "With any somatic data:", sum(vus_somatic_data$has_any_somatic, na.rm = TRUE), "\n",
  "Percentage with any somatic data:",
    round(100 * mean(vus_somatic_data$has_any_somatic, na.rm = TRUE), 1), "%\n"
)
```

```{r}
# All VUS: somatic data coverage
vus_all_somatic <- VUS_acmg_somatic %>%
  mutate(
    has_cosmic  = n_COSMIC_codon_count > 0,
    has_sperm   = n_sperm_exomes_codon_count > 0 | n_sperm_twinsUK_codon_count > 0,
    has_any_somatic = has_cosmic | has_sperm
  )

cat(
  "All VUS total:", nrow(vus_all_somatic), "\n",
  "With COSMIC codon count:", sum(vus_all_somatic$has_cosmic, na.rm = TRUE), "\n",
  "Percentage:", round(100 * mean(vus_all_somatic$has_cosmic, na.rm = TRUE), 1), "%\n"
)
```

---

# 6. ACMG reclassification

```{r}
# VUS class distribution without and with somatic criteria
vus_classes_before <- VUS_acmg %>%
  count(acmg_class, name = "n_before") %>%
  mutate(pct_before = round(100 * n_before / sum(n_before), 1))

vus_classes_after <- VUS_acmg_somatic %>%
  count(acmg_class, name = "n_after") %>%
  mutate(pct_after = round(100 * n_after / sum(n_after), 1))

vus_comparison <- full_join(vus_classes_before, vus_classes_after, by = "acmg_class") %>%
  arrange(acmg_class)

cat("VUS classification before and after adding somatic criteria:\n")
print(vus_comparison)
```

```{r}
# Summary: how many VUS reclassified upward (to LP or P)?
n_vus_reclassified_up_before <- VUS_acmg %>%
  filter(acmg_class %in% c("Likely pathogenic", "Pathogenic")) %>%
  nrow()

n_vus_reclassified_up_after <- VUS_acmg_somatic %>%
  filter(acmg_class %in% c("Likely pathogenic", "Pathogenic")) %>%
  nrow()

cat(
  "VUS reclassified to LP/P (without somatic):", n_vus_reclassified_up_before, "\n",
  "VUS reclassified to LP/P (with somatic):", n_vus_reclassified_up_after, "\n",
  "Additional reclassifications with somatic criteria:",
    n_vus_reclassified_up_after - n_vus_reclassified_up_before, "\n"
)
```

```{r}
# Pathogenic variants: how many downgraded by somatic criteria (benign somatic code applied)?
n_path_downgraded <- path_acmg_somatic %>%
  filter(acmg_class %in% c("VUS", "Likely benign", "Benign")) %>%
  nrow()

cat("Pathogenic variants reclassified to VUS/LB/B with somatic criteria:", n_path_downgraded, "\n")
```

```{r}
# Somatic criteria applied: how many VUS received a somatic pathogenic code?
if ("SOM_PS" %in% names(VUS_acmg_somatic)) {
  n_som_ps <- sum(VUS_acmg_somatic$SOM_PS, na.rm = TRUE)
  n_som_pm <- sum(VUS_acmg_somatic$SOM_PM, na.rm = TRUE)
  n_som_pp <- sum(VUS_acmg_somatic$SOM_PP, na.rm = TRUE)
  n_som_bp <- sum(VUS_acmg_somatic$SOM_BP, na.rm = TRUE)

  cat(
    "VUS with SOM_PS (somatic strong pathogenic):", n_som_ps, "\n",
    "VUS with SOM_PM (somatic moderate pathogenic):", n_som_pm, "\n",
    "VUS with SOM_PP (somatic supporting pathogenic):", n_som_pp, "\n",
    "VUS with SOM_BP (somatic supporting benign):", n_som_bp, "\n"
  )
}
```

---

# 7. MAVE data numbers

```{r}
# PTPN11 MAVE AUC vs somatic data
ptpn11_mave_auc <- PTPN11_dat %>%
  filter(Form == "Long") %>%
  filter(Predictor %in% c("enrichment_ave", "n_COSMIC_codon_count", "n_sperm_codon_count",
                          "AlphaMissense_score", "REVEL_score")) %>%
  dplyr::rename(feature = Predictor) %>%
  mutate(feature = case_when(
    feature == "enrichment_ave"      ~ "MAVE",
    feature == "AlphaMissense_score" ~ "AlphaMissense",
    feature == "REVEL_score"         ~ "REVEL",
    TRUE ~ feature
  ))

cat("PTPN11 AUCs (count weighted):\n")
print(ptpn11_mave_auc %>% dplyr::select(feature, AUC, CI_lower, CI_upper) %>% arrange(desc(AUC)))
```

```{r}
# PTEN MAVE AUC
pten_mave_auc <- PTEN_MAVE %>%
  filter(feature %in% c("activity_MAVE", "abundance_MAVE", "n_COSMIC_codon_count",
                        "n_sperm_codon_count", "AlphaMissense", "REVEL")) %>%
  dplyr::select(feature, AUC, CI_lower, CI_upper) %>%
  arrange(desc(AUC))

cat("PTEN AUCs:\n")
print(pten_mave_auc)
```

```{r}
# DDX3X MAVE AUC
ddx3x_mave_auc <- DDX3X_MAVE %>%
  filter(feature %in% c("MAVE_score", "n_COSMIC_codon_count", "n_sperm_codon_count",
                        "AlphaMissense", "REVEL")) %>%
  dplyr::select(feature, AUC, CI_lower, CI_upper) %>%
  arrange(desc(AUC))

cat("DDX3X AUCs:\n")
print(ddx3x_mave_auc)
```

```{r}
# HRAS MAVE AUC
hras_mave_auc <- HRAS_MAVE %>%
  filter(feature %in% c("MAVE_score", "n_COSMIC_codon_count", "n_sperm_codon_count",
                        "AlphaMissense", "REVEL")) %>%
  dplyr::select(feature, AUC, CI_lower, CI_upper) %>%
  arrange(desc(AUC))

cat("HRAS AUCs:\n")
print(hras_mave_auc)
```

```{r}
# PTPN11 p.Asn308Asp (N308D): MAVE enrichment value
# Codon 308, wt = N (Asn), mut = D (Asp)
ptpn11_n308d <- PTPN11_MAVE_vis %>%
  filter(pos == 308, wt == "N", mut == "D")

cat("PTPN11 p.Asn308Asp MAVE enrichment:\n")
print(ptpn11_n308d)
```

```{r}
# How many PTPN11 codon positions have MAVE data?
n_ptpn11_codons_with_mave <- PTPN11_MAVE_vis %>%
  filter(!is.na(enrichment_ave)) %>%
  distinct(pos) %>%
  nrow()

n_ptpn11_variants_with_mave <- PTPN11_MAVE_vis %>%
  filter(!is.na(enrichment_ave)) %>%
  nrow()

cat(
  "PTPN11 codon positions with MAVE data:", n_ptpn11_codons_with_mave, "\n",
  "PTPN11 amino acid substitutions with MAVE data:", n_ptpn11_variants_with_mave, "\n"
)
```
