---
title: "Pathway GO term analysis"
output:
  html_document:
    df_print: paged
    code_folding: hide
  pdf_document: default
date: "2026-02-24"
---

# Aim

Are the genes that have the highest utility of somatic data in germline clinical variant interpretation enriched for particular KEGG pathways, GO terms, phenotypes, or panels?

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(patchwork)
theme_set(theme_bw(14))

safe_colorblind_palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499",
                             "#44AA99", "#999933", "#882255", "#661100", "#6699CC", "#888888")

dir.create("../../Data/Tables",  recursive = TRUE, showWarnings = FALSE)
dir.create("../../Data/Figures", recursive = TRUE, showWarnings = FALSE)
```

# Gene and pathway lists

Gene list outputs were pre-generated by `Define_somatic_selection_gene_groups.Rmd` and `Define_cancer_DD_gene_mechanism_sets.Rmd`.

```{r, message=FALSE, warning=FALSE}
# Somatic selection gene vectors (hardcoded; match Define_somatic_selection_gene_groups.Rmd)
sperm_selection_genes <- c(
  "PTPN11", "MIB1", "RIT1", "FGFR3", "EP300", "FGFR2", "KDM5B", "NF1", "SMAD6",
  "CUL3", "RASA2", "PRRC2A", "PTEN", "ROBO1", "DDX3X", "CSNK2B", "KRAS", "PPM1D",
  "ARID1A", "BRAF", "HRAS", "KMT2E", "SCAF4", "BMPR2", "TCF12", "CCAR2", "DHX9", "NSD1",
  "LZTR1", "ARHGAP35", "CBL", "SSX1", "RBM12", "SMAD4", "FAM222B",
  "KDM5C", "KMT2D", "AR", "CTNNB1", "RAF1"
)

buccal_selection_genes <- c(
  "NOTCH1", "FAT1", "TP53", "NOTCH2", "ATM", "MGA", "KMT2C", "ASXL1", "CHEKC2",
  "TET2", "PPM1D", "BCORL1", "ARID1A", "SETD2", "ZFP36L2", "ARID2", "EP300",
  "NOTCH3", "KDM5C", "RAC1", "ARID1B", "RBM10", "KDM6A", "STAG2", "EPHA2", "CIC",
  "SF3B1", "ZFP36L1", "EGFR", "DNMT3A", "PIK3R1", "TP63", "ARID5B", "AJUBA",
  "FGFR3", "ZNF750", "FUBP1", "CUL3", "KLF5", "PAX9", "FOXP1", "ETV6", "RHOA",
  "RARG", "SPOP", "SMARCB1", "CCND1", "HLA-B", "EIF1AX", "PIK3CA"
)

somatic_selection_genes <- union(sperm_selection_genes, buccal_selection_genes)

# Load pre-generated gene lists from Data/Gene_definitions/
somatic_selection_and_cancer_genes <- read_csv(
  "../../Data/Gene_definitions/Somatic_selection_cancer_genes.csv",
  show_col_types = FALSE
) %>% pull(gene)

altfunc <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_AltFunc.csv",
                    show_col_types = FALSE) %>% pull(gene_symbol)
lof     <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_LOF.csv",
                    show_col_types = FALSE) %>% pull(gene_symbol)
other   <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_other.csv",
                    show_col_types = FALSE) %>% pull(gene_symbol)
VCEP_rasopathy <- read_csv("../../Data/Gene_definitions/RASopathy_genes_VCEP_030425.csv",
                            show_col_types = FALSE) %>% pull(gene)

panelapp_macrocephaly_somatic <- read_csv(
  "../../Data/Gene_definitions/Macrocephaly_monoallelic_DD_somatic_genes.csv",
  show_col_types = FALSE
) %>% pull(gene)
```

```{r, message=FALSE, warning=FALSE}
# DDG2P (for HPO-based macrocephaly definition)
DDG2P <- read_csv("../../Data/DDG2P/G2P_DD_2025-04-16.csv", show_col_types = FALSE) %>%
  clean_names()
DDG2P_monoallelic <- DDG2P %>%
  filter(allelic_requirement %in% c("monoallelic_autosomal", "monoallelic_X_heterozygous"))

macrocephaly_HPO    <- "HP:0000256"
DDG2P_macrocephaly_mono <- DDG2P_monoallelic %>%
  filter(str_detect(phenotypes, macrocephaly_HPO)) %>%
  pull(gene_symbol) %>% unique()

DDG2P_macrocephaly_mono_somatic <- intersect(DDG2P_macrocephaly_mono,
                                              somatic_selection_and_cancer_genes)
```

```{r, message=FALSE, warning=FALSE}
# DATA_SOURCE: PanelApp Australia Macrocephaly-Megalencephaly panel TSV
# — not bundled with project; obtain from https://panelapp.agha.umccr.org/
# panelapp_macrocephaly <- read_tsv("...") %>% clean_names() %>%
#   filter(gene_symbol %in% DDG2P_monoallelic$gene_symbol) %>%
#   pull(gene_symbol) %>% unique()
# (using the pre-generated macrocephaly somatic list instead)
panelapp_macrocephaly <- panelapp_macrocephaly_somatic  # fallback

# DATA_SOURCE: UK PanelApp segmental overgrowth panels TSV
# — not bundled with project; obtain from https://panelapp.genomicsengland.co.uk/
# seg_panel_1 <- read_tsv("...") %>% clean_names() %>% pull(gene_symbol) %>% unique()
# seg_panel_2 <- read_tsv("...") %>% clean_names() %>% pull(gene_symbol) %>% unique()
seg_panel <- character(0)  # fallback: populate from PanelApp downloads

# DATA_SOURCE: UK PanelApp RASopathy green genes TSV
# — not bundled with project; obtain from https://panelapp.genomicsengland.co.uk/
# panelapp_RAS <- read_tsv("...") %>% clean_names() %>% pull(gene_symbol) %>% unique()
panelapp_RAS <- VCEP_rasopathy  # fallback: use VCEP list
```

```{r}
# VCEP-defined PIK3CA-MTOR brain malformations pathway genes (hardcoded)
VCEP_PIK3CA_MTOR <- c("AKT3", "PIK3CA", "PIK3R2", "MTOR", "DEPDC5")
```

# Load AUC data

```{r, message=FALSE, warning=FALSE}
date <- "240426"

gene_AUCs_individual_features <- read_csv(
  paste0("../../Data/AUC_tables/Gene_group_hold_out_models/Single_feature_AUCs_individual_genes_", date, ".csv"),
  show_col_types = FALSE
)

all_genes <- gene_AUCs_individual_features$gene %>% unique()

n_vars <- read_csv(
  paste0("../../Data/AUC_tables/Gene_group_hold_out_models/Variant_counts_per_gene_", date, ".csv"),
  show_col_types = FALSE
)

clinvar_long_COSMIC <- gene_AUCs_individual_features %>%
  filter(Predictor == "n_COSMIC_codon_count") %>%
  filter(dataset == "clinvar_long") %>%
  arrange(desc(AUC_full_lower)) %>%
  left_join(n_vars %>% filter(data_set == "clinvar_mutation_data_long"),
            by = "gene")

clinvar_unique_COSMIC <- gene_AUCs_individual_features %>%
  filter(Predictor %in% "n_COSMIC_codon_count") %>%
  filter(dataset == "clinvar_unique") %>%
  arrange(desc(AUC_full_lower)) %>%
  left_join(n_vars %>% filter(data_set == "clinvar_mutation_data_unique"),
            by = "gene")
```

# Panelapp defined genes

1. Segmental overgrowth genes from UK panelApp (see DATA_SOURCE note in setup)

2. RASopathy genes from UK PanelApp (fallback: VCEP list loaded above)

# VCEP defined gene lists

1. Brain malformations (PIK3CA-MTOR pathway genes): `VCEP_PIK3CA_MTOR` (hardcoded above)

2. RASopathy genes: `VCEP_rasopathy` (loaded above from `Data/Gene_definitions/`)

# GO terms defined gene lists

Gene symbols were annotated with high-level molecular function classes using Gene Ontology (GO). Human gene annotations were obtained from the Bioconductor package org.Hs.eg.db (Homo sapiens). For each gene, all associated GO terms were retrieved using Entrez Gene identifiers.

```{r, warning=FALSE, message=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(AnnotationDbi)
  library(org.Hs.eg.db)
  library(GO.db)
})

GO_PARENTS <- tibble::tribble(
  ~class,                        ~goid,        ~ontology,

  "Kinase",                      "GO:0004672", "MF",
  "Protein phosphatase",         "GO:0004725", "MF",
  "Protein phosphatase",         "GO:0004722", "MF",
  "Protein phosphatase",         "GO:0034594", "MF",
  "RAS GEF",                     "GO:0005085", "MF",
  "RAS GAP",                     "GO:0017016", "MF",
  "Pore-forming ion channel",    "GO:0005244", "MF",
  "Pore-forming ion channel",    "GO:0015276", "MF",
  "Pore-forming ion channel",    "GO:0015278", "MF",
  "Pore-forming ion channel",    "GO:0008381", "MF",
  "Pore-forming ion channel",    "GO:0005261", "MF",
  "Chromatin modifier",          "GO:0004402", "MF",
  "Chromatin modifier",          "GO:0004407", "MF",
  "Chromatin modifier",          "GO:0018024", "MF",
  "Chromatin modifier",          "GO:0071558", "MF",
  "Chromatin modifier",          "GO:0006338", "BP",
  "Transcription factor",        "GO:0003700", "MF",
  "GPCR",                        "GO:0004930", "MF",
  "RTK",                         "GO:0004714", "MF",
  "Ubiquitin ligase",            "GO:0061630", "MF",
  "GTPase",                      "GO:0003924", "MF"
)

get_descendants <- function(goid, ontology) {
  if (ontology == "MF") return(unique(c(goid, unlist(GOMFOFFSPRING[[goid]]))))
  if (ontology == "BP") return(unique(c(goid, unlist(GOBPOFFSPRING[[goid]]))))
  if (ontology == "CC") return(unique(c(goid, unlist(GOCCOFFSPRING[[goid]]))))
  stop("Unknown ontology: ", ontology)
}

annotate_by_go_parents <- function(symbols, parents = GO_PARENTS) {
  symbols <- trimws(symbols)
  symbols <- unique(na.omit(symbols))
  symbols <- symbols[symbols != ""]

  parent_sets <- parents %>%
    mutate(desc = purrr::map2(goid, ontology, get_descendants)) %>%
    group_by(class) %>%
    summarise(desc = list(unique(unlist(desc))), .groups = "drop")

  class_levels <- parent_sets$class

  map <- AnnotationDbi::select(
    org.Hs.eg.db, keys = symbols, keytype = "SYMBOL",
    columns = c("SYMBOL", "ENTREZID")
  ) %>% distinct()

  go <- AnnotationDbi::select(
    org.Hs.eg.db, keys = na.omit(map$ENTREZID), keytype = "ENTREZID",
    columns = c("GO", "ONTOLOGY")
  ) %>%
    dplyr::filter(!is.na(GO), !is.na(ONTOLOGY)) %>%
    dplyr::distinct(ENTREZID, GO, ONTOLOGY)

  gene_flags_long <- go %>%
    dplyr::inner_join(map, by = "ENTREZID") %>%
    dplyr::group_by(SYMBOL) %>%
    dplyr::summarise(gene_go = list(unique(GO)), .groups = "drop") %>%
    dplyr::mutate(
      flags = purrr::map(gene_go, function(terms) {
        tibble::tibble(
          class   = parent_sets$class,
          present = purrr::map_lgl(parent_sets$desc, function(desc) any(terms %in% desc))
        )
      })
    ) %>%
    dplyr::select(SYMBOL, flags) %>%
    tidyr::unnest(flags)

  gene_flags_wide <- gene_flags_long %>%
    tidyr::pivot_wider(names_from = class, values_from = present, values_fill = FALSE)

  for (cls in class_levels) {
    if (!cls %in% colnames(gene_flags_wide)) gene_flags_wide[[cls]] <- FALSE
  }

  gene_flags_wide <- gene_flags_wide %>%
    dplyr::select(SYMBOL, dplyr::all_of(class_levels))

  out <- tibble::tibble(symbol = symbols) %>%
    dplyr::left_join(map, by = c("symbol" = "SYMBOL")) %>%
    dplyr::left_join(gene_flags_wide, by = c("symbol" = "SYMBOL")) %>%
    dplyr::mutate(mapped = !is.na(ENTREZID))

  for (cls in class_levels) {
    out[[cls]] <- tidyr::replace_na(out[[cls]], FALSE)
  }

  out %>%
    dplyr::arrange(symbol) %>%
    dplyr::distinct(symbol, .keep_all = TRUE)
}
```

```{r, warning=FALSE, message=FALSE}
class_groups <- annotate_by_go_parents(all_genes)
```

# Signalling pathways defined gene lists

```{r}
pi3k_mtor_genes <- c(
  "RICTOR","MTOR","RHEB","BAD","PIK3R2","PIK3CA","PIK3R3",
  "GRB10","KRAS","RB1CC1","RPTOR","HRAS","FOXO3","ULK1","EIF4EBP1",
  "TSC2","CDKN1B","NRAS","PDK1","PIK3CB","NOS3","PTEN","PIK3CG","FOXO1","GSK3B"
)

ras_mapk_genes <- c(
  "KRAS","HRAS","NRAS","MRAS",
  "RAF1","ARAF","BRAF",
  "MAP2K1","MAP2K2","MAPK1","MAPK3",
  "SOS1","SOS2","GRB2","RASA1","RASA2",
  "EGFR","FGFR1","FGFR2","FGFR3","PDGFRA","PDGFRB","ERBB2","ERBB3","NTRK1","NTRK2","IGF1R"
)
```

# Generate annotated table with flags for involvement in pathways and GO classes

```{r}
clinvar_unique_COSMIC_annot <- clinvar_unique_COSMIC %>%
  mutate(gene = trimws(gene)) %>%
  left_join(class_groups, by = c("gene" = "symbol")) %>%
  dplyr::select(gene, AUC_full, AUC_full_upper, AUC_full_lower, `Chromatin modifier`, Kinase, RTK, GPCR,
                `Pore-forming ion channel`, `Protein phosphatase`, `Transcription factor`,
                `Ubiquitin ligase`, GTPase, `RAS GEF`, `RAS GAP`) %>% clean_names() %>%
  mutate(
    `PI3K-MTOR pathway`         = gene %in% pi3k_mtor_genes,
    `RAS-MAPK pathway`          = gene %in% ras_mapk_genes,
    Macrocephaly_DDG2P          = gene %in% DDG2P_macrocephaly_mono,
    Macrocephaly_DDG2P_somatic  = gene %in% DDG2P_macrocephaly_mono_somatic,
    `RAS-MAPK_PanelApp`         = gene %in% panelapp_RAS,
    Macrocephaly_PanelApp       = gene %in% panelapp_macrocephaly,
    Macrocephaly_PanelApp_somatic = gene %in% panelapp_macrocephaly_somatic,
    Segmental_overgrowth_PanelApp = gene %in% seg_panel,
    Brain_malformations_VCEP    = gene %in% VCEP_PIK3CA_MTOR,
    RASopathy_VCEP              = gene %in% VCEP_rasopathy
  )

clinvar_long_COSMIC_annot <- clinvar_long_COSMIC %>%
  mutate(gene = trimws(gene)) %>%
  left_join(class_groups, by = c("gene" = "symbol")) %>%
  dplyr::select(gene, AUC_full, AUC_full_upper, AUC_full_lower, `Chromatin modifier`, Kinase, RTK, GPCR,
                `Pore-forming ion channel`, `Protein phosphatase`, `Transcription factor`,
                `Ubiquitin ligase`, GTPase, `RAS GEF`, `RAS GAP`) %>% clean_names() %>%
  mutate(
    `PI3K-MTOR pathway`         = gene %in% pi3k_mtor_genes,
    `RAS-MAPK pathway`          = gene %in% ras_mapk_genes,
    Macrocephaly_DDG2P          = gene %in% DDG2P_macrocephaly_mono,
    Macrocephaly_DDG2P_somatic  = gene %in% DDG2P_macrocephaly_mono_somatic,
    `RAS-MAPK_PanelApp`         = gene %in% panelapp_RAS,
    Macrocephaly_PanelApp       = gene %in% panelapp_macrocephaly,
    Macrocephaly_PanelApp_somatic = gene %in% panelapp_macrocephaly_somatic,
    Segmental_overgrowth_PanelApp = gene %in% seg_panel,
    Brain_malformations_VCEP    = gene %in% VCEP_PIK3CA_MTOR,
    RASopathy_VCEP              = gene %in% VCEP_rasopathy
  )

all_AUCs_annot <- gene_AUCs_individual_features %>%
  mutate(gene = trimws(gene)) %>%
  left_join(class_groups, by = c("gene" = "symbol")) %>%
  dplyr::select(gene, Predictor, dataset, AUC_full, AUC_full_upper, AUC_full_lower,
                `Chromatin modifier`, Kinase, RTK, GPCR,
                `Pore-forming ion channel`, `Protein phosphatase`, `Transcription factor`,
                `Ubiquitin ligase`, GTPase, `RAS GEF`, `RAS GAP`) %>% clean_names() %>%
  mutate(
    `PI3K-MTOR pathway`         = gene %in% pi3k_mtor_genes,
    `RAS-MAPK pathway`          = gene %in% ras_mapk_genes,
    Macrocephaly_DDG2P          = gene %in% DDG2P_macrocephaly_mono,
    Macrocephaly_DDG2P_somatic  = gene %in% DDG2P_macrocephaly_mono_somatic,
    `RAS-MAPK_PanelApp`         = gene %in% panelapp_RAS,
    Macrocephaly_PanelApp       = gene %in% panelapp_macrocephaly,
    Macrocephaly_PanelApp_somatic = gene %in% panelapp_macrocephaly_somatic,
    Segmental_overgrowth_PanelApp = gene %in% seg_panel,
    Brain_malformations_VCEP    = gene %in% VCEP_PIK3CA_MTOR,
    RASopathy_VCEP              = gene %in% VCEP_rasopathy
  )
```

# Ranking utility of different gene groups

Which of all the gene groups defined above (phenotypes, panelapp, VCEP, GO terms, signalling pathways) are informative discriminators for genes that have high AUCs for a somatic predictor?

I've subset to genes that have at least 10 pathogenic UNIQUE observations and at least 10 benign.

Three methods:

1. Compare mean AUCs in vs out of group (wilcoxon)
2. Compare enrichment of top quartile AUC genes in vs out of group (fisher's exact test)
3. What percentage of genes in each group have a 'significant' AUC from cancer data (defined as lower CI of AUC > 0.6)

## Mean AUC difference and top quartile enrichment {.tabset}

```{r}
evaluate_gene_groups <- function(df,
                                 auc_col = "auc_full",
                                 gene_col = "gene",
                                 top_quantile = 0.75) {
  df <- df %>%
    dplyr::filter(!is.na(.data[[auc_col]])) %>%
    dplyr::mutate(
      .gene = trimws(.data[[gene_col]]),
      .auc  = .data[[auc_col]]
    )

  threshold <- stats::quantile(df$.auc, probs = top_quantile, na.rm = TRUE)

  df <- df %>%
    dplyr::mutate(top_quantile_flag = .auc >= threshold)

  class_cols <- df %>%
    dplyr::select(where(is.logical)) %>%
    dplyr::select(-top_quantile_flag) %>%
    colnames()

  results <- purrr::map_dfr(class_cols, function(cls) {
    in_group  <- df[[cls]] == TRUE
    n_in_group  <- sum(in_group,  na.rm = TRUE)
    n_out_group <- sum(!in_group, na.rm = TRUE)

    if (n_in_group == 0 || n_out_group == 0) return(NULL)

    auc_in  <- df$.auc[in_group]
    auc_out <- df$.auc[!in_group]

    mean_in  <- mean(auc_in,  na.rm = TRUE)
    mean_out <- mean(auc_out, na.rm = TRUE)
    delta_mean <- mean_in - mean_out

    wilc <- stats::wilcox.test(auc_in, auc_out)

    n_top_in    <- sum(df$top_quantile_flag[in_group], na.rm = TRUE)
    n_top_total <- sum(df$top_quantile_flag, na.rm = TRUE)

    contingency <- matrix(c(
      n_top_in,
      n_in_group - n_top_in,
      n_top_total - n_top_in,
      nrow(df) - n_top_total - (n_in_group - n_top_in)
    ), nrow = 2, byrow = TRUE)

    fisher_res <- stats::fisher.test(contingency)

    tibble::tibble(
      class                   = cls,
      n_in_group              = n_in_group,
      n_out_group             = n_out_group,
      mean_auc_in_group       = mean_in,
      mean_auc_not_in_group   = mean_out,
      delta_mean_auc          = delta_mean,
      wilcox_stat             = unname(wilc$statistic),
      wilcox_p                = wilc$p.value,
      n_top_quantile_in_group = n_top_in,
      pct_top_quantile_in_group = 100 * n_top_in / n_in_group,
      odds_ratio_top_quantile = unname(fisher_res$estimate),
      fisher_p_top_quantile   = fisher_res$p.value
    )
  })

  results %>%
    dplyr::mutate(
      wilcox_fdr = p.adjust(wilcox_p, method = "BH"),
      fisher_fdr = p.adjust(fisher_p_top_quantile, method = "BH")
    ) %>%
    dplyr::arrange(wilcox_fdr, fisher_fdr, desc(delta_mean_auc))
}
```

```{r}
enrichment_results_clinvar_unique_COSMIC_codon_count <- evaluate_gene_groups(
  clinvar_unique_COSMIC_annot, auc_col = "auc_full", gene_col = "gene", top_quantile = 0.75
)

enrichment_results_clinvar_long_COSMIC_codon_count <- evaluate_gene_groups(
  clinvar_long_COSMIC_annot %>% filter(gene %in% clinvar_unique_COSMIC_annot$gene),
  auc_col = "auc_full", gene_col = "gene", top_quantile = 0.75
)
```

```{r}
label_categories <- function(df) {
  df %>%
    dplyr::rename(any_of(c(category = "class"))) %>%
    mutate(category = case_when(
      category == "PI3K-MTOR pathway"            ~ "Pathway: PI3K-MTOR (WikiPathways)",
      category == "RAS-MAPK pathway"             ~ "Pathway: RAS-MAPK (KEGG)",
      category == "Macrocephaly_DDG2P"           ~ "Phenotype: Macrocephaly (DDG2P HPO)",
      category == "Macrocephaly_DDG2P_somatic"   ~ "Phenotype: Macrocephaly (DDG2P HPO, somatic genes)",
      category == "RAS-MAPK_PanelApp"            ~ "Panel: RAS-MAPK (UK PanelApp)",
      category == "Macrocephaly_PanelApp"        ~ "Panel: Macrocephaly (Aus PanelApp, monoallelic DD genes)",
      category == "Macrocephaly_PanelApp_somatic"~ "Panel: Macrocephaly somatic (Aus PanelApp, monoallelic DD genes, somatic genes)",
      category == "Segmental_overgrowth_PanelApp"~ "Panel: Segmental overgrowth (UK PanelApp)",
      category == "Brain_malformations_VCEP"     ~ "VCEP: Brain malformations (PIK3CA/MTOR)",
      category == "RASopathy_VCEP"               ~ "VCEP: RASopathies",
      category == "protein_phosphatase"          ~ "GO: Protein phosphatase activity",
      category == "pore_forming_ion_channel"     ~ "GO: Pore-forming ion channel activity",
      category == "ras_gef"                      ~ "GO: RAS GEF",
      category == "rtk"                          ~ "GO: Receptor tyrosine kinase activity",
      category == "Kinase"                       ~ "GO: Protein kinase activity",
      category == "chromatin_modifier"           ~ "GO: Chromatin modifier activity",
      category == "transcription_factor"         ~ "GO: DNA-binding transcription factor activity",
      category == "gpcr"                         ~ "GO: G protein-coupled receptor activity",
      category == "gt_pase"                      ~ "GO: GTPase",
      category == "kinase"                       ~ "GO: kinase activity",
      category == "ubiquitin_ligase"             ~ "GO: Ubiquitin ligase activity",
      TRUE ~ category
    ))
}
```

### Clinvar unique

```{r}
enrichment_results_clinvar_unique_COSMIC_codon_count %>%
  arrange(wilcox_p) %>%
  dplyr::select(-delta_mean_auc, -wilcox_stat, -n_out_group) %>%
  label_categories()
```

### Clinvar long
(removing genes with <10 unique observations)

```{r}
enrichment_results_clinvar_long_COSMIC_codon_count %>%
  arrange(wilcox_p) %>%
  dplyr::select(-delta_mean_auc, -wilcox_stat, -n_out_group) %>%
  label_categories()
```

## Percentage of genes with somatic data value {.tabset}

What % of genes in each list above that have a meaningful AUC uplift from cancer data (defined as lower CI of AUC > 0.6):

### Clinvar unique

```{r}
clinvar_unique_COSMIC_annot %>%
  dplyr::select(where(is.logical), auc_full_lower, auc_full_upper) %>%
  pivot_longer(cols = where(is.logical), names_to = "category", values_to = "flag") %>%
  filter(flag) %>%
  group_by(category) %>%
  summarise(
    `n genes` = n(),
    `n AUC lower > 0.6` = sprintf("%d (%.1f%%)",
      sum(auc_full_lower > 0.6, na.rm = TRUE),
      100 * mean(auc_full_lower > 0.6, na.rm = TRUE)),
    `n AUC lower > 0.8` = sprintf("%d (%.1f%%)",
      sum(auc_full_lower > 0.8, na.rm = TRUE),
      100 * mean(auc_full_lower > 0.8, na.rm = TRUE)),
    `n AUC upper < 0.4` = sprintf("%d (%.1f%%)",
      sum(auc_full_upper < 0.4, na.rm = TRUE),
      100 * mean(auc_full_upper < 0.4, na.rm = TRUE)),
    `n underpowered/inconclusive` = sprintf("%d (%.1f%%)",
      sum(!(auc_full_lower > 0.6 | auc_full_upper < 0.4), na.rm = TRUE),
      100 * mean(!(auc_full_lower > 0.6 | auc_full_upper < 0.4), na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  label_categories()
```

### Clinvar long
(removing genes with <10 unique observations)

```{r}
powered_genes <- clinvar_unique_COSMIC_annot %>% pull(gene) %>% unique()

clinvar_long_COSMIC_annot %>%
  dplyr::select(where(is.logical), auc_full_lower, auc_full_upper) %>%
  pivot_longer(cols = where(is.logical), names_to = "category", values_to = "flag") %>%
  filter(flag) %>%
  group_by(category) %>%
  summarise(
    `n genes` = n(),
    `n AUC lower > 0.6` = sprintf("%d (%.1f%%)",
      sum(auc_full_lower > 0.6, na.rm = TRUE),
      100 * mean(auc_full_lower > 0.6, na.rm = TRUE)),
    `n AUC lower > 0.8` = sprintf("%d (%.1f%%)",
      sum(auc_full_lower > 0.8, na.rm = TRUE),
      100 * mean(auc_full_lower > 0.8, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  label_categories()
```

Version without GO terms (Table 3):

```{r}
table_3 <- clinvar_long_COSMIC_annot %>%
  dplyr::select(where(is.logical), auc_full_lower, auc_full_upper) %>%
  pivot_longer(cols = where(is.logical), names_to = "category", values_to = "flag") %>%
  filter(flag) %>%
  group_by(category) %>%
  summarise(
    `n genes` = n(),
    `n AUC lower > 0.6` = sprintf("%d (%.1f%%)",
      sum(auc_full_lower > 0.6, na.rm = TRUE),
      100 * mean(auc_full_lower > 0.6, na.rm = TRUE)),
    `n AUC lower > 0.8` = sprintf("%d (%.1f%%)",
      sum(auc_full_lower > 0.8, na.rm = TRUE),
      100 * mean(auc_full_lower > 0.8, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(category = case_when(
    category == "PI3K-MTOR pathway"            ~ "Pathway: PI3K-MTOR (WikiPathways)",
    category == "RAS-MAPK pathway"             ~ "Pathway: RAS-MAPK (KEGG)",
    category == "Macrocephaly_PanelApp"        ~ "Phenotype: Dominant Macrocephaly",
    category == "RAS-MAPK_PanelApp"            ~ "Panel: RAS-MAPK (UK PanelApp)",
    category == "Segmental_overgrowth_PanelApp"~ "Panel: Segmental overgrowth (UK PanelApp)",
    category == "Brain_malformations_VCEP"     ~ "VCEP: Brain malformations (PIK3CA/MTOR)",
    category == "RASopathy_VCEP"               ~ "VCEP: RASopathies",
    category %in% c(
      "Macrocephaly_DDG2P", "Macrocephaly_DDG2P_somatic", "Macrocephaly_PanelApp_somatic"
    ) ~ NA_character_,
    TRUE ~ category
  )) %>%
  filter(!is.na(category)) %>%
  filter(grepl("^(VCEP|Panel|Phenotype|Pathway):", category)) %>%
  mutate(category = factor(category, levels = c(
    "VCEP: Brain malformations (PIK3CA/MTOR)",
    "VCEP: RASopathies",
    "Panel: RAS-MAPK (UK PanelApp)",
    "Panel: Segmental overgrowth (UK PanelApp)",
    "Phenotype: Dominant Macrocephaly",
    "Pathway: PI3K-MTOR (WikiPathways)",
    "Pathway: RAS-MAPK (KEGG)"
  ))) %>%
  arrange(category)

table_3
```

```{r}
write_csv(table_3, "../../Data/Tables/Table_3_gene_groups_and_somatic_AUCs.csv")
```

### Generate version for export to table for paper

```{r}
clinvar_long_COSMIC_annot %>%
  mutate(
    `Gene group: AltFunc` = gene %in% altfunc,
    `Gene group: LOF`     = gene %in% lof,
    `Gene group: Other`   = gene %in% other
  ) %>%
  filter(gene %in% powered_genes) %>%
  mutate(somatic_helpful = case_when(
    auc_full_lower > 0.8 ~ "Somatic data helpful, AUC lower CI > 0.8",
    auc_full_lower > 0.7 ~ "Somatic data helpful, AUC lower CI > 0.7",
    auc_full_lower > 0.6 ~ "Somatic data helpful, AUC lower CI > 0.6",
    auc_full_upper < 0.4 ~ "Somatic data not helpful, AUC upper CI < 0.4",
    TRUE ~ "Underpowered/inconclusive"
  )) %>%
  mutate(cancer_data_helpful = ifelse(auc_full_lower > 0.6, "Yes", "No")) %>%
  dplyr::select(where(is.logical), cancer_data_helpful) %>%
  pivot_longer(cols = where(is.logical), names_to = "category", values_to = "flag") %>%
  filter(flag) %>%
  group_by(category) %>%
  summarise(
    n_true     = n(),
    n_helpful  = sum(cancer_data_helpful == "Yes", na.rm = TRUE),
    perc_helpful = 100 * n_helpful / n_true,
    .groups = "drop"
  ) %>%
  arrange(desc(perc_helpful)) %>%
  label_categories()
```

# Visualising top genes and their ontology

## AUC confidence heatmap

```{r, warning=FALSE, message=FALSE}
gene_AUCs_LR_model <- read_csv(
  paste0("../../Data/AUC_tables/Gene_group_hold_out_models/AUC_table_gene_holdout_LR_models_", date, ".csv"),
  show_col_types = FALSE
)

n_vars_per_gene <- read_csv(
  paste0("../../Data/AUC_tables/Gene_group_hold_out_models/Variant_counts_per_gene_", date, ".csv"),
  show_col_types = FALSE
)

n_vars_per_gene_clinvar <- n_vars_per_gene %>%
  filter(data_set %in% c("clinvar_mutation_data_unique", "clinvar_mutation_data_long")) %>%
  pivot_wider(names_from = data_set, values_from = c(n_path_variants, n_benign_variants)) %>%
  dplyr::select(-n_benign_variants_clinvar_mutation_data_long) %>%
  dplyr::rename(
    n_benign            = n_benign_variants_clinvar_mutation_data_unique,
    n_unique_pathogenic = n_path_variants_clinvar_mutation_data_unique,
    n_pathogenic        = n_path_variants_clinvar_mutation_data_long
  )
```

```{r}
var_table <- n_vars_per_gene_clinvar %>%
  left_join(
    all_AUCs_annot %>%
      dplyr::select(gene, `PI3K-MTOR pathway`, `RAS-MAPK pathway`, `RAS-MAPK_PanelApp`,
                    RASopathy_VCEP, `Macrocephaly_PanelApp_somatic`,
                    kinase, rtk, protein_phosphatase, chromatin_modifier, transcription_factor) %>%
      unique(),
    by = "gene"
  ) %>%
  mutate(
    `AltFunc gene group` = gene %in% altfunc,
    LOF_gene             = gene %in% lof
  )

make_gene_panel <- function(genes, dataset_name, gene_group_cols = NULL) {

  inclusion_genes <- var_table %>%
    dplyr::filter(n_unique_pathogenic >= 10, n_benign >= 10) %>%
    dplyr::pull(gene)

  target_genes <- intersect(genes, inclusion_genes)

  gene_df_somatic <- gene_AUCs_individual_features %>%
    dplyr::filter(dataset == dataset_name) %>%
    dplyr::filter(Predictor %in% c(
      "n_COSMIC_codon_count", "n_sperm_codon_count", "n_buccal_codon_count"
    )) %>%
    dplyr::mutate(
      Predictor_simple = dplyr::case_when(
        Predictor == "n_COSMIC_codon_count" ~ "Cancer",
        Predictor == "n_sperm_codon_count"  ~ "Sperm",
        Predictor == "n_buccal_codon_count" ~ "Buccal",
        TRUE ~ Predictor
      ),
      Predictor_simple = factor(Predictor_simple, levels = c("Cancer", "Sperm", "Buccal"))
    ) %>%
    dplyr::select(Predictor_simple, gene, AUC_full, AUC_full_lower, AUC_full_upper) %>%
    dplyr::mutate(
      somatic_ci_cat = dplyr::case_when(
        is.na(AUC_full) | is.na(AUC_full_lower) | is.na(AUC_full_upper) ~ "missing",
        AUC_full_lower > 0.8 ~ "Somatic data helpful, AUC lower CI > 0.8",
        AUC_full_lower > 0.7 ~ "Somatic data helpful, AUC lower CI > 0.7",
        AUC_full_lower > 0.6 ~ "Somatic data helpful, AUC lower CI > 0.6",
        AUC_full_upper < 0.4 ~ "Somatic data not helpful, AUC upper CI < 0.4",
        TRUE ~ "Underpowered/inconclusive"
      ),
      auc_label = ifelse(is.na(AUC_full), "", sprintf("%.2f", AUC_full))
    ) %>%
    dplyr::filter(gene %in% target_genes)

  gene_order_df <- gene_df_somatic %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(max_auc_lower = max(AUC_full_lower, na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(max_auc_lower))

  gene_levels    <- rev(gene_order_df$gene)
  common_genes   <- gene_levels

  gene_df_somatic <- gene_df_somatic %>%
    dplyr::mutate(gene = factor(as.character(gene), levels = gene_levels))

  somatic_ci_cols <- c(
    "Somatic data helpful, AUC lower CI > 0.8"       = "#a63642",
    "Somatic data helpful, AUC lower CI > 0.7"       = "#d6605c",
    "Somatic data helpful, AUC lower CI > 0.6"       = "#f07f78",
    "Underpowered/inconclusive"                       = "#fffde7",
    "Somatic data not helpful, AUC upper CI < 0.4"   = "#f6e58d",
    "missing"                                         = "grey70"
  )

  p_heat <- ggplot() +
    geom_tile(data = gene_df_somatic,
              aes(x = Predictor_simple, y = gene, fill = somatic_ci_cat), colour = "grey20") +
    geom_text(data = gene_df_somatic,
              aes(x = Predictor_simple, y = gene, label = auc_label), size = 3) +
    scale_fill_manual(name = "Somatic AUC confidence group", values = somatic_ci_cols) +
    theme_bw(base_size = 12) +
    theme(
      axis.text.y = element_text(size = 10),
      axis.text.x = element_text(angle = 60, hjust = 1),
      panel.grid  = element_blank(),
      legend.position = "left"
    ) +
    labs(x = "AUC \n Clinvar Vs UK Biobank", y = " ")

  gene_counts_df <- var_table %>%
    dplyr::filter(gene %in% common_genes) %>%
    tidyr::pivot_longer(
      cols = c(n_unique_pathogenic, n_pathogenic, n_benign),
      names_to = "metric", values_to = "value"
    ) %>%
    dplyr::mutate(
      metric = dplyr::recode(metric,
        n_unique_pathogenic = "n unique pathogenic variants",
        n_pathogenic        = "n pathogenic variants",
        n_benign            = "n benign variants"
      ),
      metric = factor(metric, levels = c(
        "n unique pathogenic variants", "n pathogenic variants", "n benign variants"
      )),
      gene  = factor(gene, levels = gene_levels),
      label = ifelse(is.na(value), "", format(value, big.mark = ","))
    )

  p_table <- ggplot(gene_counts_df, aes(x = metric, y = gene)) +
    geom_tile(fill = "white", colour = "grey80") +
    geom_text(aes(label = label), size = 3) +
    scale_x_discrete(position = "top") +
    labs(x = NULL, y = NULL) +
    theme_bw(base_size = 12) +
    theme(
      panel.grid   = element_blank(),
      axis.text.y  = element_blank(),
      axis.ticks.y = element_blank(),
      axis.text.x  = element_text(angle = 80, hjust = 0),
      plot.margin  = margin(t = 5.5, r = 5.5, b = 5.5, l = 0)
    )

  p_groups <- NULL

  if (!is.null(gene_group_cols) && length(gene_group_cols) > 0) {
    gene_group_cols <- intersect(gene_group_cols, colnames(var_table))

    if (length(gene_group_cols) > 0) {
      gene_groups_df <- var_table %>%
        dplyr::filter(gene %in% common_genes) %>%
        dplyr::select(gene, dplyr::all_of(gene_group_cols)) %>%
        tidyr::pivot_longer(
          cols = dplyr::all_of(gene_group_cols), names_to = "group", values_to = "value"
        ) %>%
        dplyr::mutate(
          gene      = factor(gene, levels = gene_levels),
          group     = factor(group, levels = gene_group_cols),
          value_chr = tolower(as.character(value)),
          is_true   = dplyr::case_when(
            value_chr %in% c("true","t","1","yes","y") ~ TRUE,
            value_chr %in% c("false","f","0","no","n") ~ FALSE,
            TRUE ~ FALSE
          ),
          fill_val  = dplyr::if_else(is_true, "TRUE", NA_character_)
        )

      p_groups <- ggplot(gene_groups_df, aes(x = group, y = gene)) +
        geom_tile(aes(fill = fill_val), colour = "grey80") +
        scale_x_discrete(position = "top") +
        scale_fill_manual(values = c("TRUE" = "#2171B5"), na.value = "white", guide = "none") +
        labs(x = NULL, y = NULL) +
        theme_bw(base_size = 12) +
        theme(
          panel.grid   = element_blank(),
          axis.text.y  = element_blank(),
          axis.ticks.y = element_blank(),
          axis.text.x  = element_text(angle = 80, hjust = 0),
          plot.margin  = margin(t = 5.5, r = 5.5, b = 5.5, l = 0)
        )
    }
  }

  if (is.null(p_groups)) {
    (p_heat + p_table) + patchwork::plot_layout(widths = c(4.5, 4))
  } else {
    (p_heat + p_table + p_groups) +
      patchwork::plot_layout(widths = c(4.5, 4, length(gene_group_cols)))
  }
}
```

### Top 25 genes

Genes with the highest AUC_lower_CI for the COSMIC_codon_count predictor:

```{r, fig.width=10, fig.height=6}
genes_top_25_long <- gene_AUCs_individual_features %>%
  filter(Predictor == "n_COSMIC_codon_count", dataset == "clinvar_long") %>%
  arrange(desc(AUC_full_lower)) %>%
  head(25) %>% pull(gene)

make_gene_panel(genes_top_25_long, dataset_name = "clinvar_long",
                gene_group_cols = c("AltFunc gene group", "PI3K-MTOR pathway", "RAS-MAPK pathway"))

ggsave("../../Data/Figures/Top_25_genes_clinvar_long.png", width = 10, height = 10)
```

### Pathway defined genes: PIK3CA-MTOR and RAS-MAPK pathway genes

Showing clinvar long data:

```{r, fig.width=10, fig.height=10}
genes_ras_mtor <- unique(c(pi3k_mtor_genes, ras_mapk_genes))

make_gene_panel(genes_ras_mtor, dataset_name = "clinvar_long",
                gene_group_cols = c("chromatin_modifier", "kinase", "protein_phosphatase",
                                    "pore_forming_ion_channel"))

ggsave("../../Data/Figures/PI3K_MTOR_RAS_MAPK_genes_clinvar_long.png", width = 10, height = 10)
```

### Macrocephaly genes that are also somatic selection monoallelic DDG2P genes

Showing clinvar long data:

```{r, fig.width=10, fig.height=10}
make_gene_panel(panelapp_macrocephaly_somatic, dataset_name = "clinvar_long")
ggsave("../../Data/Figures/Macrocephaly_genes_clinvar_long.png", width = 10, height = 10)
```

# Conclusions

Clinical definitions of gene groups (Brain malformations, macrocephaly, RASopathy) are more enriched for somatic AUC helpfulness than GO or pathway defined groups.
