---
title: "Gene group somatic utility analysis"
output:
  html_document:
    df_print: paged
    code_folding: hide
  pdf_document: default
date: "2026-05-07"
---

# Aim

Summarise the utility of somatic (COSMIC) data for germline variant interpretation
across clinically and biologically defined gene groups (VCEP, PanelApp, KEGG
pathways). Produces Table 3 for the manuscript.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
theme_set(theme_bw(14))

dir.create("../../Data/Tables",  recursive = TRUE, showWarnings = FALSE)
dir.create("../../Data/Figures", recursive = TRUE, showWarnings = FALSE)

recode_data_name <- function(x) {
  dplyr::case_match(x,
    c("clinvar_long", "clinvar_long_uncontaminated") ~ "clinvar_count_weighted",
    c("DNM_long", "DNM_long_uncontaminated")         ~ "DNM_count_weighted",
    .default = x
  )
}
```

# Gene lists

## DDG2P

```{r ddg2p, message=FALSE, warning=FALSE}
DDG2P <- read_csv("../../Data/DDG2P/G2P_DD_2025-04-16.csv",
                  show_col_types = FALSE) %>%
  clean_names()

DDG2P_monoallelic <- DDG2P %>%
  filter(allelic_requirement %in% c("monoallelic_autosomal",
                                    "monoallelic_X_heterozygous"))

cat("DDG2P monoallelic genes:", length(unique(DDG2P_monoallelic$gene_symbol)), "\n")
```

## VCEP-defined gene lists

```{r vcep, message=FALSE, warning=FALSE}
VCEP_rasopathy <- read_csv(
  "../../Data/Gene_definitions/RASopathy_genes_VCEP_030425.csv",
  show_col_types = FALSE
) %>% pull(gene)

# PIK3CA/MTOR pathway brain malformation VCEP genes (hardcoded)
VCEP_PIK3CA_MTOR <- c("AKT3", "PIK3CA", "PIK3R2", "MTOR", "DEPDC5")

cat("VCEP RASopathy genes:", length(VCEP_rasopathy), "\n")
cat("VCEP brain malformation genes:", length(VCEP_PIK3CA_MTOR), "\n")
```

## PanelApp gene lists

```{r panelapp, message=FALSE, warning=FALSE}
# UK PanelApp: RAS-MAPK pathway (RASopathy-associated genes)
panelapp_RAS <- read_tsv(
  "../../Data/Panels/RASopathy_panelapp_green_list_150126.tsv",
  show_col_types = FALSE
) %>% clean_names() %>%
  pull(gene_symbol) %>% unique()

# UK PanelApp: Segmental overgrowth — union of two panels
# (Segmental overgrowth disorders -- Deep sequencing, panel 98; and
#  Neurological segmental overgrowth, panel 736)
seg_panel <- union(
  read_tsv(
    "../../Data/Panels/PanelAll_UK_Neurological_segmental_overgrowth_green_genes_120326.tsv",
    show_col_types = FALSE
  ) %>% clean_names() %>% pull(gene_symbol),
  read_tsv(
    "../../Data/Panels/PanelApp_UK_segmental_overgrowth_green_genes_120326.tsv",
    show_col_types = FALSE
  ) %>% clean_names() %>% pull(gene_symbol)
) %>% unique()

# Aus PanelApp: Macrocephaly/megalencephaly, restricted to monoallelic DDG2P genes
panelapp_macrocephaly <- read_tsv(
  "../../Data/Panels/PanelApp_Aus_Macrocephaly_Megalencephaly_green_genes_120326.tsv",
  show_col_types = FALSE
) %>% clean_names() %>%
  filter(gene_symbol %in% DDG2P_monoallelic$gene_symbol) %>%
  pull(gene_symbol) %>% unique()

cat("Panel: RAS-MAPK genes:", length(panelapp_RAS), "\n")
cat("Panel: Segmental overgrowth genes:", length(seg_panel), "\n")
cat("Panel: Dominant macrocephaly genes (Aus PanelApp x DDG2P monoallelic):",
    length(panelapp_macrocephaly), "\n")
```

## Signalling pathway gene lists

```{r pathways}
# WikiPathways: PI3K-MTOR signalling
pi3k_mtor_genes <- c(
  "RICTOR", "MTOR",   "RHEB",    "BAD",    "PIK3R2",  "PIK3CA",
  "PIK3R3", "GRB10",  "KRAS",    "RB1CC1", "RPTOR",   "HRAS",
  "FOXO3",  "ULK1",   "EIF4EBP1","TSC2",   "CDKN1B",  "NRAS",
  "PDK1",   "PIK3CB", "NOS3",    "PTEN",   "PIK3CG",  "FOXO1",   "GSK3B"
)

# KEGG: RAS-MAPK signalling
ras_mapk_genes <- c(
  "KRAS",   "HRAS",   "NRAS",   "MRAS",
  "RAF1",   "ARAF",   "BRAF",
  "MAP2K1", "MAP2K2", "MAPK1",  "MAPK3",
  "SOS1",   "SOS2",   "GRB2",   "RASA1",  "RASA2",
  "EGFR",   "FGFR1",  "FGFR2",  "FGFR3",
  "PDGFRA", "PDGFRB", "ERBB2",  "ERBB3",  "NTRK1",   "NTRK2",   "IGF1R"
)

cat("PI3K-MTOR pathway genes:", length(pi3k_mtor_genes), "\n")
cat("RAS-MAPK (KEGG) genes:", length(ras_mapk_genes), "\n")
```

# Load AUC data

```{r load-auc, message=FALSE, warning=FALSE}
date <- "240426"

gene_AUCs_individual_features <- read_csv(
  paste0("../../Data/AUC_tables/Gene_group_hold_out_models/",
         "Single_feature_AUCs_individual_genes_", date, ".csv"),
  show_col_types = FALSE
) %>%
  mutate(dataset = recode_data_name(dataset))

n_vars <- read_csv(
  paste0("../../Data/AUC_tables/Gene_group_hold_out_models/",
         "Variant_counts_per_gene_", date, ".csv"),
  show_col_types = FALSE
)

# Restrict to COSMIC codon count predictor in the clinvar_count_weighted dataset
clinvar_long_COSMIC <- gene_AUCs_individual_features %>%
  filter(Predictor == "n_COSMIC_codon_count",
         dataset  == "clinvar_count_weighted") %>%
  arrange(desc(AUC_full_lower)) %>%
  left_join(
    n_vars %>% filter(data_set == "clinvar_mutation_data_long"),
    by = "gene"
  )

cat("Genes with clinvar_long COSMIC codon count AUC:", nrow(clinvar_long_COSMIC), "\n")
```

# Annotate genes with gene group flags

```{r annotate}
clinvar_long_COSMIC_annot <- clinvar_long_COSMIC %>%
  mutate(gene = trimws(gene)) %>%
  mutate(
    Brain_malformations_VCEP      = gene %in% VCEP_PIK3CA_MTOR,
    RASopathy_VCEP                = gene %in% VCEP_rasopathy,
    `RAS-MAPK_PanelApp`           = gene %in% panelapp_RAS,
    Segmental_overgrowth_PanelApp = gene %in% seg_panel,
    Macrocephaly_PanelApp         = gene %in% panelapp_macrocephaly,
    `PI3K-MTOR pathway`           = gene %in% pi3k_mtor_genes,
    `RAS-MAPK pathway`            = gene %in% ras_mapk_genes
  )

# Diagnostic: genes per group (among those with computed AUC)
clinvar_long_COSMIC_annot %>%
  summarise(across(where(is.logical), sum, na.rm = TRUE)) %>%
  pivot_longer(everything(), names_to = "group", values_to = "n_genes_with_AUC")
```

# Table 3: Percentage of genes with somatic data utility

```{r table-3}
table_3 <- clinvar_long_COSMIC_annot %>%
  dplyr::select(where(is.logical), AUC_full_lower, AUC_full_upper) %>%
  pivot_longer(cols = where(is.logical),
               names_to  = "category",
               values_to = "flag") %>%
  filter(flag) %>%
  group_by(category) %>%
  summarise(
    `n genes` = n(),
    `n AUC lower > 0.6` = sprintf(
      "%d (%.1f%%)",
      sum(AUC_full_lower > 0.6, na.rm = TRUE),
      100 * mean(AUC_full_lower > 0.6, na.rm = TRUE)
    ),
    `n AUC lower > 0.8` = sprintf(
      "%d (%.1f%%)",
      sum(AUC_full_lower > 0.8, na.rm = TRUE),
      100 * mean(AUC_full_lower > 0.8, na.rm = TRUE)
    ),
    .groups = "drop"
  ) %>%
  mutate(category = case_when(
    category == "Brain_malformations_VCEP"      ~ "VCEP: Brain malformations (PIK3CA/MTOR)",
    category == "RASopathy_VCEP"                ~ "VCEP: RASopathies",
    category == "RAS-MAPK_PanelApp"             ~ "Panel: RAS-MAPK (UK PanelApp)",
    category == "Segmental_overgrowth_PanelApp" ~ "Panel: Segmental overgrowth (UK PanelApp)",
    category == "Macrocephaly_PanelApp"         ~ "Phenotype: Dominant Macrocephaly",
    category == "PI3K-MTOR pathway"             ~ "Pathway: PI3K-MTOR (WikiPathways)",
    category == "RAS-MAPK pathway"              ~ "Pathway: RAS-MAPK (KEGG)",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(category)) %>%
  mutate(category = factor(category, levels = c(
    "VCEP: Brain malformations (PIK3CA/MTOR)",
    "VCEP: RASopathies",
    "Panel: RAS-MAPK (UK PanelApp)",
    "Panel: Segmental overgrowth (UK PanelApp)",
    "Phenotype: Dominant Macrocephaly",
    "Pathway: PI3K-MTOR (WikiPathways)",
    "Pathway: RAS-MAPK (KEGG)"
  ))) %>%
  arrange(category)

table_3
```

```{r write-table-3}
write_csv(table_3, "../../Data/Tables/Table_3_gene_groups_and_somatic_AUCs.csv")
```
