---
title: "Somatic germline paper tables"
output:
  html_document:
    df_print: paged
    code_folding: hide
date: "2026-02-05"
---

# Aim
Generate and export all manuscript tables in order of appearance.

Note: Tables 1 and 2 are hard-coded in LaTeX and do not require R generation.

Supplementary Table 3 (list of MAVE studies) is compiled manually and not generated here.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(DT)
library(knitr)

date <- "240426"

dir.create("../../Data/Tables", recursive = TRUE, showWarnings = FALSE)

recode_data_name <- function(x) {
  dplyr::case_match(x,
    "clinvar_long_uncontaminated" ~ "clinvar_count_weighted",
    "DNM_long_uncontaminated"     ~ "DNM_count_weighted",
    "clinvar_long"                ~ "clinvar_count_weighted",
    "DNM_long"                    ~ "DNM_count_weighted",
    .default = x
  )
}
```

```{r, warning=FALSE, include=FALSE}
Clinvar_vs_Biobank <- read_csv("../../Data/ML_model_input/Gene_group_models/Clinvar_vs_Biobank_unique_set1_240426.csv.gz")
DNM_vs_Biobank     <- read_csv("../../Data/ML_model_input/Gene_group_models/DNM_vs_Biobank_unique_set1_240426.csv.gz")
```

```{r, warning=FALSE, message=FALSE}
altfunc           <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_AltFunc.csv") %>% pull(gene_symbol)
altfunc_stringent <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_AltFunc_stringent.csv") %>% pull(gene_symbol)
lof               <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_LOF.csv") %>% pull(gene_symbol)
other             <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_other.csv") %>% pull(gene_symbol)
RASopathies       <- read_csv("../../Data/Gene_definitions/RASopathy_genes_VCEP_030425.csv", show_col_types = FALSE) %>% pull(gene)
altfunc_minus_RAS <- setdiff(altfunc, RASopathies)
macrocephaly      <- read_csv("../../Data/Gene_definitions/Macrocephaly_monoallelic_DD_somatic_genes.csv") %>% pull(gene)
```

```{r, warning=FALSE, include=FALSE}
AUCs_individual_features <- read_csv(paste0("../../Data/AUC_tables/Gene_group_models/AUC_table_gene_groups_single_features_", date, ".csv")) %>%
  mutate(data_name = recode_data_name(data_name))

AUCs_LR_model <- read_csv(paste0("../../Data/AUC_tables/Gene_group_models/AUC_table_gene_groups_LR_models_", date, ".csv")) %>%
  mutate(Predictor = paste0(cep, "_LR_model_", test_or_train, "_data"),
         data_name = recode_data_name(data_name))

AUCs <- bind_rows(AUCs_LR_model, AUCs_individual_features) %>%
  dplyr::select(Predictor, data_name, gene_group,
                AUC_full, AUC_full_lower, AUC_full_upper, everything()) %>%
  dplyr::select(-any_of("gene_set"))

n_genes <- read_csv(paste0("../../Data/AUC_tables/Gene_group_models/Gene_counts_per_group_", date, ".csv")) %>%
  mutate(data_name = recode_data_name(data_name))
n_vars  <- read_csv(paste0("../../Data/AUC_tables/Gene_group_models/Variant_counts_per_group_", date, ".csv")) %>%
  mutate(data_name = recode_data_name(data_name))

performance <- read_csv("../../Data/Tables/Gene_group_single_feature_metrics.csv") %>%
  mutate(data_name = recode_data_name(data_name))
```

```{r, warning=FALSE, message=FALSE}
gene_AUCs_individual_features <- read_csv("../../Data/AUC_tables/Gene_group_hold_out_models/Single_feature_AUCs_individual_genes_240426.csv") %>%
  mutate(dataset = recode_data_name(dataset))
gene_AUCs_LR_model            <- read_csv("../../Data/AUC_tables/Gene_group_hold_out_models/AUC_table_gene_holdout_LR_models_240426.csv") %>%
  mutate(dataset = recode_data_name(dataset))
n_vars_per_gene               <- read_csv("../../Data/AUC_tables/Gene_group_hold_out_models/Variant_counts_per_gene_240426.csv")

n_vars_per_gene_clinvar <- n_vars_per_gene %>%
  filter(data_set %in% c("clinvar_mutation_data_unique", "clinvar_mutation_data_long")) %>%
  pivot_wider(names_from = data_set, values_from = c(n_path_variants, n_benign_variants)) %>%
  dplyr::select(-n_benign_variants_clinvar_mutation_data_long) %>%
  dplyr::rename(
    n_benign            = n_benign_variants_clinvar_mutation_data_unique,
    n_unique_pathogenic = n_path_variants_clinvar_mutation_data_unique,
    n_pathogenic        = n_path_variants_clinvar_mutation_data_long
  )
```

```{r, warning=FALSE, message=FALSE}
LR <- read_csv("../../Data/LR_calculations/LR_table.csv")
```

```{r}
# Per-gene somatic AUC summary in clinvar_long dataset
genes_clinvar_long_auc <- gene_AUCs_individual_features %>%
  filter(dataset == "clinvar_count_weighted") %>%
  filter(Predictor %in% c("n_COSMIC_codon_count", "n_sperm_codon_count", "n_buccal_codon_count")) %>%
  group_by(gene) %>%
  summarise(max_AUC_lower = max(AUC_full_lower, na.rm = TRUE), .groups = "drop") %>%
  filter(is.finite(max_AUC_lower)) %>%
  mutate(
    auc_lower_06 = max_AUC_lower > 0.6,
    auc_lower_08 = max_AUC_lower > 0.8
  )

genes_clinvar_long <- genes_clinvar_long_auc$gene
```

---

# Table 3: Number of genes with somatic predictive power

```{r}
table3 <- read_csv("../../Data/Tables/Table_3_gene_groups_and_somatic_AUCs.csv",
                   show_col_types = FALSE)
knitr::kable(table3, caption = "Table 3: Genes with somatic predictive power by gene group")
```

---

# Supplementary Table 1: Full benchmarking results (all gene groups)

Single-feature predictor performance metrics for all gene groups across dataset types.

```{r}
supp_table1 <- performance %>%
  dplyr::select(-any_of("gene_set")) %>%
  dplyr::select(gene_group, data_name, everything())

write_csv(supp_table1, "../../Data/Tables/Supplementary_Table1.csv")

DT::datatable(supp_table1,
              caption = "Supplementary Table 1: Full benchmarking results",
              filter = "top", rownames = FALSE,
              options = list(pageLength = 20, scrollX = TRUE))
```

---

# Supplementary Table 2: AUC for each feature for each gene

Single-feature AUC values for individual genes in both clinvar_unique and clinvar_count_weighted datasets.

```{r}
supp_table2 <- gene_AUCs_individual_features %>%
  filter(dataset %in% c("clinvar_unique", "clinvar_count_weighted")) %>%
  dplyr::select(gene, dataset, everything())

write_csv(supp_table2, "../../Data/Tables/Supplementary_Table2.csv")

DT::datatable(supp_table2,
              caption = "Supplementary Table 2: Individual gene single-feature AUCs",
              filter = "top", rownames = FALSE,
              options = list(pageLength = 20, scrollX = TRUE))
```

---

# Supplementary Table 3: List of MAVE studies

_This table lists the deep mutational scanning studies used in this analysis. It is compiled manually and not generated by R._

---

# Supplementary Table 4: Performance of LR models with DeLong p-values

Gene group-level logistic regression model performance (test set), with DeLong test comparing REVEL-only vs REVEL + somatic data.

```{r}
supp_table4 <- AUCs_LR_model %>%
  filter(test_or_train == "test") %>%
  dplyr::select(gene_group, data_name, cep,
                AUC_full, AUC_full_lower, AUC_full_upper,
                AUC_cep_only, AUC_cep_lower, AUC_cep_upper,
                AUC_diff, AUC_diff_lower, AUC_diff_upper,
                delong_p_value)

write_csv(supp_table4, "../../Data/Tables/Supplementary_Table4.csv")

DT::datatable(supp_table4,
              caption = "Supplementary Table 4: LR model performance with DeLong p-values",
              filter = "top", rownames = FALSE,
              options = list(pageLength = 20, scrollX = TRUE))
```

---

# Supplementary Table 5: Logistic regression model results for individual genes

Individual gene-level logistic regression model performance (test set), with DeLong test comparing REVEL-only vs REVEL + somatic data.

```{r}
supp_table5 <- gene_AUCs_LR_model %>%
  filter(test_or_train == "test") %>%
  dplyr::select(gene, dataset, everything(), -test_or_train)

write_csv(supp_table5, "../../Data/Tables/Supplementary_Table5.csv")

DT::datatable(supp_table5,
              caption = "Supplementary Table 5: Individual gene LR model performance with DeLong p-values",
              filter = "top", rownames = FALSE,
              options = list(pageLength = 20, scrollX = TRUE))
```

---

# Supplementary Table 6: Likelihood ratios for somatic feature cutoffs

Positive and negative likelihood ratios for COSMIC and Nanoseq feature thresholds across all four pathogenic variant dataset forms (ClinVar count-weighted and unique, and de novo mutation count-weighted and unique).

```{r}
supp_table6 <- LR %>%
  filter(data_set %in% c("clinvar_long_public", "clinvar_unique_public",
                         "DNMs_31k_long_public", "DNMs_31k_unique_public")) %>%
  arrange(data_set, gene_group, feature)

write_csv(supp_table6, "../../Data/Tables/Supplementary_Table6.csv")

DT::datatable(supp_table6,
              caption = "Supplementary Table 6: Likelihood ratios",
              filter = "top", rownames = FALSE,
              options = list(pageLength = 20, scrollX = TRUE))
```
