---
title: "Likelihood ratio threshold plots"
output:
  html_document:
    df_print: paged
    code_folding: hide
  pdf_document: default
date: "2026-02-27"
---

# Aim

Examine the LRs of different somatic features across:

* gene groups
* pathogenic variant sources (clinvar vs DNMs)
* long vs unique form data

```{r, message=FALSE}
library(tidyverse)
library(patchwork)
library(ggrepel)

theme_set(theme_bw(14))

safe_colorblind_palette <- c("#88CCEE", "#CC6677", "#117733", "#DDCC77", "#332288", "#AA4499",
                             "#44AA99", "#999933", "#882255", "#661100", "#6699CC", "#888888")

dir.create("../../Data/Figures", recursive = TRUE, showWarnings = FALSE)
```

```{r, message=FALSE}
LR <- read_csv("../../Data/LR_calculations/LR_table.csv")
```

```{r}
acmg_lr_pos <- tibble::tibble(
  level = c("Supporting", "Moderate", "Strong"),
  lr = c(2.08, 4.33, 18.7)
)

acmg_lr_neg <- tibble::tibble(
  level = c("Supporting", "Moderate", "Strong"),
  lr = c(0.48, 0.23, 0.05)
)
```

```{r}
LR_edit <- LR %>%
  mutate(gene_group = case_when(
    gene_group == "AltFunc"          ~ "Altered function genes",
    gene_group == "RASopathies"      ~ "RASopathy genes",
    gene_group == "AltFunc_minus_RAS"~ "Non-RASopathy Altered function genes",
    gene_group == "LOF"              ~ "Loss of function genes",
    gene_group == "Other"            ~ "Other dominant cancer-DD genes",
    gene_group == "AllGenes"         ~ "All DD genes",
    gene_group == "MacrocephalyGenes"~ "Macrocephaly genes"
  ))
```


# All gene groups, single features

## Clinvar unique with twinsUK sperm data {.tabset}

### COSMIC codon count

```{r, fig.width=9, fig.height=5}
LR_edit %>%
  filter(data_set == "clinvar_unique_public") %>%
  filter(!threshold == "1-4") %>%
  dplyr::filter(feature == "n_COSMIC_codon_count") %>%
  dplyr::arrange(gene_group, sensitivity) %>%
  ggplot(aes(x = sensitivity, y = LR_pos, colour = gene_group, group = gene_group)) +

  geom_errorbar(
    aes(ymin = LR_pos_low95, ymax = LR_pos_high95),
    width = 0.005,
    linewidth = 0.4,
    alpha = 0.7
  ) +

  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +

  geom_hline(
    data = acmg_lr_pos,
    aes(yintercept = lr),
    linetype = "dotted",
    colour = "grey40",
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +

  geom_text(
    data = acmg_lr_pos,
    aes(x = 1.1, y = lr, label = level),
    hjust = 1.05, vjust = -0.1,
    colour = "grey40",
    size = 3,
    inherit.aes = FALSE
  ) +

  geom_text_repel(
    data = ~ dplyr::filter(.x, gene_group == "RASopathy genes"),
    aes(label = threshold),
    size = 3,
    show.legend = FALSE
  ) +

  scale_y_log10() +
  coord_cartesian(clip = "off") +
  labs(y = "Positive likelihood ratio") +
  scale_colour_manual(values = safe_colorblind_palette) +
  theme_bw() +
  theme(plot.margin = margin(5.5, 30, 5.5, 5.5))

ggsave("../../Data/Figures/COSMIC_codon_count_LR_plot.png", width = 9, height = 5, dpi = 300)
```

### COSMIC count

```{r, fig.width=9, fig.height=5}
LR_edit %>%
  filter(data_set == "clinvar_unique_public") %>%
  filter(!threshold == "1-4") %>%
  dplyr::filter(feature == "n_COSMIC") %>%
  dplyr::arrange(gene_group, sensitivity) %>%

  ggplot(aes(x = sensitivity, y = LR_pos, colour = gene_group, group = gene_group)) +

  geom_errorbar(
    aes(ymin = LR_pos_low95, ymax = LR_pos_high95),
    width = 0.005,
    linewidth = 0.4,
    alpha = 0.7
  ) +

  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +

  geom_hline(
    data = acmg_lr_pos,
    aes(yintercept = lr),
    linetype = "dotted",
    colour = "grey40",
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +

  geom_text(
    data = acmg_lr_pos,
    aes(x = 1.1, y = lr, label = level),
    hjust = 1.05, vjust = -0.1,
    colour = "grey40",
    size = 3,
    inherit.aes = FALSE
  ) +

  geom_text_repel(
    data = ~ dplyr::filter(.x, gene_group == "RASopathy genes"),
    aes(label = threshold),
    size = 3,
    show.legend = FALSE
  ) +

  scale_y_log10() +
  coord_cartesian(clip = "off") +
  labs(y = "Positive likelihood ratio") +
  scale_colour_manual(values = safe_colorblind_palette) +
  theme_bw() +
  theme(plot.margin = margin(5.5, 30, 5.5, 5.5))
```


# RASopathy genes, all features {.tabset}

```{r}
plot_LR_thresholds <- function(data_set_name) {
  plot_data <- LR_edit %>%
    filter(data_set == data_set_name) %>%
    filter(!threshold == "1-4") %>%
    filter(gene_group == "RASopathy genes") %>%
    arrange(feature, sensitivity)

  ggplot(plot_data,
         aes(x = sensitivity,
             y = LR_pos,
             colour = feature,
             group = feature)) +

    geom_errorbar(
      aes(ymin = LR_pos_low95,
          ymax = LR_pos_high95),
      width = 0.005,
      linewidth = 0.4,
      alpha = 0.7
    ) +

    geom_line(linewidth = 0.8) +
    geom_point(size = 1.5) +

    geom_hline(
      data = acmg_lr_pos,
      aes(yintercept = lr),
      linetype = "dotted",
      colour = "grey40",
      linewidth = 0.5,
      inherit.aes = FALSE
    ) +

    geom_text(
      data = acmg_lr_pos,
      aes(x = 1.1, y = lr, label = level),
      hjust = 1.05,
      vjust = -0.1,
      colour = "grey40",
      size = 3,
      inherit.aes = FALSE
    ) +

    geom_text_repel(
      aes(label = threshold),
      size = 3,
      direction = "y",
      force = 4,
      box.padding = 0.4,
      point.padding = 0.2,
      max.overlaps = Inf,
      segment.alpha = 0.6,
      show.legend = FALSE
    ) +

    scale_y_log10() +
    coord_cartesian(clip = "off") +
    labs(y = "Positive likelihood ratio") +
    scale_colour_manual(values = safe_colorblind_palette) +
    theme_bw() +
    theme(plot.margin = margin(5.5, 30, 5.5, 5.5))
}
```

## Clinvar unique with twinsUK sperm data

```{r}
LR_test_dat <- LR_edit %>%
  filter(data_set == "clinvar_unique_public") %>%
  filter(!threshold == "1-4") %>%
  filter(gene_group == "RASopathy genes") %>%
  arrange(feature, sensitivity)

plot_data <- LR_test_dat %>%
  dplyr::filter(
    is.finite(sensitivity),
    is.finite(LR_pos),
    LR_pos > 0
  )
```

```{r, fig.width=9, fig.height=5}
plot_LR_thresholds("clinvar_unique_public")
```

## Clinvar long with twinsUK sperm data

```{r}
plot_LR_thresholds("clinvar_long_public")
```

## DNM 31k long with twinsUK sperm data

```{r}
plot_LR_thresholds("DNMs_31k_long_public")
```


# Use hotcodon for strong evidence in any gene

Showing the LR of hotcodon for different data types:

```{r, fig.width=9, fig.height=3}
LR_hotcodon <- LR_edit %>%
  filter(feature == "hotcodon_cancer") %>%
  filter(grepl("public", data_set)) %>%
  arrange(gene_group, sensitivity) %>%
  mutate(data_set = case_when(
    data_set == "clinvar_unique_public"  ~ "ClinVar unique",
    data_set == "clinvar_long_public"    ~ "ClinVar long",
    data_set == "DNMs_31k_long_public"   ~ "DNM 31k long",
    TRUE ~ data_set
  ))

ggplot(LR_hotcodon, aes(y = data_set, x = LR_pos)) +
  geom_errorbarh(
    aes(xmin = LR_pos_low95, xmax = LR_pos_high95),
    width = 0.2,
    alpha = 0.7
  ) +
  geom_vline(
    data = acmg_lr_pos,
    aes(xintercept = lr),
    linetype = "dotted",
    colour = "grey40",
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = acmg_lr_pos,
    aes(x = lr, y = Inf, label = level),
    angle = 90,
    colour = "grey50",
    size = 3,
    vjust = 1,
    hjust = 1.1,
    inherit.aes = FALSE
  ) +
  geom_point(size = 2) +
  scale_x_log10() +
  facet_wrap(~ gene_group) +
  labs(x = "Positive likelihood ratio", y = "Data set") +
  theme_bw() +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(5.5, 20, 5.5, 5.5))
```


# Use n_COSMIC_codon_count > 1 for supporting evidence in RASopathy genes

```{r, fig.width=7, fig.height=3}
LR_plot <- LR_edit %>%
  filter(feature == "n_COSMIC_codon_count") %>%
  filter(threshold == ">1") %>%
  filter(grepl("public", data_set)) %>%
  arrange(gene_group, sensitivity)

ggplot(LR_plot, aes(y = data_set, x = LR_pos)) +
  geom_errorbarh(
    aes(xmin = LR_pos_low95, xmax = LR_pos_high95),
    width = 0.2,
    alpha = 0.7
  ) +
  geom_vline(
    data = acmg_lr_pos,
    aes(xintercept = lr),
    linetype = "dotted",
    colour = "grey40",
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = acmg_lr_pos,
    aes(x = lr, y = Inf, label = level),
    angle = 90,
    colour = "grey50",
    size = 3,
    vjust = 1,
    hjust = 1.1,
    inherit.aes = FALSE
  ) +
  geom_point(size = 2) +
  scale_x_log10() +
  facet_wrap(~ gene_group) +
  labs(x = "Positive likelihood ratio", y = "Data set") +
  theme_bw() +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(5.5, 20, 5.5, 5.5))
```


# Use n_COSMIC_codon_count > 2 for moderate evidence in RASopathy genes

```{r, fig.width=7, fig.height=3}
LR_plot <- LR_edit %>%
  filter(feature == "n_COSMIC_codon_count") %>%
  filter(threshold == ">2") %>%
  filter(grepl("public", data_set)) %>%
  arrange(gene_group, sensitivity)

ggplot(LR_plot, aes(y = data_set, x = LR_pos)) +
  geom_errorbarh(
    aes(xmin = LR_pos_low95, xmax = LR_pos_high95),
    width = 0.2,
    alpha = 0.7
  ) +
  geom_vline(
    data = acmg_lr_pos,
    aes(xintercept = lr),
    linetype = "dotted",
    colour = "grey40",
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = acmg_lr_pos,
    aes(x = lr, y = Inf, label = level),
    angle = 90,
    colour = "grey50",
    size = 3,
    vjust = 1,
    hjust = 1.1,
    inherit.aes = FALSE
  ) +
  geom_point(size = 2) +
  scale_x_log10() +
  facet_wrap(~ gene_group) +
  labs(x = "Positive likelihood ratio", y = "Data set") +
  theme_bw() +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(5.5, 20, 5.5, 5.5))
```

# Use n_COSMIC_codon_count thresholds in RASopathy genes

## Positive LRs

```{r, fig.width=3, fig.height=7}
LR_plot <- LR_edit %>%
  filter(feature == "n_COSMIC_codon_count") %>%
  filter(!threshold %in% c("1-4", ">20", ">50")) %>%
  filter(gene_group == "RASopathy genes") %>%
  filter(grepl("public", data_set)) %>%
  mutate(data_set = case_when(
    data_set == "clinvar_unique_public" ~ "ClinVar unique",
    data_set == "clinvar_long_public"   ~ "ClinVar long",
    data_set == "DNMs_31k_long_public"  ~ "DNM 31k long",
    TRUE ~ data_set
  )) %>%
  mutate(threshold = factor(threshold, levels = c(">0", ">1", ">2", ">3", ">5", ">10")))

ggplot(LR_plot, aes(y = data_set, x = LR_pos)) +
  geom_errorbarh(
    aes(xmin = LR_pos_low95, xmax = LR_pos_high95),
    width = 0.2,
    alpha = 0.7
  ) +
  geom_vline(
    data = acmg_lr_pos,
    aes(xintercept = lr),
    linetype = "dotted",
    colour = "grey40",
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = acmg_lr_pos,
    aes(x = lr, y = Inf, label = level),
    angle = 90,
    colour = "grey50",
    size = 3,
    vjust = 1,
    hjust = 1.1,
    inherit.aes = FALSE
  ) +
  geom_point(size = 2) +
  scale_x_log10() +
  facet_wrap(~ threshold, ncol = 1) +
  labs(x = "Positive likelihood ratio", y = "") +
  theme_bw() +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(5.5, 20, 5.5, 5.5))

ggsave("../../Data/Figures/COSMIC_codon_count_LR_thresholds_plot.png", width = 3, height = 7, dpi = 300)
```

## Negative LR

```{r, fig.width=4, fig.height=2}
ggplot(LR_plot %>% filter(threshold == ">0") %>%
         mutate(threshold = "0"),
       aes(y = data_set, x = LR_neg)) +
  geom_errorbarh(
    aes(xmin = LR_neg_low95, xmax = LR_neg_high95),
    width = 0.2,
    alpha = 0.7
  ) +
  geom_vline(
    data = acmg_lr_neg,
    aes(xintercept = lr),
    linetype = "dotted",
    colour = "grey40",
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = acmg_lr_neg,
    aes(x = lr, y = Inf, label = level),
    angle = 90,
    colour = "grey50",
    size = 3,
    vjust = 1,
    hjust = 1.1,
    inherit.aes = FALSE
  ) +
  geom_point(size = 2) +
  facet_wrap(~ factor(threshold, levels = c("0")), ncol = 1) +
  scale_x_log10() +
  labs(x = "Negative likelihood ratio", y = "") +
  theme_bw() +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(5.5, 20, 5.5, 5.5))

ggsave("../../Data/Figures/COSMIC_codon_count_LR_thresholds_neg_plot.png", width = 4, height = 2, dpi = 300)
```
