---
title: "Likelihood ratio threshold calculations"
output: html_document
date: "2026-02-20"
---

# Aim

1. Calculate likelihood ratio thresholds for different somatic data features and counts to see what level of ACMG evidence this translates to if somatic data were to be used for germline clinical variant interpretation.

   - Contrast different data sources (ClinVar long vs ClinVar unique vs DNMs)
   - Add confidence intervals of LRs

2. Choose somatic evidence levels and annotate variant lists of interest

```{r, message=FALSE}
library(tidyverse)

dir.create("../../Data/LR_calculations", recursive = TRUE, showWarnings = FALSE)
```

# Load data

```{r, message=FALSE}
read_variants <- function(path) {
  read_tsv(path, show_col_types = FALSE) %>%
    dplyr::rename(
      n_sperm             = n_sperm_twinsUK,
      n_sperm_codon_count = n_sperm_twinsUK_codon_count
    )
}

clinvar_path   <- read_variants("../../Data/Variant_lists/Clinvar_DDG2P_missense_path_annot_260526_public.tsv.gz")
clinvar_VUS    <- read_variants("../../Data/Variant_lists/Clinvar_DDG2P_missense_VUS_annot_260526_public.tsv.gz")
clinvar_benign <- read_variants("../../Data/Variant_lists/Clinvar_DDG2P_missense_benign_annot_260526_public.tsv.gz")
DNMs_31k       <- read_variants("../../Data/Variant_lists/DNMs_31k_missense_annot_260526_public.tsv.gz")
biobank_benign <- read_variants("../../Data/Variant_lists/Biobank_missense_variants_annot_260526_public.tsv.gz")
```

```{r, message=FALSE}
altfunc       <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_AltFunc.csv",
                          show_col_types = FALSE) %>% pull(gene_symbol)
lof           <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_LOF.csv",
                          show_col_types = FALSE) %>% pull(gene_symbol)
other         <- read_csv("../../Data/Gene_definitions/DDG2P_2025_dominant_selection_genes_other.csv",
                          show_col_types = FALSE) %>% pull(gene_symbol)
RASopathies   <- read_csv("../../Data/Gene_definitions/RASopathy_genes_VCEP_030425.csv",
                          show_col_types = FALSE) %>% pull(gene)
altfunc_minus_RAS <- setdiff(altfunc, RASopathies)
macrocephaly  <- read_csv("../../Data/Gene_definitions/Macrocephaly_monoallelic_DD_somatic_genes.csv",
                          show_col_types = FALSE) %>% pull(gene)

all_genes <- unique(c(clinvar_path$gene, biobank_benign$gene, DNMs_31k$gene))
```

# Calculate LRs

```{r}
transmute_variants <- function(df, class_label) {
  df %>% transmute(
    gene,
    n_COSMIC             = as.numeric(n_COSMIC),
    n_sperm              = as.numeric(n_sperm),
    n_sperm_codon_count  = as.numeric(n_sperm_codon_count),
    n_COSMIC_codon_count = as.numeric(n_COSMIC_codon_count),
    hotcodon_cancer      = as.logical(hotcodon_cancer),
    class                = class_label
  )
}

ben <- transmute_variants(biobank_benign, "Benign")

Clinvar_unique_vs_biobank <- bind_rows(transmute_variants(clinvar_path, "Pathogenic"), ben)

Clinvar_long_vs_biobank <- bind_rows(
  clinvar_path %>%
    mutate(NumberSubmitters_clean = ifelse(is.na(NumberSubmitters_clean), 1L, NumberSubmitters_clean)) %>%
    uncount(NumberSubmitters_clean) %>%
    transmute_variants("Pathogenic"),
  ben
)

DNMs_31k_vs_biobank <- bind_rows(
  DNMs_31k %>%
    mutate(n_DNM = ifelse(is.na(n_DNM), 1L, n_DNM)) %>%
    uncount(n_DNM) %>%
    transmute_variants("Pathogenic"),
  ben
)

DNMs_31k_unique_vs_biobank <- bind_rows(
  transmute_variants(DNMs_31k, "Pathogenic"),
  ben
)

data_sets <- list(
  Clinvar_unique_vs_biobank   = Clinvar_unique_vs_biobank,
  Clinvar_long_vs_biobank     = Clinvar_long_vs_biobank,
  DNMs_31k_vs_biobank         = DNMs_31k_vs_biobank,
  DNMs_31k_unique_vs_biobank  = DNMs_31k_unique_vs_biobank
)
```

```{r}
gene_groups <- list(
  AllGenes          = all_genes,
  AltFunc           = altfunc,
  LOF               = lof,
  Other             = other,
  RASopathies       = RASopathies,
  AltFunc_minus_RAS = altfunc_minus_RAS,
  MacrocephalyGenes = macrocephaly
)
```

```{r}
rule_tbl <- tibble::tribble(
  ~rule, ~test_pos_expr,

  "n_COSMIC > 0",   quote(n_COSMIC > 0),
  "n_COSMIC > 1",   quote(n_COSMIC > 1),
  "n_COSMIC > 2",   quote(n_COSMIC > 2),
  "n_COSMIC > 3",   quote(n_COSMIC > 3),
  "n_COSMIC 1-4",   quote(n_COSMIC >= 1 & n_COSMIC <= 4),
  "n_COSMIC > 5",   quote(n_COSMIC > 5),
  "n_COSMIC > 10",  quote(n_COSMIC > 10),
  "n_COSMIC > 20",  quote(n_COSMIC > 20),
  "n_COSMIC > 50",  quote(n_COSMIC > 50),

  "n_sperm > 0",              quote(n_sperm > 0),
  "n_sperm > 1",              quote(n_sperm > 1),
  "n_sperm > 5",              quote(n_sperm > 5),

  "n_sperm_codon_count > 0",  quote(n_sperm_codon_count > 0),
  "n_sperm_codon_count > 1",  quote(n_sperm_codon_count > 1),
  "n_sperm_codon_count > 2",  quote(n_sperm_codon_count > 2),
  "n_sperm_codon_count > 3",  quote(n_sperm_codon_count > 3),
  "n_sperm_codon_count > 5",  quote(n_sperm_codon_count > 5),

  "n_COSMIC_codon_count > 0",  quote(n_COSMIC_codon_count > 0),
  "n_COSMIC_codon_count > 1",  quote(n_COSMIC_codon_count > 1),
  "n_COSMIC_codon_count > 2",  quote(n_COSMIC_codon_count > 2),
  "n_COSMIC_codon_count > 3",  quote(n_COSMIC_codon_count > 3),
  "n_COSMIC_codon_count 1-4",  quote(n_COSMIC_codon_count >= 1 & n_COSMIC_codon_count <= 4),
  "n_COSMIC_codon_count > 5",  quote(n_COSMIC_codon_count > 5),
  "n_COSMIC_codon_count > 10", quote(n_COSMIC_codon_count > 10),
  "n_COSMIC_codon_count > 20", quote(n_COSMIC_codon_count > 20),
  "n_COSMIC_codon_count > 50", quote(n_COSMIC_codon_count > 50),

  "hotcodon_cancer", quote(hotcodon_cancer == TRUE)
)
```

```{r}
calc_lr <- function(df, rule_name, test_pos_expr) {
  df2 <- df %>% mutate(test_pos = as.logical(eval(test_pos_expr)))

  TP <- sum(df2$class == "Pathogenic" & df2$test_pos,  na.rm = TRUE)
  FN <- sum(df2$class == "Pathogenic" & !df2$test_pos, na.rm = TRUE)
  FP <- sum(df2$class == "Benign"     & df2$test_pos,  na.rm = TRUE)
  TN <- sum(df2$class == "Benign"     & !df2$test_pos, na.rm = TRUE)

  if (any(c(TP, FN, FP, TN) == 0)) {
    TP <- TP + 0.5; FN <- FN + 0.5; FP <- FP + 0.5; TN <- TN + 0.5
  }

  sens <- TP / (TP + FN)
  spec <- TN / (TN + FP)

  lr_pos <- sens / (1 - spec)
  lr_neg <- (1 - sens) / spec

  se_log_lr_pos <- sqrt((1/TP) - (1/(TP+FN)) + (1/FP) - (1/(FP+TN)))
  se_log_lr_neg <- sqrt((1/FN) - (1/(TP+FN)) + (1/TN) - (1/(FP+TN)))

  tibble(
    rule         = rule_name,
    TP = TP, FN = FN, FP = FP, TN = TN,
    sensitivity    = sens,
    specificity    = spec,
    LR_pos         = lr_pos,
    LR_pos_low95   = exp(log(lr_pos) - 1.96 * se_log_lr_pos),
    LR_pos_high95  = exp(log(lr_pos) + 1.96 * se_log_lr_pos),
    LR_neg         = lr_neg,
    LR_neg_low95   = exp(log(lr_neg) - 1.96 * se_log_lr_neg),
    LR_neg_high95  = exp(log(lr_neg) + 1.96 * se_log_lr_neg)
  )
}
```

```{r}
run_lr <- function(df, data_set_label, path_source, long_or_unique) {
  imap_dfr(gene_groups, function(genes, group_name) {
    df_group <- df %>% filter(gene %in% genes)
    map2_dfr(rule_tbl$rule, rule_tbl$test_pos_expr, function(r, expr) {
      calc_lr(df_group, r, expr) %>% mutate(gene_group = group_name, .before = 1)
    })
  }) %>%
    mutate(data_set = data_set_label, path_source = path_source,
           long_or_unique = long_or_unique)
}

lr_results <- bind_rows(
  run_lr(Clinvar_unique_vs_biobank,   "clinvar_unique_public",   "clinvar", "unique"),
  run_lr(Clinvar_long_vs_biobank,     "clinvar_long_public",     "clinvar", "long"),
  run_lr(DNMs_31k_vs_biobank,         "DNMs_31k_long_public",    "DNM_31k", "long"),
  run_lr(DNMs_31k_unique_vs_biobank,  "DNMs_31k_unique_public",  "DNM_31k", "unique")
) %>%
  mutate(feature = case_when(
    str_detect(rule, "n_COSMIC ")           ~ "n_COSMIC",
    str_detect(rule, "n_sperm_codon_count") ~ "n_sperm_codon_count",
    str_detect(rule, "n_sperm ")            ~ "n_sperm",
    str_detect(rule, "n_COSMIC_codon_count")~ "n_COSMIC_codon_count",
    str_detect(rule, "hotcodon_cancer")     ~ "hotcodon_cancer"
  )) %>%
  mutate(threshold = case_when(
    str_detect(rule, "n_COSMIC > ")              ~ paste0(">", str_extract(rule, "(?<=n_COSMIC > )\\d+")),
    str_detect(rule, "n_COSMIC 1-4")             ~ "1-4",
    str_detect(rule, "n_sperm > ")               ~ paste0(">", str_extract(rule, "(?<=n_sperm > )\\d+")),
    str_detect(rule, "n_sperm_codon_count > ")   ~ paste0(">", str_extract(rule, "(?<=n_sperm_codon_count > )\\d+")),
    str_detect(rule, "n_COSMIC_codon_count > ")  ~ paste0(">", str_extract(rule, "(?<=n_COSMIC_codon_count > )\\d+")),
    str_detect(rule, "n_COSMIC_codon_count 1-4") ~ "1-4",
    str_detect(rule, "hotcodon_cancer")          ~ "TRUE"
  ))
```

## Export LR results

```{r}
write_csv(lr_results, "../../Data/LR_calculations/LR_table.csv")
```
