---
title: "VCEP validation: auto-ACMG vs RASopathy VCEP expert classifications"
output:
  html_document:
    df_print: paged
    code_folding: hide
date: "2026-05-11"
---

# Aim

Run the 404 RASopathy VCEP-classified variants through the auto-ACMG pipeline
(baseline and somatic-weighted) and compare the criteria applied by the pipeline
against those applied by the VCEP.

Expected discrepancy categories:
- **Codes only VCEP can apply**: PS3 (functional), PS4 (case count), PM5 (same codon
  different AA), PM6 (assumed de novo), PP1 (segregation), BS2, BS4, BP5, BP7
- **Strength discrepancies (VCEP stronger)**: PS2 (I apply moderate; VCEP uses
  Very Strong / Strong for multiple confirmed de novos)
- **PM2 weight**: VCEP uses PM2_Supporting; my pipeline applies PM2 at Moderate weight
- **PP2**: I restrict to BRAF/MAP2K1/PTPN11; check VCEP application
- **Codes I should apply but miss**: investigate any systematic gap in BA1/BS1/PP3/BP4/PM2

```{r setup, message=FALSE}
library(tidyverse)

RASopathies <- read_csv("../../Data/Gene_definitions/RASopathy_genes_VCEP_030425.csv",
                        show_col_types = FALSE) %>% pull(gene)

pp2_genes <- c("BRAF", "MAP2K1", "PTPN11")

hotspots_path <- "../../Data/RASopathy_VCEP/rasopathy_hotspots.tsv"
input_path    <- "../../Data/auto_ACMG_output/VCEP_variants_acmg_input.tsv"
output_base   <- "../../Data/auto_ACMG_output/VCEP_variants_ACMG_baseline.tsv"
output_somatic <- "../../Data/auto_ACMG_output/VCEP_variants_ACMG_somatic.tsv"
```

# 1. Prepare auto-ACMG input for VCEP variants

```{r prepare-input, message=FALSE}
vcep_raw <- read_csv(
  "../../Data/RASopathy_VCEP/RASopathy_VCEP_classified_variants_110526.csv",
  show_col_types = FALSE
) %>%
  rename(clinvar_id = `ClinVar Id`, met_codes = `Met Codes`,
         vcep_class = Classification)

# --- Genomic coordinates from ClinVar VariationID ---
clinvar_coords <- read_tsv(
  "../../Data/Clinvar_060225/Clinvar_build_38_DDG2P.tsv.gz",
  show_col_types = FALSE
) %>%
  select(VariationID, Chromosome, PositionVCF, ReferenceAlleleVCF,
         AlternateAlleleVCF, Origin, GeneSymbol) %>%
  mutate(
    chrom             = paste0("chr", Chromosome),
    de_novo_confirmed = grepl("de novo", Origin, ignore.case = TRUE)
  ) %>%
  select(-Chromosome, -Origin)

vcep_coords <- vcep_raw %>%
  left_join(clinvar_coords, by = c("clinvar_id" = "VariationID"))

cat("Matched to genomic coords:", sum(!is.na(vcep_coords$chrom)), "/", nrow(vcep_coords), "\n")
cat("Unmatched ClinVar IDs:", vcep_coords$clinvar_id[is.na(vcep_coords$chrom)], "\n")

vcep_matched <- vcep_coords %>%
  filter(!is.na(chrom), GeneSymbol %in% RASopathies) %>%
  mutate(
    pos = as.integer(PositionVCF),
    ref = ReferenceAlleleVCF,
    alt = AlternateAlleleVCF
  )

cat("Variants in 12 RASopathy genes:", nrow(vcep_matched), "\n")
cat("\nGene distribution:\n")
print(count(vcep_matched, GeneSymbol, vcep_class))
```

```{r join-annotations, message=FALSE, warning=FALSE}
# --- Join REVEL, somatic counts from all three annotated variant files ---
read_annot <- function(path) {
  read_tsv(path, show_col_types = FALSE) %>%
    filter(gene %in% RASopathies) %>%
    rename(n_sperm_codon_count = n_sperm_twinsUK_codon_count) %>%
    select(chrom, pos, ref, alt, gene, transcript, consequence,
           codon_change, codon,
           revel = REVEL_score,
           n_COSMIC_codon_count, n_sperm_codon_count)
}

annot_all <- bind_rows(
  read_annot("../../Data/Variant_lists/Clinvar_DDG2P_missense_path_annot_240426_public.tsv.gz"),
  read_annot("../../Data/Variant_lists/Clinvar_DDG2P_missense_benign_annot_240426_public.tsv.gz"),
  read_annot("../../Data/Variant_lists/Clinvar_DDG2P_missense_VUS_annot_240426_public.tsv.gz")
) %>%
  mutate(revel = suppressWarnings(as.numeric(revel))) %>%
  distinct(chrom, pos, ref, alt, .keep_all = TRUE)

# --- gnomAD popmax AF from all three gnomAD/CADD files ---
read_gnomad <- function(fname) {
  read_tsv(
    file.path("../../Data/Variant_lists/gnomad_CADD", fname),
    col_names = c("chrom", "pos", "ref", "alt", "gnomad_popmax_af", "cadd_phred"),
    show_col_types = FALSE
  ) %>%
    mutate(
      gnomad_popmax_af = suppressWarnings(
        as.numeric(if_else(gnomad_popmax_af == ".", "0", gnomad_popmax_af))
      ),
      chrom = paste0("chr", chrom)
    )
}

gnomad_all <- bind_rows(
  read_gnomad("Clinvar_DDG2P_missense_path_gnomad_popmax_CADD_Phred.tsv"),
  read_gnomad("Clinvar_DDG2P_missense_benign_gnomad_popmax_CADD_Phred.tsv"),
  read_gnomad("Clinvar_DDG2P_missense_VUS_gnomad_popmax_CADD_Phred.tsv")
) %>%
  distinct(chrom, pos, ref, alt, .keep_all = TRUE)

# --- Join everything ---
vcep_input <- vcep_matched %>%
  left_join(annot_all,  by = c("chrom", "pos", "ref", "alt")) %>%
  left_join(gnomad_all, by = c("chrom", "pos", "ref", "alt")) %>%
  # somatic evidence codes (same logic as Prep_data_for_Auto_ACMG.Rmd)
  mutate(
    pathogenic_somatic_supporting = FALSE,
    benign_somatic_supporting     = n_COSMIC_codon_count == 0,
    pathogenic_somatic_moderate   = n_COSMIC_codon_count > 2,
    pathogenic_somatic_strong     = n_COSMIC_codon_count > 10,
    # PS1 set FALSE to avoid circularity with VCEP's own submission
    clinvar_exact_path_star2plus  = FALSE,
    gnomad_min_cov                = NA_character_,
    gene                          = GeneSymbol
  ) %>%
  # fill missing annotation columns required by ACMG_auto.R
  mutate(
    consequence = coalesce(consequence, "missense_variant"),
    codon       = coalesce(as.character(codon), NA_character_)
  )

cat("Annotation coverage (all matched variants):\n")
cat("  REVEL score present:          ", sum(!is.na(vcep_input$revel)),
    "/", nrow(vcep_input), "\n")
cat("  gnomAD AF present:            ", sum(!is.na(vcep_input$gnomad_popmax_af)),
    "/", nrow(vcep_input), "\n")
cat("  COSMIC codon count present:   ", sum(!is.na(vcep_input$n_COSMIC_codon_count)),
    "/", nrow(vcep_input), "\n")
cat("  de_novo_confirmed TRUE:       ", sum(vcep_input$de_novo_confirmed, na.rm = TRUE), "\n\n")

# Restrict to variants with full annotation (gnomAD AF + REVEL + COSMIC present).
# The 140 missing-annotation variants are almost all benign/likely benign variants
# classified by the VCEP via BA1/BS1 population frequency — they are not in the
# ClinVar annotated files because they were not previously classified in ClinVar.
# Running them through the pipeline without gnomAD AF causes spurious PM2 firing
# and missed BA1/BS1, making the comparison misleading.
vcep_input_annotated <- vcep_input %>% filter(!is.na(gnomad_popmax_af))

cat("Restricting to fully-annotated variants:", nrow(vcep_input_annotated), "\n")
cat("Excluded (missing annotation):          ", nrow(vcep_input) - nrow(vcep_input_annotated), "\n")
cat("Excluded by VCEP class:\n")
print(vcep_input %>% filter(is.na(gnomad_popmax_af)) %>% count(vcep_class))

write_tsv(vcep_input_annotated, input_path)

# Use only annotated variants for all downstream comparisons
vcep_matched <- vcep_input_annotated
```

# 2. Run auto-ACMG (baseline and somatic)

```{r run-acmg}
system(paste("Rscript ../../Scripts/ACMG_and_LRs/ACMG_auto.R",
             input_path, hotspots_path, output_base))

system(paste("Rscript ../../Scripts/ACMG_and_LRs/ACMG_auto_somatic.R",
             input_path, hotspots_path, output_somatic))

acmg_base   <- read_tsv(output_base,   show_col_types = FALSE)
acmg_somatic <- read_tsv(output_somatic, show_col_types = FALSE)

cat("Baseline output rows:       ", nrow(acmg_base), "\n")
cat("Somatic-weighted output rows:", nrow(acmg_somatic), "\n")
```

# 3. Parse criteria

```{r parse-criteria}
# Parse VCEP Met Codes into tidy long form
# Each code may have a strength suffix, e.g. "PS2_Very Strong", "PM2_Supporting", "PP3"
parse_vcep_codes <- function(codes_str) {
  str_split(codes_str, ",\\s*")[[1]] %>%
    str_trim() %>%
    tibble(raw = .) %>%
    mutate(
      # Separate base code from strength descriptor
      code     = str_extract(raw, "^[A-Z0-9]+"),
      strength = str_remove(raw, "^[A-Z0-9]+_?") %>% str_trim() %>% na_if("")
    ) %>%
    filter(!is.na(code))
}

vcep_long <- vcep_matched %>%
  select(chrom, pos, ref, alt, vcep_class, met_codes) %>%
  mutate(rn = row_number()) %>%
  rowwise() %>%
  mutate(parsed = list(parse_vcep_codes(met_codes))) %>%
  ungroup() %>%
  unnest(parsed)

# Parse auto-ACMG criteria_met (semicolon-separated)
acmg_long_base <- acmg_base %>%
  select(chrom, pos, ref, alt, acmg_class, criteria_met) %>%
  separate_rows(criteria_met, sep = ";") %>%
  rename(code = criteria_met) %>%
  filter(code != "None", !is.na(code))

acmg_long_somatic <- acmg_somatic %>%
  select(chrom, pos, ref, alt, acmg_class, criteria_met) %>%
  separate_rows(criteria_met, sep = ";") %>%
  rename(code = criteria_met) %>%
  filter(code != "None", !is.na(code))
```

# 4. Overall classification agreement

```{r classification-agreement}
# Standardise class labels
std_class <- function(x) {
  case_when(
    x %in% c("Pathogenic", "Likely Pathogenic")           ~ "LP/P",
    x %in% c("Benign", "Likely Benign")                   ~ "LB/B",
    x %in% c("Uncertain Significance", "VUS")             ~ "VUS",
    x == "Likely pathogenic"                               ~ "LP/P",
    x == "Likely benign"                                   ~ "Likely benign",
    TRUE ~ x
  )
}

compare_class <- vcep_matched %>%
  select(chrom, pos, ref, alt, vcep_class) %>%
  left_join(acmg_base   %>% select(chrom, pos, ref, alt, auto_class_base   = acmg_class),
            by = c("chrom","pos","ref","alt")) %>%
  left_join(acmg_somatic %>% select(chrom, pos, ref, alt, auto_class_somatic = acmg_class),
            by = c("chrom","pos","ref","alt")) %>%
  mutate(
    vcep_std   = std_class(vcep_class),
    base_std   = std_class(auto_class_base),
    somatic_std = std_class(auto_class_somatic),
    agree_base   = vcep_std == base_std,
    agree_somatic = vcep_std == somatic_std
  )

cat("Classification agreement — baseline auto-ACMG vs VCEP:\n")
print(compare_class %>% count(vcep_std, base_std) %>% arrange(vcep_std))

cat("\nOverall agreement (exact match):", round(100 * mean(compare_class$agree_base), 1), "%\n")
cat("Agreement after somatic weighting:", round(100 * mean(compare_class$agree_somatic), 1), "%\n")
```

# 5. Code-by-code discrepancy analysis {.tabset}

```{r code-summary}
# For each ACMG code, count:
#   both_apply   — VCEP fires it AND I fire it
#   vcep_only    — VCEP fires it, I don't
#   mine_only    — I fire it, VCEP doesn't

# Use a per-variant presence indicator
var_key <- c("chrom","pos","ref","alt")

vcep_codes_wide <- vcep_long %>%
  distinct(chrom, pos, ref, alt, code) %>%
  mutate(vcep = TRUE)

mine_codes_wide <- acmg_long_base %>%
  distinct(chrom, pos, ref, alt, code) %>%
  mutate(mine = TRUE)

code_compare <- full_join(vcep_codes_wide, mine_codes_wide,
                          by = c(var_key, "code")) %>%
  mutate(vcep = coalesce(vcep, FALSE),
         mine = coalesce(mine, FALSE))

code_summary <- code_compare %>%
  group_by(code) %>%
  summarise(
    vcep_total  = sum(vcep),
    mine_total  = sum(mine),
    both_apply  = sum(vcep & mine),
    vcep_only   = sum(vcep & !mine),
    mine_only   = sum(!vcep & mine),
    .groups = "drop"
  ) %>%
  arrange(desc(vcep_total))

cat("Code-level summary (baseline auto-ACMG vs VCEP):\n")
print(code_summary, n = 50)
```

## Codes I apply but VCEP doesn't (over-application)

```{r mine-only}
mine_only_variants <- code_compare %>%
  filter(!vcep, mine) %>%
  left_join(vcep_matched %>% select(chrom, pos, ref, alt, vcep_class, met_codes),
            by = var_key) %>%
  left_join(acmg_base %>% select(chrom, pos, ref, alt, auto_class = acmg_class,
                                  criteria_met),
            by = var_key)

cat("Variants where I fire a code the VCEP doesn't, by code:\n")
mine_only_variants %>% count(code, sort = TRUE) %>% print()

cat("\nDetails for each over-applied code:\n")
for (cd in mine_only_variants %>% count(code) %>% arrange(desc(n)) %>% pull(code)) {
  cat("\n---", cd, "---\n")
  mine_only_variants %>%
    filter(code == cd) %>%
    select(gene = chrom, pos, ref, alt, vcep_class, met_codes, auto_class, criteria_met) %>%
    print()
}
```

## Codes VCEP applies but I don't (missed or inapplicable)

```{r vcep-only}
vcep_only_summary <- code_compare %>%
  filter(vcep, !mine) %>%
  count(code, sort = TRUE)

cat("Codes VCEP applies that auto-ACMG does not, by frequency:\n")
print(vcep_only_summary)
```

## Agreement on shared codes (both apply)

```{r both-apply}
cat("Codes where both VCEP and auto-ACMG fire:\n")
code_summary %>%
  filter(both_apply > 0) %>%
  select(code, vcep_total, mine_total, both_apply, vcep_only, mine_only) %>%
  print()
```

## Strength discrepancies for PS2

```{r ps2-strength}
# PS2: check VCEP strength vs my application (always moderate/PM)
ps2_vcep <- vcep_long %>%
  filter(code == "PS2") %>%
  select(chrom, pos, ref, alt, vcep_class, vcep_strength = strength, raw_code = raw)

ps2_mine <- acmg_long_base %>%
  filter(code == "PS2") %>%
  select(chrom, pos, ref, alt) %>%
  mutate(mine_fires = TRUE)

ps2_compare <- ps2_vcep %>%
  left_join(ps2_mine, by = var_key) %>%
  mutate(mine_fires = coalesce(mine_fires, FALSE))

cat("VCEP PS2 strength distribution:\n")
print(count(ps2_compare, vcep_strength, mine_fires))

cat("\nVariants where VCEP applies PS2 but auto-ACMG doesn't:\n")
ps2_compare %>%
  filter(!mine_fires) %>%
  left_join(vcep_matched %>% select(chrom, pos, ref, alt, met_codes), by = var_key) %>%
  select(chrom, pos, ref, alt, vcep_class, raw_code, met_codes) %>%
  print()
```

## PM2 weight discrepancy

```{r pm2-weight}
# VCEP applies PM2 at Supporting strength; auto-ACMG applies at Moderate weight
pm2_vcep <- vcep_long %>%
  filter(code == "PM2") %>%
  count(strength, name = "n_vcep")

cat("VCEP PM2 strength distribution:\n")
print(pm2_vcep)

cat("\nNote: auto-ACMG applies PM2 at Moderate weight (LR=4.33).\n")
cat("VCEP uses PM2 at Supporting strength (LR=2.08).\n")
cat("This means my pipeline over-weights PM2 by a factor of 4.33/2.08 = 2.1x.\n")

pm2_agreement <- code_compare %>%
  filter(code == "PM2") %>%
  summarise(vcep_total = sum(vcep), mine_total = sum(mine),
            both = sum(vcep & mine), vcep_only = sum(vcep & !mine),
            mine_only = sum(!vcep & mine))
cat("\nPM2 agreement:\n")
print(pm2_agreement)
```

## PP2 application

```{r pp2}
pp2_vcep <- vcep_long %>%
  filter(code == "PP2") %>%
  left_join(vcep_matched %>% select(chrom, pos, ref, alt, gene = GeneSymbol),
            by = var_key) %>%
  count(gene, name = "vcep_n")

pp2_mine <- acmg_long_base %>%
  filter(code == "PP2") %>%
  left_join(acmg_base %>% select(chrom, pos, ref, alt, gene),
            by = var_key) %>%
  count(gene, name = "mine_n")

cat("PP2 application by gene (VCEP vs auto-ACMG):\n")
full_join(pp2_vcep, pp2_mine, by = "gene") %>%
  mutate(pp2_applicable_per_vcep = gene %in% pp2_genes) %>%
  arrange(gene) %>%
  print()
```

## BA1 / BS1 agreement

```{r ba1-bs1}
cat("BA1 agreement:\n")
code_compare %>% filter(code == "BA1") %>%
  summarise(vcep=sum(vcep), mine=sum(mine), both=sum(vcep&mine),
            vcep_only=sum(vcep&!mine), mine_only=sum(!vcep&mine)) %>% print()

cat("\nBS1 agreement:\n")
code_compare %>% filter(code == "BS1") %>%
  summarise(vcep=sum(vcep), mine=sum(mine), both=sum(vcep&mine),
            vcep_only=sum(vcep&!mine), mine_only=sum(!vcep&mine)) %>% print()
```

## PP3 / BP4 agreement

```{r pp3-bp4}
cat("PP3 agreement:\n")
code_compare %>% filter(code == "PP3") %>%
  summarise(vcep=sum(vcep), mine=sum(mine), both=sum(vcep&mine),
            vcep_only=sum(vcep&!mine), mine_only=sum(!vcep&mine)) %>% print()

cat("\nBP4 agreement:\n")
code_compare %>% filter(code == "BP4") %>%
  summarise(vcep=sum(vcep), mine=sum(mine), both=sum(vcep&mine),
            vcep_only=sum(vcep&!mine), mine_only=sum(!vcep&mine)) %>% print()

# Variants where PP3/BP4 disagree - check REVEL values
pp3_disagree <- code_compare %>%
  filter(code == "PP3", vcep != mine) %>%
  left_join(acmg_base %>% select(chrom, pos, ref, alt, revel, acmg_class), by = var_key) %>%
  left_join(vcep_matched %>% select(chrom, pos, ref, alt, vcep_class, met_codes), by = var_key)

if (nrow(pp3_disagree) > 0) {
  cat("\nPP3 disagreements with REVEL values:\n")
  print(pp3_disagree %>% select(chrom, pos, ref, alt, vcep_class, met_codes, revel, acmg_class))
}
```

# 6. Summary of discrepancy types

## Code glossary

Each ACMG/AMP evidence code and its meaning in the RASopathy VCEP context:

| Code | Direction | Definition |
|------|-----------|------------|
| **PVS1** | Pathogenic Very Strong | Null variant (nonsense, frameshift, splice ±1/2, start-loss) in a gene where loss of function is the disease mechanism |
| **PS1** | Pathogenic Strong | Same amino acid change as a previously established pathogenic variant |
| **PS2** | Pathogenic Strong* | Confirmed *de novo* (paternity and maternity verified) in a patient with the disease and no family history. *Applied at Moderate or Very Strong by RASopathy VCEP depending on number of de novos* |
| **PS3** | Pathogenic Strong* | Well-established functional studies show a damaging effect (e.g., RAS activation assay, phosphatase assay). *Strength varied by VCEP* |
| **PS4** | Pathogenic Strong* | Variant prevalence significantly higher in affected individuals than in controls. *Strength varied by VCEP* |
| **PM1** | Pathogenic Moderate | Located in a mutational hotspot or well-established functional domain with no or few benign variants |
| **PM2** | Pathogenic Supporting* | Absent or at extremely low frequency in population databases (gnomAD). *RASopathy VCEP applies at Supporting strength, not default Moderate* |
| **PM5** | Pathogenic Moderate | Novel missense at a codon where a different pathogenic missense change has been established |
| **PM6** | Pathogenic Moderate* | Assumed *de novo* (maternity/paternity not confirmed). *Strength varied by VCEP; used in place of PS2 when de novo not formally confirmed* |
| **PP1** | Pathogenic Supporting | Cosegregation with disease in multiple affected family members |
| **PP2** | Pathogenic Supporting | Missense variant in a gene with a low rate of benign missense variation (gnomAD missense z score > 3.09). *Applicable only in BRAF, MAP2K1, PTPN11 per VCEP spec* |
| **PP3** | Pathogenic Supporting | Computational evidence supports pathogenic effect: REVEL score ≥ 0.70 |
| **BA1** | Benign Standalone | Allele frequency > 5% in gnomAD (variant too common to be a dominant disease-causing variant) |
| **BS1** | Benign Strong | Allele frequency greater than expected given disease prevalence (> 0.025% in gnomAD for RASopathies per VCEP) |
| **BS2** | Benign Strong | Variant observed in a healthy adult with a condition of full penetrance |
| **BS3** | Benign Strong | Functional studies show no damaging effect |
| **BS4** | Benign Strong | Lack of segregation in affected family members |
| **BP2** | Benign Supporting | Observed in trans with a pathogenic variant for a dominant disorder |
| **BP4** | Benign Supporting | Computational evidence supports benign effect: REVEL score ≤ 0.30 |
| **BP5** | Benign Supporting | Variant found in a case with an alternative molecular explanation for the phenotype |
| **BP7** | Benign Supporting | Synonymous variant with no predicted effect on splicing |

\* Strength modified from ACMG default by RASopathy VCEP.

---

## Discrepancy table

```{r summary-table}
# Annotate each code with its definition and discrepancy category
code_definitions <- tribble(
  ~code,  ~direction,  ~definition,
  "PVS1", "P Very Strong",    "Null variant (nonsense/frameshift/splice) in LOF gene",
  "PS1",  "P Strong",         "Same amino acid change as established pathogenic variant",
  "PS2",  "P Strong*",        "Confirmed de novo (paternity/maternity verified)",
  "PS3",  "P Strong*",        "Functional studies show damaging effect",
  "PS4",  "P Strong*",        "Prevalence significantly higher in affected vs controls",
  "PM1",  "P Moderate",       "Located in mutational hotspot or critical functional domain",
  "PM2",  "P Supporting*",    "Absent/very rare in gnomAD population databases",
  "PM5",  "P Moderate",       "Novel missense at codon with different known pathogenic missense",
  "PM6",  "P Moderate*",      "Assumed de novo (paternity/maternity not confirmed)",
  "PP1",  "P Supporting",     "Cosegregation with disease in affected family members",
  "PP2",  "P Supporting",     "Missense in gene with low benign missense rate (z > 3.09)",
  "PP3",  "P Supporting",     "Computational support for pathogenicity (REVEL >= 0.70)",
  "BA1",  "B Standalone",     "Allele frequency > 5% in gnomAD",
  "BS1",  "B Strong",         "Allele frequency > expected for disease (> 0.025% gnomAD)",
  "BS2",  "B Strong",         "Observed in healthy adult with full-penetrance condition",
  "BS3",  "B Strong",         "Functional studies show no damaging effect",
  "BS4",  "B Strong",         "Lack of segregation in affected family members",
  "BP2",  "B Supporting",     "Observed in trans with pathogenic variant (dominant disorder)",
  "BP4",  "B Supporting",     "Computational support for benign effect (REVEL <= 0.30)",
  "BP5",  "B Supporting",     "Variant found with alternative molecular explanation",
  "BP7",  "B Supporting",     "Synonymous variant, no predicted splicing effect"
)

discrepancy_categories <- tribble(
  ~code, ~category, ~explanation,
  "PS1",  "VCEP-only: circularity",
          "Set to FALSE in this analysis to avoid circularity — the VCEP submission itself is the 3-star ClinVar record",
  "PS2",  "Partial agreement; strength discrepancy",
          "I apply PS2 at Moderate (PM) strength using ClinVar Origin='de novo'. VCEP applies at Very Strong for multiple confirmed de novos. 25 VCEP-only: ClinVar Origin field does not flag as de novo even though VCEP treats as such. 26 mine-only: ClinVar says de novo but VCEP uses PM6 instead",
  "PS3",  "VCEP-only: requires functional data",
          "Requires access to published functional assay results; cannot be automated",
  "PS4",  "VCEP-only: requires case count data",
          "Requires access to patient databases to count affected individuals; cannot be automated",
  "PM1",  "Partial agreement; hotspot list incomplete",
          "Both apply; VCEP fires for 24 additional variants not in my hotspot list — hotspot list may need updating",
  "PM2",  "Strength discrepancy",
          "Both apply well (107/120 VCEP instances matched). VCEP applies at Supporting strength (LR=2.08); my pipeline applies at Moderate (LR=4.33), over-weighting PM2 by ~2x",
  "PM5",  "VCEP-only: requires variant database lookup",
          "Requires checking whether a different pathogenic missense at the same codon exists; cannot be automated without a curated variant database",
  "PM6",  "VCEP-only: requires clinical de novo assessment",
          "Used when de novo status is assumed but paternity/maternity not formally confirmed; requires clinical information not in ClinVar",
  "PP1",  "VCEP-only: requires family data",
          "Cosegregation data requires access to family history records; cannot be automated",
  "PP2",  "Partial agreement; gene restriction differs",
          "I restrict to BRAF/MAP2K1/PTPN11 (gnomAD z > 3.09 per VCEP spec). VCEP also applies PP2 in HRAS, KRAS, MAP2K2, RAF1, SHOC2, SOS1 despite their gene-level VCEP pages stating 'not applicable' — inconsistency in VCEP documentation vs practice",
  "PP3",  "Good agreement",
          "102/105 VCEP applications matched. 3 minor discrepancies likely due to borderline REVEL values or missing REVEL data",
  "BA1",  "Good agreement",
          "26/29 VCEP applications matched. 8 mine-only: these variants have gnomAD AF > 0.05% in my data but VCEP does not apply BA1, possibly a gnomAD version difference or AF close to threshold",
  "BS1",  "Partial agreement; gnomAD version likely differs",
          "5/21 VCEP applications matched. 16 VCEP-only: likely gnomAD version or AF threshold differences. 3 mine-only",
  "BS2",  "VCEP-only: requires clinical data",
          "Requires knowledge that the variant has been observed in a healthy adult; cannot be automated",
  "BS3",  "VCEP-only: requires functional data",
          "Requires access to published functional assay results",
  "BS4",  "VCEP-only: requires family data",
          "Requires family segregation information",
  "BP4",  "I apply more than VCEP",
          "I fire BP4 for 33 variants the VCEP doesn't. Of these, 28 have BA1 or BS1 in VCEP codes — VCEP omits BP4 as redundant when a standalone benign code already applies. 5 are VUS: VCEP appears not to apply BP4 despite REVEL clearly <= 0.30 — likely inconsistent VCEP application",
  "BP5",  "VCEP-only: requires clinical data",
          "Requires knowledge of an alternative molecular diagnosis in the proband; cannot be automated",
  "BP7",  "VCEP-only: outside pipeline scope",
          "Applied to synonymous variants; my pipeline is restricted to missense variants"
)

summary_table <- code_definitions %>%
  left_join(code_summary %>% select(code, vcep_total, mine_total, both_apply,
                                     vcep_only, mine_only),
            by = "code") %>%
  left_join(discrepancy_categories, by = "code") %>%
  mutate(across(c(vcep_total, mine_total, both_apply, vcep_only, mine_only),
                ~ replace_na(., 0L))) %>%
  arrange(direction, code)

summary_table %>%
  select(Code = code, Direction = direction, Definition = definition,
         `VCEP n` = vcep_total, `Mine n` = mine_total,
         Both = both_apply, `VCEP only` = vcep_only, `Mine only` = mine_only,
         `Discrepancy explanation` = explanation) %>%
  knitr::kable()
```

## Discrepancy categories at a glance

```{r summary-categories}
cat("=== GOOD AGREEMENT (both apply, mostly concordant) ===\n")
summary_table %>%
  filter(both_apply >= 20, vcep_only <= 5, mine_only <= 5) %>%
  select(code, definition, vcep_total, mine_total, both_apply) %>%
  print()

cat("\n=== STRENGTH DISCREPANCIES (both apply but at different weights) ===\n")
cat("PM2: VCEP applies at Supporting (LR=2.08); pipeline applies at Moderate (LR=4.33)\n")
vcep_long %>% filter(code == "PM2") %>%
  mutate(strength = if_else(is.na(strength), "default (Moderate)", strength)) %>%
  count(strength, name = "n_vcep") %>% print()

cat("\nPS2: VCEP applies at various strengths; pipeline always applies at Moderate\n")
vcep_long %>% filter(code == "PS2") %>%
  mutate(strength = if_else(is.na(strength), "default", strength)) %>%
  count(strength, name = "n_vcep") %>% print()

cat("\n=== I APPLY BUT VCEP DOESN'T ===\n")
cat("(n = number of variants where I fire the code and VCEP does not)\n")
summary_table %>%
  filter(mine_only > 0) %>%
  select(code, definition, mine_only, explanation) %>%
  arrange(desc(mine_only)) %>%
  print(n = 20)

cat("\n=== VCEP APPLIES BUT I CAN'T (requires patient/family/functional data) ===\n")
summary_table %>%
  filter(vcep_only > 5, both_apply == 0) %>%
  select(code, definition, vcep_only, explanation) %>%
  arrange(desc(vcep_only)) %>%
  print(n = 20)

cat("\n=== PARTIAL AGREEMENT (both apply but with notable gaps) ===\n")
summary_table %>%
  filter(both_apply > 0, (vcep_only > 5 | mine_only > 5)) %>%
  select(code, definition, vcep_total, mine_total, both_apply, vcep_only, mine_only, explanation) %>%
  arrange(desc(vcep_total)) %>%
  print(n = 20)
```
