---
title: "VCEP validation: auto-ACMG vs RASopathy VCEP expert classifications"
output:
  html_document:
    df_print: paged
    code_folding: hide
date: "2026-05-11"
---

# Aim

Run the 404 RASopathy VCEP-classified variants through the auto-ACMG pipeline
(baseline and somatic-weighted) and compare the criteria applied by the pipeline
against those applied by the VCEP.

Expected discrepancy categories:
- **Codes only VCEP can apply**: PS3 (functional), PS4 (case count), PM5 (same codon
  different AA), PM6 (assumed de novo), PP1 (segregation), BS2, BS4, BP5, BP7
- **Strength discrepancies (VCEP stronger)**: PS2 (I apply moderate; VCEP uses
  Very Strong / Strong for multiple confirmed de novos)
- **PM2 weight**: VCEP uses PM2_Supporting; my pipeline applies PM2 at Moderate weight
- **PP2**: I restrict to BRAF/MAP2K1/PTPN11; check VCEP application
- **Codes I should apply but miss**: investigate any systematic gap in BA1/BS1/PP3/BP4/PM2

```{r setup, message=FALSE}
library(tidyverse)

RASopathies <- read_csv("../../Data/Gene_definitions/RASopathy_genes_VCEP_030425.csv",
                        show_col_types = FALSE) %>% pull(gene)

pp2_genes <- c("BRAF", "MAP2K1", "PTPN11")

hotspots_path <- "../../Data/RASopathy_VCEP/rasopathy_hotspots.tsv"
input_path    <- "../../Data/auto_ACMG_output/VCEP_variants_acmg_input.tsv"
output_base   <- "../../Data/auto_ACMG_output/VCEP_variants_ACMG_baseline.tsv"
output_somatic <- "../../Data/auto_ACMG_output/VCEP_variants_ACMG_somatic.tsv"
```

# 1. Prepare auto-ACMG input for VCEP variants

```{r prepare-input, message=FALSE}
vcep_raw <- read_csv(
  "../../Data/RASopathy_VCEP/RASopathy_VCEP_classified_variants_110526.csv",
  show_col_types = FALSE
) %>%
  rename(clinvar_id = `ClinVar Id`, met_codes = `Met Codes`,
         vcep_class = Classification)

# --- Genomic coordinates from ClinVar VariationID ---
clinvar_coords <- read_tsv(
  "../../Data/Clinvar_060225/Clinvar_build_38_DDG2P.tsv.gz",
  show_col_types = FALSE
) %>%
  select(VariationID, Chromosome, PositionVCF, ReferenceAlleleVCF,
         AlternateAlleleVCF, Origin, GeneSymbol) %>%
  mutate(
    chrom             = paste0("chr", Chromosome),
    de_novo_confirmed = grepl("de novo", Origin, ignore.case = TRUE)
  ) %>%
  select(-Chromosome, -Origin)

vcep_coords <- vcep_raw %>%
  left_join(clinvar_coords, by = c("clinvar_id" = "VariationID"))

cat("Matched to genomic coords:", sum(!is.na(vcep_coords$chrom)), "/", nrow(vcep_coords), "\n")
cat("Unmatched ClinVar IDs:", vcep_coords$clinvar_id[is.na(vcep_coords$chrom)], "\n")

vcep_matched <- vcep_coords %>%
  filter(!is.na(chrom), GeneSymbol %in% RASopathies) %>%
  mutate(
    pos = as.integer(PositionVCF),
    ref = ReferenceAlleleVCF,
    alt = AlternateAlleleVCF
  )

cat("Variants in 12 RASopathy genes:", nrow(vcep_matched), "\n")
cat("\nGene distribution:\n")
print(count(vcep_matched, GeneSymbol, vcep_class))
```

```{r join-annotations, message=FALSE, warning=FALSE}
# --- Join REVEL, somatic counts from all three annotated variant files ---
read_annot <- function(path) {
  read_tsv(path, show_col_types = FALSE) %>%
    filter(gene %in% RASopathies) %>%
    rename(n_sperm_codon_count = n_sperm_twinsUK_codon_count) %>%
    select(chrom, pos, ref, alt, gene, transcript, consequence,
           codon_change, codon,
           revel = REVEL_score,
           n_COSMIC_codon_count, n_sperm_codon_count)
}

annot_all <- bind_rows(
  read_annot("../../Data/Variant_lists/Clinvar_DDG2P_missense_path_annot_240426_public.tsv.gz"),
  read_annot("../../Data/Variant_lists/Clinvar_DDG2P_missense_benign_annot_240426_public.tsv.gz"),
  read_annot("../../Data/Variant_lists/Clinvar_DDG2P_missense_VUS_annot_240426_public.tsv.gz")
) %>%
  mutate(revel = suppressWarnings(as.numeric(revel))) %>%
  distinct(chrom, pos, ref, alt, .keep_all = TRUE)

# --- gnomAD popmax AF from all three gnomAD/CADD files ---
read_gnomad <- function(fname) {
  read_tsv(
    file.path("../../Data/Variant_lists/gnomad_CADD", fname),
    col_names = c("chrom", "pos", "ref", "alt", "gnomad_popmax_af", "cadd_phred"),
    show_col_types = FALSE
  ) %>%
    mutate(
      gnomad_popmax_af = suppressWarnings(
        as.numeric(if_else(gnomad_popmax_af == ".", "0", gnomad_popmax_af))
      ),
      chrom = paste0("chr", chrom)
    )
}

gnomad_all <- bind_rows(
  read_gnomad("Clinvar_DDG2P_missense_path_gnomad_popmax_CADD_Phred.tsv"),
  read_gnomad("Clinvar_DDG2P_missense_benign_gnomad_popmax_CADD_Phred.tsv"),
  read_gnomad("Clinvar_DDG2P_missense_VUS_gnomad_popmax_CADD_Phred.tsv")
) %>%
  distinct(chrom, pos, ref, alt, .keep_all = TRUE)

# --- Join everything ---
vcep_input <- vcep_matched %>%
  left_join(annot_all,  by = c("chrom", "pos", "ref", "alt")) %>%
  left_join(gnomad_all, by = c("chrom", "pos", "ref", "alt")) %>%
  # somatic evidence codes (same logic as Prep_data_for_Auto_ACMG.Rmd)
  mutate(
    pathogenic_somatic_supporting = FALSE,
    benign_somatic_supporting     = n_COSMIC_codon_count == 0,
    pathogenic_somatic_moderate   = n_COSMIC_codon_count > 2,
    pathogenic_somatic_strong     = n_COSMIC_codon_count > 10,
    # PS1 set FALSE to avoid circularity with VCEP's own submission
    clinvar_exact_path_star2plus  = FALSE,
    gnomad_min_cov                = NA_character_,
    gene                          = GeneSymbol
  ) %>%
  # fill missing annotation columns required by ACMG_auto.R
  mutate(
    consequence = coalesce(consequence, "missense_variant"),
    codon       = coalesce(as.character(codon), NA_character_)
  )

cat("Annotation coverage (all matched variants):\n")
cat("  REVEL score present:          ", sum(!is.na(vcep_input$revel)),
    "/", nrow(vcep_input), "\n")
cat("  gnomAD AF present:            ", sum(!is.na(vcep_input$gnomad_popmax_af)),
    "/", nrow(vcep_input), "\n")
cat("  COSMIC codon count present:   ", sum(!is.na(vcep_input$n_COSMIC_codon_count)),
    "/", nrow(vcep_input), "\n")
cat("  de_novo_confirmed TRUE:       ", sum(vcep_input$de_novo_confirmed, na.rm = TRUE), "\n\n")

# Restrict to variants with full annotation (gnomAD AF + REVEL + COSMIC present).
# The 140 missing-annotation variants are almost all benign/likely benign variants
# classified by the VCEP via BA1/BS1 population frequency — they are not in the
# ClinVar annotated files because they were not previously classified in ClinVar.
# Running them through the pipeline without gnomAD AF causes spurious PM2 firing
# and missed BA1/BS1, making the comparison misleading.
vcep_input_annotated <- vcep_input %>% filter(!is.na(gnomad_popmax_af))

cat("Restricting to fully-annotated variants:", nrow(vcep_input_annotated), "\n")
cat("Excluded (missing annotation):          ", nrow(vcep_input) - nrow(vcep_input_annotated), "\n")
cat("Excluded by VCEP class:\n")
print(vcep_input %>% filter(is.na(gnomad_popmax_af)) %>% count(vcep_class))

write_tsv(vcep_input_annotated, input_path)

# Use only annotated variants for all downstream comparisons
vcep_matched <- vcep_input_annotated
```

# 2. Run auto-ACMG (baseline and somatic)

```{r run-acmg}
system(paste("Rscript ../../Scripts/ACMG_and_LRs/ACMG_auto.R",
             input_path, hotspots_path, output_base))

system(paste("Rscript ../../Scripts/ACMG_and_LRs/ACMG_auto_somatic.R",
             input_path, hotspots_path, output_somatic))

acmg_base   <- read_tsv(output_base,   show_col_types = FALSE)
acmg_somatic <- read_tsv(output_somatic, show_col_types = FALSE)

cat("Baseline output rows:       ", nrow(acmg_base), "\n")
cat("Somatic-weighted output rows:", nrow(acmg_somatic), "\n")
```

# 3. Parse criteria

```{r parse-criteria}
# Parse VCEP Met Codes into tidy long form
# Each code may have a strength suffix, e.g. "PS2_Very Strong", "PM2_Supporting", "PP3"
parse_vcep_codes <- function(codes_str) {
  str_split(codes_str, ",\\s*")[[1]] %>%
    str_trim() %>%
    tibble(raw = .) %>%
    mutate(
      # Separate base code from strength descriptor
      code     = str_extract(raw, "^[A-Z0-9]+"),
      strength = str_remove(raw, "^[A-Z0-9]+_?") %>% str_trim() %>% na_if("")
    ) %>%
    filter(!is.na(code))
}

vcep_long <- vcep_matched %>%
  select(chrom, pos, ref, alt, vcep_class, met_codes) %>%
  mutate(rn = row_number()) %>%
  rowwise() %>%
  mutate(parsed = list(parse_vcep_codes(met_codes))) %>%
  ungroup() %>%
  unnest(parsed)

# Parse auto-ACMG criteria_met (semicolon-separated)
acmg_long_base <- acmg_base %>%
  select(chrom, pos, ref, alt, acmg_class, criteria_met) %>%
  separate_rows(criteria_met, sep = ";") %>%
  rename(code = criteria_met) %>%
  filter(code != "None", !is.na(code))

acmg_long_somatic <- acmg_somatic %>%
  select(chrom, pos, ref, alt, acmg_class, criteria_met) %>%
  separate_rows(criteria_met, sep = ";") %>%
  rename(code = criteria_met) %>%
  filter(code != "None", !is.na(code))
```

# 4. Overall classification agreement

```{r classification-agreement}
# Standardise class labels
std_class <- function(x) {
  case_when(
    x %in% c("Pathogenic", "Likely Pathogenic")           ~ "LP/P",
    x %in% c("Benign", "Likely Benign")                   ~ "LB/B",
    x %in% c("Uncertain Significance", "VUS")             ~ "VUS",
    x == "Likely pathogenic"                               ~ "LP/P",
    x == "Likely benign"                                   ~ "Likely benign",
    TRUE ~ x
  )
}

compare_class <- vcep_matched %>%
  select(chrom, pos, ref, alt, vcep_class) %>%
  left_join(acmg_base   %>% select(chrom, pos, ref, alt, auto_class_base   = acmg_class),
            by = c("chrom","pos","ref","alt")) %>%
  left_join(acmg_somatic %>% select(chrom, pos, ref, alt, auto_class_somatic = acmg_class),
            by = c("chrom","pos","ref","alt")) %>%
  mutate(
    vcep_std   = std_class(vcep_class),
    base_std   = std_class(auto_class_base),
    somatic_std = std_class(auto_class_somatic),
    agree_base   = vcep_std == base_std,
    agree_somatic = vcep_std == somatic_std
  )

cat("Classification agreement — baseline auto-ACMG vs VCEP:\n")
print(compare_class %>% count(vcep_std, base_std) %>% arrange(vcep_std))

cat("\nOverall agreement (exact match):", round(100 * mean(compare_class$agree_base), 1), "%\n")
cat("Agreement after somatic weighting:", round(100 * mean(compare_class$agree_somatic), 1), "%\n")
```

# 5. Code-by-code discrepancy analysis {.tabset}

```{r code-summary}
# For each ACMG code, count:
#   both_apply   — VCEP fires it AND I fire it
#   vcep_only    — VCEP fires it, I don't
#   mine_only    — I fire it, VCEP doesn't

# Use a per-variant presence indicator
var_key <- c("chrom","pos","ref","alt")

vcep_codes_wide <- vcep_long %>%
  distinct(chrom, pos, ref, alt, code) %>%
  mutate(vcep = TRUE)

mine_codes_wide <- acmg_long_base %>%
  distinct(chrom, pos, ref, alt, code) %>%
  mutate(mine = TRUE)

code_compare <- full_join(vcep_codes_wide, mine_codes_wide,
                          by = c(var_key, "code")) %>%
  mutate(vcep = coalesce(vcep, FALSE),
         mine = coalesce(mine, FALSE))

code_summary <- code_compare %>%
  group_by(code) %>%
  summarise(
    vcep_total  = sum(vcep),
    mine_total  = sum(mine),
    both_apply  = sum(vcep & mine),
    vcep_only   = sum(vcep & !mine),
    mine_only   = sum(!vcep & mine),
    .groups = "drop"
  ) %>%
  arrange(desc(vcep_total))

cat("Code-level summary (baseline auto-ACMG vs VCEP):\n")
print(code_summary, n = 50)
```

## Codes I apply but VCEP doesn't (over-application)

```{r mine-only}
mine_only_variants <- code_compare %>%
  filter(!vcep, mine) %>%
  left_join(vcep_matched %>% select(chrom, pos, ref, alt, vcep_class, met_codes),
            by = var_key) %>%
  left_join(acmg_base %>% select(chrom, pos, ref, alt, auto_class = acmg_class,
                                  criteria_met),
            by = var_key)

cat("Variants where I fire a code the VCEP doesn't, by code:\n")
mine_only_variants %>% count(code, sort = TRUE) %>% print()

cat("\nDetails for each over-applied code:\n")
for (cd in mine_only_variants %>% count(code) %>% arrange(desc(n)) %>% pull(code)) {
  cat("\n---", cd, "---\n")
  mine_only_variants %>%
    filter(code == cd) %>%
    select(gene = chrom, pos, ref, alt, vcep_class, met_codes, auto_class, criteria_met) %>%
    print()
}
```

## Codes VCEP applies but I don't (missed or inapplicable)

```{r vcep-only}
vcep_only_summary <- code_compare %>%
  filter(vcep, !mine) %>%
  count(code, sort = TRUE)

cat("Codes VCEP applies that auto-ACMG does not, by frequency:\n")
print(vcep_only_summary)
```

## Agreement on shared codes (both apply)

```{r both-apply}
cat("Codes where both VCEP and auto-ACMG fire:\n")
code_summary %>%
  filter(both_apply > 0) %>%
  select(code, vcep_total, mine_total, both_apply, vcep_only, mine_only) %>%
  print()
```

## Strength discrepancies for PS2

```{r ps2-strength}
# PS2: check VCEP strength vs my application (always moderate/PM)
ps2_vcep <- vcep_long %>%
  filter(code == "PS2") %>%
  select(chrom, pos, ref, alt, vcep_class, vcep_strength = strength, raw_code = raw)

ps2_mine <- acmg_long_base %>%
  filter(code == "PS2") %>%
  select(chrom, pos, ref, alt) %>%
  mutate(mine_fires = TRUE)

ps2_compare <- ps2_vcep %>%
  left_join(ps2_mine, by = var_key) %>%
  mutate(mine_fires = coalesce(mine_fires, FALSE))

cat("VCEP PS2 strength distribution:\n")
print(count(ps2_compare, vcep_strength, mine_fires))

cat("\nVariants where VCEP applies PS2 but auto-ACMG doesn't:\n")
ps2_compare %>%
  filter(!mine_fires) %>%
  left_join(vcep_matched %>% select(chrom, pos, ref, alt, met_codes), by = var_key) %>%
  select(chrom, pos, ref, alt, vcep_class, raw_code, met_codes) %>%
  print()
```

## PM2 weight discrepancy

```{r pm2-weight}
# VCEP applies PM2 at Supporting strength; auto-ACMG applies at Moderate weight
pm2_vcep <- vcep_long %>%
  filter(code == "PM2") %>%
  count(strength, name = "n_vcep")

cat("VCEP PM2 strength distribution:\n")
print(pm2_vcep)

cat("\nNote: auto-ACMG applies PM2 at Moderate weight (LR=4.33).\n")
cat("VCEP uses PM2 at Supporting strength (LR=2.08).\n")
cat("This means my pipeline over-weights PM2 by a factor of 4.33/2.08 = 2.1x.\n")

pm2_agreement <- code_compare %>%
  filter(code == "PM2") %>%
  summarise(vcep_total = sum(vcep), mine_total = sum(mine),
            both = sum(vcep & mine), vcep_only = sum(vcep & !mine),
            mine_only = sum(!vcep & mine))
cat("\nPM2 agreement:\n")
print(pm2_agreement)
```

## PP2 application

```{r pp2}
pp2_vcep <- vcep_long %>%
  filter(code == "PP2") %>%
  left_join(vcep_matched %>% select(chrom, pos, ref, alt, gene = GeneSymbol),
            by = var_key) %>%
  count(gene, name = "vcep_n")

pp2_mine <- acmg_long_base %>%
  filter(code == "PP2") %>%
  left_join(acmg_base %>% select(chrom, pos, ref, alt, gene),
            by = var_key) %>%
  count(gene, name = "mine_n")

cat("PP2 application by gene (VCEP vs auto-ACMG):\n")
full_join(pp2_vcep, pp2_mine, by = "gene") %>%
  mutate(pp2_applicable_per_vcep = gene %in% pp2_genes) %>%
  arrange(gene) %>%
  print()
```

## BA1 / BS1 agreement

```{r ba1-bs1}
cat("BA1 agreement:\n")
code_compare %>% filter(code == "BA1") %>%
  summarise(vcep=sum(vcep), mine=sum(mine), both=sum(vcep&mine),
            vcep_only=sum(vcep&!mine), mine_only=sum(!vcep&mine)) %>% print()

cat("\nBS1 agreement:\n")
code_compare %>% filter(code == "BS1") %>%
  summarise(vcep=sum(vcep), mine=sum(mine), both=sum(vcep&mine),
            vcep_only=sum(vcep&!mine), mine_only=sum(!vcep&mine)) %>% print()
```

## PP3 / BP4 agreement

```{r pp3-bp4}
cat("PP3 agreement:\n")
code_compare %>% filter(code == "PP3") %>%
  summarise(vcep=sum(vcep), mine=sum(mine), both=sum(vcep&mine),
            vcep_only=sum(vcep&!mine), mine_only=sum(!vcep&mine)) %>% print()

cat("\nBP4 agreement:\n")
code_compare %>% filter(code == "BP4") %>%
  summarise(vcep=sum(vcep), mine=sum(mine), both=sum(vcep&mine),
            vcep_only=sum(vcep&!mine), mine_only=sum(!vcep&mine)) %>% print()

# Variants where PP3/BP4 disagree - check REVEL values
pp3_disagree <- code_compare %>%
  filter(code == "PP3", vcep != mine) %>%
  left_join(acmg_base %>% select(chrom, pos, ref, alt, revel, acmg_class), by = var_key) %>%
  left_join(vcep_matched %>% select(chrom, pos, ref, alt, vcep_class, met_codes), by = var_key)

if (nrow(pp3_disagree) > 0) {
  cat("\nPP3 disagreements with REVEL values:\n")
  print(pp3_disagree %>% select(chrom, pos, ref, alt, vcep_class, met_codes, revel, acmg_class))
}
```

# 6. Summary of discrepancy types

```{r summary}
cat("=== DISCREPANCY SUMMARY ===\n\n")

cat("1. CODES I CANNOT APPLY (VCEP-only, require patient data):\n")
vcep_only_summary %>%
  filter(!code %in% c("PS1","PS2","PM1","PM2","PP2","PP3","BA1","BS1","BP4")) %>%
  print()

cat("\n2. PM2 WEIGHT: I apply at Moderate (4.33x); VCEP applies at Supporting (2.08x)\n")
pm2_vcep %>% print()

cat("\n3. PS2 STRENGTH: I apply at Moderate; VCEP uses:\n")
vcep_long %>% filter(code == "PS2") %>% count(strength, sort=TRUE) %>% print()

cat("\n4. PP2 RESTRICTION: I restrict to BRAF/MAP2K1/PTPN11 —\n")
cat("   VCEP genes where PP2 is applied:\n")
vcep_long %>%
  filter(code == "PP2") %>%
  left_join(vcep_matched %>% select(chrom, pos, ref, alt, gene = GeneSymbol), by = var_key) %>%
  count(gene) %>% print()

cat("\n5. CODES I FIRE THAT VCEP DOESN'T:\n")
mine_only_variants %>% count(code, sort=TRUE) %>% print()
```
