---
title: "Gene group single-feature AUC, AUPRC, F1, MCC"
output: html_document
date: "2026-02-13"
---

# Aim

For each gene group and dataset, calculate classification metrics for individual features (somatic evidence and CEPs):

- AUC, AUPRC
- Sensitivity, specificity, precision, recall, accuracy, balanced accuracy, F1, MCC at MCC-optimal threshold

Output:
- `Data/Tables/Gene_group_single_feature_metrics.csv` (supplementary table with all metrics)
- `Data/AUC_tables/Gene_group_models/AUC_table_gene_groups_single_features_<date>.csv` (downstream input)

```{r, message=FALSE}
library(tidyverse)
library(pROC)
library(PRROC)

date <- "240426"

dir.create("../../../Data/Tables", recursive = TRUE, showWarnings = FALSE)
dir.create("../../../Data/AUC_tables/Gene_group_models", recursive = TRUE, showWarnings = FALSE)

buccal_panel_genes <- read_csv(
  "../../../Data/Buccal_data/Lawson_supplementary_table1_buccal_nanoseq_panel.csv",
  show_col_types = FALSE) %>% pull(gene) %>% unique()
buccal_predictors <- c("n_buccal", "n_buccal_codon_count")
```

# Import data

```{r, message=FALSE}
bases <- c("Clinvar_vs_Biobank_unique", "Clinvar_vs_Biobank_long",
           "DNM_vs_Biobank_unique",     "DNM_vs_Biobank_long")
sets  <- c("set1")

for (base in bases) {
  for (set in sets) {
    obj_name  <- paste0(base, "_", set, "_", date)
    file_path <- file.path(
      "../../../Data/ML_model_input/Gene_group_models/",
      paste0(base, "_", set, "_", date, ".csv.gz")
    )
    assign(obj_name, read_csv(file_path, show_col_types = FALSE))
  }
}
```

# Metrics functions

```{r}
inverted_predictors <- c("ESM1b", "ESM1v", "popEVE", "SIFT")

midpoint_thresholds <- function(x, max_candidates = 500L) {
  vals <- sort(unique(x[!is.na(x)]))
  if (length(vals) <= 1L) return(vals)
  mids <- (vals[-1L] + vals[-length(vals)]) / 2
  candidates <- c(vals[1L] - 1e-12, mids, vals[length(vals)] + 1e-12)
  if (length(candidates) > max_candidates)
    candidates <- unname(quantile(candidates, probs = seq(0, 1, length.out = max_candidates)))
  candidates
}

safe_div <- function(num, den) ifelse(den > 0, num / den, NA_real_)

add_metrics_single_feature <- function(dat, predictor_col_name,
                                       inverted_list = inverted_predictors) {
  dat <- dat %>%
    mutate(Pathogenicity_category = case_when(
      Pathogenicity_category == "Pathogenic" ~ 1,
      Pathogenicity_category == "Benign"     ~ 0,
      TRUE                                   ~ NA_real_
    ))

  dat2 <- dat %>%
    filter(!is.na(Pathogenicity_category) & !is.na(.data[[predictor_col_name]]))

  na_row <- list(
    Predictor = predictor_col_name,
    AUC_full = NA_real_, AUC_full_lower = NA_real_, AUC_full_upper = NA_real_,
    AUPRC = NA_real_, Threshold = NA_real_,
    Sensitivity = NA_real_, Specificity = NA_real_,
    Precision = NA_real_, Recall = NA_real_, Accuracy = NA_real_,
    BalancedAccuracy = NA_real_, F1 = NA_real_, MCC = NA_real_,
    Observed_direction = NA_character_,
    Expected_direction = if (predictor_col_name %in% inverted_list)
      "lower=pathogenic (higher=benign)" else "higher=pathogenic",
    Consistent_with_expected = NA
  )

  if (nrow(dat2) == 0 || length(unique(dat2$Pathogenicity_category)) < 2) return(na_row)

  score <- as.numeric(dat2[[predictor_col_name]])
  y     <- as.integer(dat2$Pathogenicity_category)

  r <- pROC::roc(response = y, predictor = score, levels = c(0, 1),
                 direction = "auto", quiet = TRUE, na.rm = TRUE)

  auc_val <- round(as.numeric(pROC::auc(r)), 4)
  auc_ci  <- pROC::ci.auc(r)
  dir_sym <- r$direction
  obs_dir <- if (identical(dir_sym, "<")) "higher=pathogenic (controls < cases)"
             else                         "higher=benign (controls > cases)"
  exp_dir <- if (predictor_col_name %in% inverted_list)
               "lower=pathogenic (higher=benign)" else "higher=pathogenic"
  consistent <- (grepl("benign", obs_dir)     && grepl("benign", exp_dir)) ||
                (grepl("pathogenic", obs_dir) && grepl("pathogenic", exp_dir))

  inverted  <- predictor_col_name %in% inverted_list
  pr_score  <- if (inverted) -score else score
  pos_s     <- pr_score[y == 1]
  neg_s     <- pr_score[y == 0]
  auprc_val <- if (length(pos_s) > 0 && length(neg_s) > 0)
    round(as.numeric(PRROC::pr.curve(pos_s, neg_s, curve = FALSE)$auc.integral), 4)
  else NA_real_

  candidates <- midpoint_thresholds(score)
  best_mcc <- -Inf; best_thr <- NA_real_; best_sens <- NA_real_; best_spec <- NA_real_

  for (thr_cand in candidates) {
    pred_pos <- if (!inverted) as.integer(score >= thr_cand) else as.integer(score <= thr_cand)
    TP <- as.numeric(sum(pred_pos == 1 & y == 1, na.rm = TRUE))
    TN <- as.numeric(sum(pred_pos == 0 & y == 0, na.rm = TRUE))
    FP <- as.numeric(sum(pred_pos == 1 & y == 0, na.rm = TRUE))
    FN <- as.numeric(sum(pred_pos == 0 & y == 1, na.rm = TRUE))
    s  <- safe_div(TP, TP + FN); p <- safe_div(TN, TN + FP)
    dn <- sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))
    m  <- if (!is.na(dn) && dn > 0) ((TP * TN) - (FP * FN)) / dn else NA_real_
    if (!is.na(m) &&
        (m > best_mcc ||
         (isTRUE(all.equal(m, best_mcc)) && !is.na(s) && (is.na(best_sens) || s > best_sens)) ||
         (isTRUE(all.equal(m, best_mcc)) && isTRUE(all.equal(s, best_sens)) &&
          !is.na(p) && (is.na(best_spec) || p > best_spec)))) {
      best_mcc <- m; best_thr <- thr_cand; best_sens <- s; best_spec <- p
    }
  }

  pred_pos <- if (!is.na(best_thr)) {
    if (!inverted) as.integer(score >= best_thr) else as.integer(score <= best_thr)
  } else rep(NA_integer_, length(score))

  TP <- as.numeric(sum(pred_pos == 1 & y == 1, na.rm = TRUE))
  TN <- as.numeric(sum(pred_pos == 0 & y == 0, na.rm = TRUE))
  FP <- as.numeric(sum(pred_pos == 1 & y == 0, na.rm = TRUE))
  FN <- as.numeric(sum(pred_pos == 0 & y == 1, na.rm = TRUE))
  N  <- TP + TN + FP + FN

  prec <- safe_div(TP, TP + FP); rec <- safe_div(TP, TP + FN)
  acc  <- if (N > 0) (TP + TN) / N else NA_real_
  bal  <- mean(c(best_sens, best_spec), na.rm = TRUE)
  f1   <- if (!is.na(prec) && !is.na(rec) && (prec + rec) > 0) 2 * prec * rec / (prec + rec) else NA_real_
  dn2  <- sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))
  mcc  <- if (!is.na(dn2) && dn2 > 0) ((TP * TN) - (FP * FN)) / dn2 else NA_real_

  list(
    Predictor = predictor_col_name,
    AUC_full = auc_val, AUC_full_lower = round(auc_ci[1], 4), AUC_full_upper = round(auc_ci[3], 4),
    AUPRC = auprc_val,
    Threshold        = if (!is.na(best_thr))  round(best_thr, 6) else NA_real_,
    Sensitivity      = if (!is.na(best_sens)) round(best_sens, 4) else NA_real_,
    Specificity      = if (!is.na(best_spec)) round(best_spec, 4) else NA_real_,
    Precision        = if (!is.na(prec))      round(prec, 4) else NA_real_,
    Recall           = if (!is.na(rec))       round(rec, 4) else NA_real_,
    Accuracy         = if (!is.na(acc))       round(acc, 4) else NA_real_,
    BalancedAccuracy = if (!is.na(bal))       round(bal, 4) else NA_real_,
    F1               = if (!is.na(f1))        round(f1, 4) else NA_real_,
    MCC              = if (!is.na(mcc))       round(mcc, 4) else NA_real_,
    Observed_direction       = obs_dir,
    Expected_direction       = exp_dir,
    Consistent_with_expected = consistent
  )
}

make_metrics_df <- function(my_data, cols, inverted_list = inverted_predictors) {
  bind_rows(lapply(cols, \(col)
    as_tibble(add_metrics_single_feature(my_data, col, inverted_list))
  ))
}
```

# Compute metrics per gene group

```{r, message=FALSE}
predictors <- c(
  "AlphaMissense", "REVEL", "ESM1b",
  "SIFT", "Polyphen2_HVAR", "MutationAssessor", "MetaSVM", "M-CAP",
  "PrimateAI", "VARITY_R", "CADD_raw", "popEVE", "ESM1v",
  "n_COSMIC", "n_COSMIC_genome", "n_COSMIC_targeted",
  "n_COSMIC_codon_count", "n_COSMIC_genome_codon_count", "n_COSMIC_targeted_codon_count",
  "hotcodon_cancer",
  "n_sperm", "n_sperm_exomes", "n_sperm_codon_count", "n_sperm_exomes_codon_count",
  "n_buccal", "n_buccal_codon_count"
)

data_sets <- paste0(
  c("Clinvar_vs_Biobank_unique", "Clinvar_vs_Biobank_long",
    "DNM_vs_Biobank_unique",     "DNM_vs_Biobank_long"),
  "_set1_", date
)

# Named list: Gene_group label → filter expression
group_filters <- list(
  All                   = \(df) df,
  AltFunc               = \(df) filter(df, altfunc_DD_cancer_gene       == TRUE),
  Non_RASopathy_AltFunc = \(df) filter(df, altfunc_minus_RASopathy_gene == TRUE),
  RASopathy             = \(df) filter(df, RASopathy_gene               == TRUE),
  LOF                   = \(df) filter(df, lof_DD_cancer_gene           == TRUE),
  Other_DD_somatic      = \(df) filter(df, other_DD_cancer_gene         == TRUE)
)

all_results <- list()

non_buccal_predictors <- setdiff(predictors, buccal_predictors)

for (group_name in names(group_filters)) {
  for (data_name in data_sets) {
    my_data        <- group_filters[[group_name]](get(data_name))
    my_data_buccal <- filter(my_data, gene %in% buccal_panel_genes)
    metrics <- bind_rows(
      make_metrics_df(my_data,        non_buccal_predictors),
      make_metrics_df(my_data_buccal, buccal_predictors)
    ) %>%
      mutate(data_name = data_name, gene_group = group_name)
    all_results[[paste(group_name, data_name, sep = "|")]] <- metrics
  }
}

all_dat <- bind_rows(all_results) %>%
  mutate(
    data_name = str_replace(data_name, "^Clinvar_vs_Biobank_long",   "clinvar_long_uncontaminated"),
    data_name = str_replace(data_name, "^Clinvar_vs_Biobank_unique",  "clinvar_unique"),
    data_name = str_replace(data_name, "^DNM_vs_Biobank_long",        "DNM_long_uncontaminated"),
    data_name = str_replace(data_name, "^DNM_vs_Biobank_unique",      "DNM_unique"),
    data_name = str_remove(data_name, "_\\d{6}$")
  ) %>%
  extract(data_name,
          into  = c("data_name", "gene_set"),
          regex = "^(.*)_(set\\d+)(?:_.*)?$",
          remove = FALSE) %>%
  mutate(gene_group = case_when(
    gene_group == "All"                   ~ "all",
    gene_group == "AltFunc"               ~ "altfunc",
    gene_group == "Non_RASopathy_AltFunc" ~ "altfunc_minus_RAS",
    gene_group == "RASopathy"             ~ "RASopathy",
    gene_group == "LOF"                   ~ "lof",
    gene_group == "Other_DD_somatic"      ~ "other",
    TRUE                                  ~ gene_group
  ))

all_dat
```

# Export

```{r}
write_csv(all_dat, "../../../Data/Tables/Gene_group_single_feature_metrics.csv")
write_csv(all_dat,
          paste0("../../../Data/AUC_tables/Gene_group_models/AUC_table_gene_groups_single_features_",
                 date, ".csv"))
```
