---
title: "Gene group LR models and single-feature AUCs"
output: html_document
date: "2025-03-20"
---

# Aim

For each gene group, calculate:

1. AUC for individual somatic and computational features (on all data)
2. Logistic regression model AUC combining one CEP + log-transformed COSMIC/sperm/buccal codon counts (on held-out test data), with DeLong test vs CEP-only model

Gene groups: AltFunc, AltFunc minus RASopathy, LOF, Other, RASopathy, All
Mutation datasets: ClinVar path vs Biobank (unique and long_uncontaminated), DNM vs Biobank (unique and long_uncontaminated)
CEPs: REVEL, AlphaMissense, ESM1b

Model training: 10 × 10-fold repeated cross-validation, downsampling for class imbalance.

```{r, message=FALSE}
library(tidyverse)
library(janitor)
library(pROC)
library(caret)
library(PRROC)

date <- "240426"

dir.create("../../../Data/AUC_tables/Gene_group_models",  recursive = TRUE, showWarnings = FALSE)
dir.create("../../../Data/LR_models/Gene_group_models",   recursive = TRUE, showWarnings = FALSE)
```

# Import data

```{r, message=FALSE}
input_folder <- "../../../Data/ML_model_input/Gene_group_models/"

# Load train/test split files (named e.g. clinvar_unique_set1_altfunc_train_240426.csv.gz)
csv_files <- list.files(input_folder, pattern = paste0(date, "\\.csv\\.gz$"), full.names = TRUE)

for (file in csv_files) {
  file_name <- tools::file_path_sans_ext(tools::file_path_sans_ext(basename(file)))
  assign(file_name, readr::read_csv(file, show_col_types = FALSE), envir = .GlobalEnv)
}

# Load whole-dataset files (unique and long, for single-feature AUCs)
bases <- c("Clinvar_vs_Biobank_unique", "Clinvar_vs_Biobank_long",
           "DNM_vs_Biobank_unique",     "DNM_vs_Biobank_long")
sets  <- c("set1")

for (base in bases) {
  for (set in sets) {
    obj_name  <- paste0(base, "_", set, "_", date)
    file_path <- file.path(input_folder, paste0(base, "_", set, "_", date, ".csv.gz"))
    assign(obj_name, read_csv(file_path, show_col_types = FALSE))
  }
}
```

# LR model training

```{r}
set.seed(42)
seeds <- vector(mode = "list", length = 101)
for (i in 1:100) seeds[[i]] <- sample.int(1000, 1)
seeds[[101]] <- sample.int(1000, 1)

twoClassSummary_pathogenic <- function(data, lev = NULL, model = NULL) {
  if (is.null(lev)) lev <- levels(data$obs)
  roc_obj <- pROC::roc(response  = data$obs,
                       predictor = data[, "Pathogenic"],
                       levels    = c("Benign", "Pathogenic"),
                       quiet     = TRUE)
  c(ROC  = as.numeric(pROC::auc(roc_obj)),
    Sens = caret::sensitivity(data$pred, data$obs, positive = "Pathogenic"),
    Spec = caret::specificity(data$pred, data$obs, negative = "Pathogenic"))
}

trctrl <- trainControl(
  method          = "repeatedcv",
  number          = 10,
  repeats         = 10,
  sampling        = "down",
  summaryFunction = twoClassSummary_pathogenic,
  classProbs      = TRUE,
  seeds           = seeds
)
```

```{r}
preprocess_df <- function(df, cep) {
  df %>%
    mutate(
      Pathogenicity_category   = factor(Pathogenicity_category, levels = c("Benign", "Pathogenic")),
      n_COSMIC_codon_count_log = log1p(n_COSMIC_codon_count),
      n_sperm_codon_count_log  = log1p(n_sperm_codon_count),
      n_buccal_codon_count_log = log1p(n_buccal_codon_count)
    ) %>%
    select(
      Pathogenicity_category,
      !!sym(cep),
      n_COSMIC_codon_count_log,
      n_sperm_codon_count_log,
      n_buccal_codon_count_log
    ) %>%
    drop_na()
}

extract_coefs <- function(model, cep) {
  cf   <- coef(model$finalModel)
  getc <- function(nm) if (nm %in% names(cf)) unname(cf[[nm]]) else NA_real_
  tibble(
    coef_intercept = getc("(Intercept)"),
    coef_cep       = getc(cep),
    coef_COSMIC    = getc("n_COSMIC_codon_count_log"),
    coef_sperm     = getc("n_sperm_codon_count_log"),
    coef_buccal    = getc("n_buccal_codon_count_log")
  )
}

train_lr <- function(train_df, formula) {
  train(formula, data = train_df, method = "glm", family = "binomial",
        trControl = trctrl, metric = "ROC")
}

compute_test_row <- function(data_name, gene_group, set, cep,
                             lr_full, lr_cep, test_df, coef_tbl) {
  probs_full <- predict(lr_full, test_df, type = "prob")[, "Pathogenic"]
  probs_cep  <- predict(lr_cep,  test_df, type = "prob")[, "Pathogenic"]

  roc_full   <- roc(test_df$Pathogenicity_category, probs_full, quiet = TRUE)
  roc_cep    <- roc(test_df$Pathogenicity_category, probs_cep,  quiet = TRUE)
  ci_full    <- ci.auc(roc_full)
  ci_cep     <- ci.auc(roc_cep)

  dl         <- roc.test(roc_full, roc_cep, method = "delong")

  bind_cols(
    tibble(
      data_name      = data_name, gene_group = gene_group, gene_set = set,
      cep            = cep,       test_or_train = "test",
      AUC_full       = round(auc(roc_full), 3),
      AUC_full_lower = round(ci_full[1], 3),  AUC_full_upper = round(ci_full[3], 3),
      AUC_cep_only   = round(auc(roc_cep),  3),
      AUC_cep_lower  = round(ci_cep[1],  3),  AUC_cep_upper  = round(ci_cep[3],  3),
      delong_p_value = signif(dl$p.value, 3),
      AUC_diff       = round(dl$estimate[[1]] - dl$estimate[[2]], 3),
      AUC_diff_lower = round(dl$conf.int[[1]], 3),
      AUC_diff_upper = round(dl$conf.int[[2]], 3)
    ),
    coef_tbl
  )
}
```

```{r}
data_names  <- c("clinvar_long_uncontaminated", "clinvar_unique",
                 "DNM_long_uncontaminated",     "DNM_unique")
sets        <- c("set1")
gene_groups <- c("altfunc", "altfunc_minus_RAS", "lof", "other", "RASopathy", "all")
CEPs        <- c("REVEL", "AlphaMissense", "ESM1b")

save_dir    <- "../../../Data/LR_models/Gene_group_models/"

all_auc_results <- tibble()

for (data_name in data_names) {
  for (gene_group in gene_groups) {
    for (set in sets) {
      for (cep in CEPs) {

        train_obj <- paste0(data_name, "_", set, "_", gene_group, "_train_", date)
        test_obj  <- paste0(data_name, "_", set, "_", gene_group, "_test_",  date)
        if (!exists(train_obj) || !exists(test_obj)) next

        train_df <- preprocess_df(get(train_obj), cep)
        test_df  <- preprocess_df(get(test_obj),  cep)

        lr_full <- train_lr(train_df, Pathogenicity_category ~ .)
        save(lr_full, file = file.path(
          save_dir, paste0(data_name, "_", set, "_", gene_group, "_", cep, "_lr_full_", date, ".rda")
        ))

        lr_cep <- train_lr(train_df, as.formula(paste0("Pathogenicity_category ~ ", cep)))

        coef_tbl <- extract_coefs(lr_full, cep)

        probs_tr <- predict(lr_full, train_df, type = "prob")[, "Pathogenic"]
        roc_tr   <- roc(train_df$Pathogenicity_category, probs_tr, quiet = TRUE)
        ci_tr    <- ci.auc(roc_tr)

        auc_train <- bind_cols(
          tibble(
            data_name = data_name, gene_group = gene_group, cep = cep,
            gene_set = set, test_or_train = "train",
            AUC_full = round(auc(roc_tr), 3),
            AUC_full_lower = round(ci_tr[1], 3),
            AUC_full_upper = round(ci_tr[3], 3)
          ),
          coef_tbl
        )

        auc_test <- compute_test_row(data_name, gene_group, set, cep,
                                     lr_full, lr_cep, test_df, coef_tbl)

        all_auc_results <- bind_rows(all_auc_results, auc_train, auc_test)
      }
    }
  }
}

write_csv(all_auc_results,
          paste0("../../../Data/AUC_tables/Gene_group_models/AUC_table_gene_groups_LR_models_", date, ".csv"))
```

# Gene and variant counts per group

```{r}
gene_count_DF    <- data.frame()
variant_count_DF <- data.frame()

for (data_name in data_names) {
  for (gene_group in gene_groups) {
    for (set in sets) {
      train_name <- paste0(data_name, "_", set, "_", gene_group, "_train_", date)
      test_name  <- paste0(data_name, "_", set, "_", gene_group, "_test_",  date)

      combined <- bind_rows(get(train_name), get(test_name))

      gene_count_DF <- rbind(gene_count_DF, data.frame(
        gene_group = gene_group, data_name = data_name, gene_set = set,
        gene_count = combined %>% count(gene) %>% nrow()
      ))

      variant_count_DF <- rbind(variant_count_DF, data.frame(
        gene_group        = gene_group, data_name = data_name, gene_set = set,
        n_path_variants   = sum(combined$Pathogenicity_category == "Pathogenic"),
        n_benign_variants = sum(combined$Pathogenicity_category == "Benign")
      ))
    }
  }
}

gene_count_DF
variant_count_DF
```

```{r}
write_csv(gene_count_DF,
          paste0("../../../Data/AUC_tables/Gene_group_models/Gene_counts_per_group_", date, ".csv"))
write_csv(variant_count_DF,
          paste0("../../../Data/AUC_tables/Gene_group_models/Variant_counts_per_group_", date, ".csv"))
```
